---
title: "Midterm Sheet"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---
## 

Lab 1_1: 
## RStudio Layout  
When you open RStudio, you will see that the window is divided into four sections. Each of these sections has a specific function that we will use throughout the course.  
1. **Source (Top Left)**: This is where you write and edit R scripts and RMarkdown files.  
2. **Console (Bottom Left)**: This is where R executes your commands.   
3. **Environment (Top Right)**: The Environment tab shows all the objects (like data frames, vectors, etc.) that you have created during your R session.  
4. **Files/Plots/Packages/Help/Viewer (Bottom Right)**: This section has multiple tabs. The Files tab allows you to navigate your file system. The Plots tab displays any graphs you create. The Packages tab lets you manage R packages. The Help tab provides access to R documentation, and the Viewer tab is used for viewing local web content. 

Lab 1_2:
## GitHub  
[GitHub](http://www.github.com) is a file storage and management site used by programmers. Programmers upload code to repositories (folders) and make it publicly available. We will use Github to store our work and manage workflows in BIS 15L. We have a main repository for our class [datascibiol](https://github.com/jmledford3115/datascibiol). You should **Star** this repository for future reference.  
## GitHub Desktop  
The easiest way to manage your GitHub repository is to use [GitHub desktop](https://desktop.github.com/). This is a  helpful tool that will allow you to manage all of your files. Open the program and login using your GitHub account.  
## Clone Your Repository  
In GitHub desktop, you will see your personal repository as an option under `clone repository`. Clone your repository to the desktop. Once you have done this any changes you make to files will be recorded and easily uploaded. Do not move this folder from your desktop.  
## Copy the Class Repository
In order to keep track of the class files, the last thing you need to do is make a copy of the class repository. I will make regular changes to the class repository including the addition of files and data that you will need. Do a search for `datascibiol` and download a copy to the desktop. You do this by navigating to the repository then going to the small green code button at top right. Click on the code button then go to Download ZIP.  
Copy the `lab1` folder from the class repository and paste it into your personal repository. By doing this, you have created your own copies of lab materials for lab 1.  
## Using GitHub  
You will use your repository to store all of your notes and work. There are three terms that we need to define:  
1. `Commit` takes a snapshot of your current progress.    
2. `Pull` this is the same as download.     
3. `Push` this is the same as upload.    
If you setup [GitHub desktop](https://desktop.github.com/), then you will see that it automatically tracks your changes. Clicking the `Commit to master` button in the bottom left corner means that you are recording a snapshot of your edits locally. Once this is done, you need to upload your changes to GitHub. This is done by clicking `Push origin`. Once you have pushed your changes you can double check by looking at your repository on github.com.    
## RMarkdown
The file format we use in class is called a markdown file. RMarkdown files have the extension `.Rmd` and can be created in RStudio by going to `File > New File > RMarkdown`. RMarkdown files allow you to combine text, code, and output in a single document.  
## Knitting
RMarkdown files need to be `knit` in order to produce a final document. This is done by clicking the `Knit` button at the top of the RStudio window. Knitting produces a final document in HTML format. You will turn in all homework assignments as knitted HTML files. If your code does not work, then the file will not knit.    

lab 2_intro:
1. What is the difference between R and RStudio?  R is the actual programming language doing the calculations/dataprocessing, RStudio is the dashboard/userinterface to make it easier to write code and manage files
2. What is GitHub and GitHub Desktop and why are they useful to programmers?  Github is a cloud-based data hosting service to store projects online and collaborate with others, Github Desktop is an application of Github outside of the website to sync files directly from the computer to the online cloud storage. 
3. What is version control and why is it useful to programmers?  Version control is a system of records changes to a file that can be accessed at any time. This lets you undo changes as needed and tracks the entire history of the file. 
4. What is a repository?  A repository is a project's folder containing all of its files, which is tracked online. 
5. What is the difference between committing and pushing?  committing will save the files locally on your computer, pushing will save the files online to the cloud

lab 2_1:
## Working Directories and Paths
Before we get started with R, it is important to understand the concept of working directories and paths. A working directory is the folder on your computer where R will look for files to read in and where it will save files that you create. You can check your current working directory using the `getwd()` function.
```{r}
getwd() #identify current working directory
```
What if you find that you are not in the correct working directory? You can change your working directory in two ways. The first is to look at the Session menu at the top of RStudio. Click on Session -> Set Working Directory -> Choose Directory. Then navigate to the folder you want to use as your working directory. 
The second way is to use the `setwd()` function. You will need to provide the full path to the folder you want to use. For example:
```{r}
#setwd("/Users/yourname/Documents/yourfolder") #uncomment and change to your path

#setwd("/Users/avongsav/Desktop/BIS15LW2026_avongsavanh/lab2")
```
What is a path? A path is the location of a file or folder on your computer. It tells R where to find files to read in or where to save files you create. Paths can be absolute or relative. An absolute path provides the full location of a file or folder, starting from the root directory. A relative path provides the location of a file or folder in relation to the current working directory.
## Objects
In order to access the potential of R we need to assign values or other types of data to `objects`. There is a specific format that I want you to follow, so please pay close attention.  
Assign a value to object 'x'. The '<-' symbol is read as 'gets'. In this case, x gets 42. Make sure that you are in the environment panel and you should see the value associated with 'x'. On a mac, you can push `option` and `-` to automatically generate the gets symbol.  
```{r}
x  <- 42 #x gets the value 42
```
Make two new objects, treatment and control. The value of treatment is 36 and the value of control is 38.
```{r}
treatment <- 36
control <- 38
```
What is the sum of treatment and control?
```{r}
treatment + control 
```
Here we make a new object `my_experiment` that is the sum of the treatment and control. Notice that I use `_` and not spaces. 
```{r}
my_experiment <- treatment + control
my_experiment
```
We can also use the function sum to do the same thing. Notice that if I give a new object the same name as an existing object, the old one is replaced.
```{r}
my_experiment <- sum(treatment,control)
```
To learn more about the sum function, use the help command `?`.
```{r}
?sum
```
What is the mean of 2+8+2? (Do the math in your head first). Does the following code match your prediction?
```{r}
sum <- 2 + 8 + 2 
mean <- sum / 3
``` 
Order of operations applies, so we need to tell R exactly what we want.
```{r}
(2+8+2)/3
```
Here we use the mean function. Notice that we use `c` which stands for concatenate. This combines the three numbers into a single object that the mean function can use. This type of data structure is called a vector.
```{r}
mean(c(2,8,2)) #c stands for combine and creates a vector
```
Why does this not work?
```{r}
mean(2,8,2) #outputs 2 
```
1. Create three new objects, `venom_GT`, `chiron`, and `veyron`. These are the fastest cars in the world. Assign each car to its top speed. The venom_GT can go 270, chiron is 261, and veyron is 268. 
```{r}
venom_GT <- 270
chiron <- 261
veyron <- 268
```
2. Use arithmetic to calculate the mean top speed for the cars.
```{r}
mean_speed <- (venom_GT + chiron + veyron)/3
mean_speed
```
3. Use the function `mean()` to calculate the mean top speed for the cars.
```{r}
mean_speed <- mean(c(venom_GT,chiron,veyron)) #make sure you use a vector using the c
```
## Types of Data
There are four frequently used `classes` of data: 1. numeric, 2. integer, 3. character, 4. logical.
```{r}
my_numeric <- 42
my_integer <- 2L #adding an L automatically denotes an integer
my_character <- "universe"
my_logical <- FALSE
```
To find out what type of data you are working with, use the `class()` function. This is important because sometimes we will need to change the type of data to perform certain analyses.
```{r}
class(my_numeric)
```
```{r}
class(my_integer)
```
You can use the `is()` and `as()` functions to clarify or specify a type of data.
```{r}
is.integer(my_numeric)
```
Let's convert my_numeric to an integer. 
```{r}
my_numeric <- as.integer(my_numeric) #important when we work with plots
```
```{r}
is.integer(my_numeric)
```
## Missing Data
R has a special way to designate missing data, the NA. NA values in R have specific properties which are very useful if your data contains any missing values. Later this quarter we will have a lab focused on dealing with NAs.  
NA values are used to designate missing data. `is.na` or `anyNA` are useful functions when dealing with NAs in data. 
```{r}
my_missing <- NA #NA stands for missing
```
```{r}
is.na(my_missing)
```
```{r}
anyNA(my_missing)
```
## Practice  
1. Let's create a vector that includes some missing data. For now, run the following code chunk.  
```{r}
new_vector <- c(7, 6.2, 5, 9, NA, 4, 9.8, 7, 3, 2)
```
2. Calculate the mean of `new_vector`.
```{r}
mean(new_vector)
```
4. Recalculate the mean using the following code chunk. Why is the useful?  
```{r}
mean(new_vector, na.rm = TRUE) #removes NA values, can use "T" instead of "TRUE"
```

Hw 2:
**4. Write a code chunk that divides `pi` by 2. Use the help command `?` to learn how to use the `round` function to limit your result to 4 significant digits.**  
```{r}
?round

piover2 <- pi/2
piover2rounded <- round(piover2,4)
piover2rounded
```
**5. Calculate the mean for: 2, 8, 4, 6, NA, 4, 9, 9, 10, NA. Please start by making a new vector `x` that holds these values then use `mean` to perform the calculation.**
```{r}
x = c(2, 8, 4, 6, NA, 4, 9, 9, 10, NA)
x
```
```{r}
mean(x, na.rm = T)
```

lab 3_1:
## Load the tidyverse
A library is a collection of R functions and data sets. The [tidyverse](https://tidyverse.org/) is a collection of R packages designed for data science. For this course, we will be using many of the packages in the tidyverse. We load the tidyverse with the command below.  
```{r}
library("tidyverse") 
#install.packages("package_name") #command to install packages
```
## Data Structures
In addition to classes of data, R also organizes data in different ways. These are called data structures and include vectors, lists, matrices, data frames, and factors. Here, we will introduce vectors and data frames.  
## Vectors
Vectors are a common way of organizing data in R.  We create vectors using the `c` command. The `c` stands for concatenate. We used this command in lab 2.  
A numeric vector.
```{r}
my_vector <- c(10, 20, 30) #this makes a new vector
```
A character vector.
```{r}
days_of_the_week <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
```
A convenient trick for creating a vector to play around with is to generate a sequence of numbers.
```{r}
my_vector_sequence <- c(1:100)
my_vector_sequence
```
## Identifying vector elements
We can use `[]` to pull out elements in a vector. We just need to specify their position in the vector; i.e. day 3 is Wednesday.
```{r}
days_of_the_week[4]
```
```{r}
my_vector_sequence[10]
```
## Practice
1. Use `[]` to determine which element in `my_vector_sequence` has a value of 15.
```{r}
my_vector_sequence[15]
```
2. We can use operators such as <, >, ==, <==, etc. Show all values in `my_vector_sequence` that are less than or equal to 10. 
```{r}
my_vector_sequence[my_vector_sequence <= 10] #important for dna sequence data 
```
3. If you use `[]` then you only get the values, not the logical evaluation of the entire vector. Experiment with this by adjusting the chunk below.    
```{r}
my_vector_sequence <= 10
```
## Data Frames
The data frame is the most common way to organize data within R. A data frame stores data of many different classes. Essentially, data frames are spreadsheets like you would find in Excel. We usually don't build data frames in RStudio from scratch, but this example will show you how they are structured.  
Let's build separate vectors that include length (in), weight (oz), and sex of three [ruby-throated hummingbirds](https://www.audubon.org/field-guide/bird/ruby-throated-hummingbird).  
```{r}
Sex <- c("male", "female", "male") #character vectors, strings of characters
Length <- c(3.2, 3.7, 3.4) #integer vectors
Weight <- c(2.9, 4.0, 3.1)
```
Since we work in the tidyverse, we use `tibble()` to create a data frame.  
```{r}
hbirds <- tibble(Sex, Length, Weight) #columns referred to as variables
hbirds 
```
What are the column names of our data frame? Notice that R defaulted to using the names of our vectors, but we could name them something else when creating the data frame, or rename them later.  
```{r}
names(hbirds)
```
What are the dimensions of the `hbirds` data frame? The `dim()` and `str()` commands provide this information.   
```{r}
dim(hbirds)
```
```{r}
str(hbirds) #str stands for structures
```
Let's use lowercase names when we create the data frame. We just changed to lowercase here, but we could use any names we wish.  
```{r}
hbirds <- tibble(sex = Sex, length = Length, weight = Weight)
hbirds
```
## Accessing Data Frame Columns and Rows 
The same methods of selecting elements in vectors and data matrices apply to data frames. We use `[]`. We have two positions where the first applies to the rows, and the second to the columns.  
The first row.  
```{r}
hbirds[1,]
```
The third column.  
```{r}
hbirds[3] #using hbirds[,3] also works
```
```{r}
hbirds[1,3] #row, column
```
## Calculations
We can use the `$` to access a column (variable) in a data frame. Here we calculate the mean length of the hummingbirds.  
```{r}
hbirdsmean <- mean(hbirds$length)
hbirdsmean
```
## Writing Data to File
We should save our hbirds data frame so we can use it again later! There are many ways to save data in R, here we write our data frame to a csv file. We use `row.names = FALSE` to avoid row numbers from printing out (on the first row when opened on a spreadsheet)
```{r}
#write.csv(hbirds, "hbirds_data.csv", row.names = FALSE) 
```

lab 3_2:
## Change column classes
When making plots or performing statistical modeling, it may be important to convert column classes. For example, if a column contains categorical data (e.g., scientist names, treatment groups, etc.) it may be best to convert that column to a factor class.  
Recall that factors are categorical variables that have a fixed number of possible values called levels.  
What is the class of the scientist column? Let's change it to factor and then show the levels of that factor.  
```{r}
names(hot_springs)
```
```{r}
class(hot_springs$scientist)
```
Did our change work? We can also check the levels of each column.
```{r}
hot_springs$scientist <- as.factor(hot_springs$scientist)
```
```{r}
class(hot_springs$scientist)
```
## Practice
1. Change the class of the springs column to factor.
```{r}
hot_springs$spring <- as.factor(hot_springs$spring)
```
2. What are the levels of the springs column?
```{r}
levels(hot_springs$spring)
```
```{r}
table(hot_springs$spring)
```
We can summarize our data frame with the`summary()` function.  (uses fish <- (another csv file))
```{r}
names(fish)
```
```{r}
str(fish)
```
```{r}
summary(fish)
```
`glimpse()` is another useful summary function.
```{r}
glimpse(fish)
```
`nrow()` gives the numbers of rows.
```{r}
nrow(fish)
```
`ncol` gives the number of columns.
```{r}
ncol(fish)
```
`dim()` gives the dimensions.
```{r}
dim(fish)
```
`names` gives the column names.
```{r}
names(fish)
```
`head()` prints the first n rows of the data frame.
```{r}
head(fish)
```
`tail()` prinst the last n rows of the data frame.
```{r}
tail(fish)
```
`table()` is useful when you have a limited number of categorical variables. It produces fast counts of the number of observations in a variable.
```{r}
table(fish$fish_id)
```

Hw 3 not included

Lab 4: 
## Libraries
```{r}
library(tidyverse)
library(palmerpenguins)
```
## Grammar of Graphics
The ability to quickly produce and customize graphs is a strength of R. Data visualizations are produced by the package `ggplot2` and it is a core part of the tidyverse. The syntax for using ggplot is specific and common to all types of plots. This is what Hadley Wickham calls a [Grammar of Graphics](http://vita.had.co.nz/papers/layered-grammar.pdf). The "gg" in `ggplot` stands for grammar of graphics.
## Philosophy
What makes a good chart? In my opinion a good chart is elegant in its simplicity. It provides a clean, clear visual of the data without being overwhelming to the reader. This can be hard to do and takes some careful thinking. Always keep in mind that the reader will almost never know the data as well as you do so you need to be mindful about how you present the facts.  
## Data Types
We first need to define some of the data types we will use to build plots.  
+ `discrete` quantitative data that only contains integers
+ `continuous` quantitative data that can take any numerical value (like weight, height)
+ `categorical` qualitative data that can take on a limited number of values 
## Basics
The syntax used by ggplot takes some practice to get used to, especially for customizing plots, but the basic elements are the same. It is helpful to think of plots as being built up in layers.  
In short, **plot= data + geom_ + aesthetics**.  
We start by calling the ggplot function, identifying the data, and specifying the axes. We then add the `geom` type to describe how we want our data represented. Each `geom_` works with specific types of data and R is capable of building plots of single variables, multiple variables, and even maps. Lastly, we add aesthetics.
Let's start by asking the question: How does body mass vary among penguin species?
To make a plot, we need to first specify the data and map the aesthetics. The aesthetics include how each variable in our data set will be used. In the example below, I am using the aes() function to identify the x and y variables in the plot.
```{r}
ggplot(data=penguins, #specifies the data
       mapping = aes(x = species,y = body_mass_g)) #map the aesthetics
```
Notice that we have a nice background, labeled axes, and even a value range of our variables on the y-axis- but no plot. This is because we need to tell ggplot how we want our data represented. This is called the geometry or `geom()`. There are many types of `geom`, see the ggplot [cheatsheet](https://posit.co/resources/cheatsheets/).
Here we want a boxplot, specified by `geom_boxplot()`. We will explore boxplots in more detail later, but for now we just need an example.
```{r}
ggplot(data=penguins, #specifies the data
       mapping = aes(x = species,y = body_mass_g)) + #map the aesthetics, note the plus sign
  geom_boxplot()
```
## Scatterplots and barplots
Now that we have a general idea of the syntax, let's explore two common plots: 1) scatter plots and 2) bar plots.
### 1. Scatter Plots
Scatter plots are good at revealing relationships that are not readily visible in the raw data.   
Let's ask the question: Is there a relationship between body mass and flipper length?
```{r}
names(penguins)
```
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point()
```
Notice the warning! R is telling us that there are some missing values in our data. This is common in real world data sets. R automatically omits these missing values when plotting. We can also deal with the NA's explicitly using the `na.rm=T` function.  
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point(na.rm = T)
```
To add a regression (best of fit) line, we add another layer.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point( na.rm = T) +
  geom_smooth(method = lm, se = T,  na.rm = T) #linear model, se = standard error
```
This graph is fine, but it doesn't distinguish between species. It might be helpful for the reader to see how each species is represented. We can do this by mapping the color aesthetic to species.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm, color = species)) +
  geom_point(na.rm = T) +
  geom_smooth(method = lm, se = T, na.rm = T) #linear model, se = standard error
```
The plot above looks good, but I think it's a bit messy having the regression line presented for each species. When we add the color aesthetic, it is passed down to all layers. To fix this, we can move the color aesthetic to just the geom_point layer.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species),na.rm = T) +
  geom_smooth(method = lm, se = T, na.rm = T)
```
Another helpful aesthetic is `shape` which can be used to distinguish points by shape instead of or in addition to color. A common approach is to map both color and shape to the same variable.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species, shape = species),na.rm = T) +
  geom_smooth(method = lm, se = T, na.rm = T)
```
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point(mapping = aes(color = bill_depth_mm, shape = species),na.rm = T) + #color for continuous example, not recommended for visuals
  geom_smooth(method = lm, se = T, na.rm = T)
```
But don't we want a title? We do this using the `labs()` function.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species, shape = species),na.rm = T) +
  geom_smooth(method = lm, se = T, na.rm = T) + 
  labs(title = "Body Mass (g) vs. Flipper Length (mm)",
       x = "Body Mass (g)",
       y = "Flipper Length (mm)") #labs stands for title
```
### Bar Plot: `geom_bar()`
The simplest type of bar plot counts the number of observations in a categorical variable. In this case, we want to know how many observations are present in the variable `species`. Notice that we do not specify a y-axis because it is count by default.  
```{r}
names(penguins)
```
How many individuals per species?
```{r}
ggplot(data=penguins, 
       mapping = aes(x = species)) + 
    geom_bar(mapping = aes(x = species))
```
What if we want to use the color aesthetic like we did for `geom_point` above? Let's try...
```{r}
ggplot(data=penguins, 
       mapping = aes(x = species)) + 
    geom_bar(mapping = aes(fill = species))
```
## Practice
1. Make a bar plot showing the number of penguins on each island.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = island)) + 
    geom_bar(mapping = aes(fill = island))
```
2. Make a bar plot showing the number of penguins of each species on each island. (Hint: use `fill` to distinguish species)
```{r}
ggplot(data=penguins, 
       mapping = aes(x = island)) + 
    geom_bar(mapping = aes(fill = species), position = "dodge") + 
    labs(title = "#Penguins for each species by island")
```

Hw 4:
## Load the tidyverse
```{r}
library("tidyverse")
options(scipen = 999) #disable scientific notation
```
For this homework, we will be using the `msleep` dataset from the `ggplot2` package. This dataset contains information about the sleep habits of various mammals. [Reference](https://ggplot2.tidyverse.org/reference/msleep.html): V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007.
```{r}
msleep <- msleep
```
**1. Use a summary function of your choice to get an idea of the structure of the msleep data.**
```{r}
glimpse(msleep)
```
**2. What are the names of the variables? To get a better idea of what the variables mean, you can use `?msleep`.**
```{r}
?msleep
names(msleep)
```
**3. Make a plot that shows the number of mammals in each vore type. Which vore type is most represented in the data?**
```{r}
ggplot(data = msleep,
      mapping = aes(x = vore)) + 
      geom_bar(mapping = aes(fill = vore)) +
      labs(title = "Number of Mammals In Each Vore Type",
       x = "Vore Type",
       y = "Number of Mammals")
```
The herbi vore type is most represented by the data.
**4. What is the average length of time that mammals sleep?**
```{r}
sleep_average = mean(msleep$sleep_total)
sleep_average
```
The average length of time that mammals sleep is 10.43373 hours.
**5. Use `summary()` to get a quick idea of the range of values for the variable `bodywt`.**
```{r}
summary(msleep$bodywt)
```
**6. Build a plot that shows the range of `bodywt.` by vore type. Because the range of `bodywt` is so variable, you will need to scale the y-axis. You do this by adding `scale_y_log10()` as another layer.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = vore,y = bodywt)) + 
      geom_boxplot(mapping = aes(fill = vore)) +
      scale_y_log10() + 
      labs(title = "Vore Type vs. Bodyweight (kg)",
       x = "Vore Type",
       y = "Bodyweight (kg)")
```
**7. Is there a relationship between body weight and total sleep time? Add appropriate labels and a title to your plot.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = bodywt,y = sleep_total)) + 
      geom_point(mapping = aes(color = vore)) + #color instead of fill for dots 
      geom_smooth(method = lm, se = F, na.rm = T) +
      scale_x_log10() + 
      labs(title = "Bodyweight (kg) vs. Total Sleep Time (hr)",
       x = "Bodyweight (kg)",
       y = "Total Sleep Time (hr)")
```
There is a relationship between bodyweight and total sleep time.
**8. Which vore type tends to sleep the most? Build a plot to support your answer.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = vore,y = sleep_total)) + 
      geom_boxplot(mapping = aes(fill = vore)) +
      labs(title = "Vore Type vs. Total Sleep Time (hr)",
       x = "Vore Type",
       y = "Total Sleep Time (hr)")
```
The insecti vore type tends to sleep the most. 
**9. Is there a relationship between brain weight and body weight? Build a plot to support your answer.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = brainwt,y = bodywt)) + 
      geom_point(mapping = aes(color = vore)) +
      geom_smooth(method = lm, se = F, na.rm = T) +
      scale_y_log10() + 
      scale_x_log10() + 
      labs(title = "Brain Weight (kg) vs. Body Weight (kg)",
       x = "Brain Weight (kg)",
       y = "Body Weight (kg)")
```
There is a relationship between brain weight and body weight.
**10. Build one plot of your choice that provides additional insight into the msleep dataset. Be sure to include appropriate labels and a title.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = conservation,y = awake)) + 
      geom_boxplot(mapping = aes(fill = vore)) +
      labs(title = "Conservation Type vs. Time Awake (hr)",
       x = "Conservation Type",
       y = "Time Awake (hr)")
```

Lab 5_1: ## Tidyverse
Recall that the [tidyverse](www.tidyverse.org) is a collection of packages that make workflow in R easier. The packages operate more intuitively than base R commands and share a common organizational philosophy. In lab 4, we learned how to use `ggplot2` to make visualization of data easier. In this lab, we will learn how to use the package `dplyr` to wrangle data. 
## dplyr
The first package that we will use to wrangle data is `dplyr`. `dplyr` is used to transform data frames by extracting, rearranging, and summarizing data such that they are focused on a question of interest. This is very helpful,  especially when wrangling large data, and makes dplyr one of most frequently used packages in the tidyverse. The two functions we will use most are `select()` and `filter()`.  
These functions are often called verbs, and the format used for each is the same. The output is always a new, more restricted dataframe.  
## `select()`
The verb `select()` allows you to pull out columns of interest from a dataframe; it does not affect the rows. To do this, just add the names of the columns of interest to the `select()` command. The order in which you add them, will determine the order in which they appear in the output.  
For smaller dataframes, `select()` may not make much sense. However, for larger dataframes with many variables, `select()` is very useful. Let's look at the fish data again.
```{r}
names(penguins)
```
We are only interested in species and body mass. We can use `select()` to extract these columns.
```{r}
select(penguins, species, body_mass_g)
```
To add a range of columns use `start_col:end_col`.
```{r}
select(penguins, species:flipper_length_mm)
```
```{r}
penguins <- penguins
penguins
```
The ! operator is useful in select. It allows us to select everything except the specified variables. (! is "naught")
```{r}
select(penguins, !body_mass_g)
```
Alternatively, you can use the `c()` function to exclude multiple columns. (c() is concatenate)
```{r}
select(penguins, !c(species, island, year, sex))
```
For very large data frames with lots of variables, `select()` utilizes lots of different operators to make things easier. Let's say we are only interested in the variables that deal with length.
```{r}
select(penguins,contains("mm"))
```
`starts_with()`
```{r}
select(penguins,starts_with("bill"))
```
`ends_with()`
```{r}
select(penguins,ends_with("mm"))
```
You can also select columns based on the class of data.  
```{r}
select(penguins, where(is.numeric))
```
There are a few verbs that go with `select()` because they deal with columns. One of these is `rename()`. `rename()` allows you to rename columns in a dataframe. The format is `new_name = old_name`.
```{r}
rename(penguins, body_mass=body_mass_g)
```
Alternatively, you can rename from within select.
```{r}
select(penguins, species, body_mass=body_mass_g)
```
The second is `relocate()`. `relocate()` allows you to move columns to different locations within a dataframe. Let's say we want to move `length` to be the first column in the dataframe.
```{r}
relocate(penguins, year)
```

Lab 5_2:
## Review  
In the previous lab, we used `select()` to extract columns of interest from a data frame. This helps us focus our attention on the variables relevant to our question. However, it doesn't allow us to extract information in rows. The `filter()` function allows us to extract data in rows that meet specific criteria.  
## `filter()`
Unlike `select()`, `filter()` allows us to extract data that meet specific criteria within a row. Let's say that we are interested only in the Adelie penguins from the `penguins` data frame. We can use `filter()` to extract these observations.  
```{r}
glimpse(penguins)
```
```{r}
filter(penguins, species == "Adelie")
```
Similarly, if we are only interested in penguins with a body mass greater than or equal to 5500g we can use `filter()` to extract these observations.  
```{r}
filter(penguins, body_mass_g >= 5500)
```
+ `filter()` allows all of the expected operators; i.e. >, >=, <, <=, != (not equal), and == (equal).  
Using the `!` operator allows for the exclusion of specific observations.
```{r}
filter(penguins, species!= "Chinstrap")
```
## Using `filter()` with multiple observations  
The `%in%` [operator](https://www.tutorialspoint.com/r/r_operators.htm) determines whether or not a value is part of a vector.    
```{r}
filter(penguins, body_mass_g %in% c(5250,5500))
```
Alternatively, you can use `between` if you are looking for a range of specific values.  
```{r}
filter(penguins, between(body_mass_g, 5250, 5500))
```
How would you use `%in%` to get the same result?
```{r}
filter(penguins, body_mass_g %in% c(5250:5500)f) # or c(5250, 5300, 5350, 5400, 5450, 5500) if you know those are the other values
```
You can also extract observations "near" a certain value but you need to specify a tolerance.  
```{r}
filter(penguins, near(body_mass_g, 5400, tol = 100))
```
## Practice
The data are from: Gaeta J., G. Sass, S. Carpenter. 2012. Biocomplexity at North Temperate Lakes LTER: Coordinated Field Studies: Large Mouth Bass Growth 2006. Environmental Data Initiative.  [link](https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-ntl&identifier=267)  
1. Load the data
```{r}
fish <- read.csv("data/Gaeta_etal_CLC_data.csv")
fish
```
2. Filter the `fish` data to include the samples from lake "BO".
```{r}
names(fish)
```
```{r}
filter(fish,lakeid == "BO")
```
3. Filter the data to include all lakes except "AR".
```{r}
filter(fish,lakeid != "AR")
```
4. Filter the fish data to include all fish with a scalelength within 0.25 of 8.
```{r}
filter(fish, near(scalelength, 8, tol = 0.25))
```
5. Filter the fish data to include fish with a scalelength between 2 and 4.
```{r}
filter(fish, between(scalelength, 2, 4))
```
## Using `filter()` on multiple conditions
You can also use `filter()` to extract data based on multiple conditions. Below we extract the Chinstrap penguins that have a body mass greater than 4500g.
```{r}
names(penguins)
```
```{r}
filter(penguins, species == "Chinstrap" & body_mass_g > 4500) #can use , instead of &
```
Notice that the `|` operator generates a different result. Why?
```{r}
filter(penguins, species == "Chinstrap" | body_mass_g > 4500) # | is an "OR" so as long as one condition is fulfilled it will be included
```
Rules:  
+ `filter(condition1, condition2)` will return rows where both conditions are met. By default the , means &.    
+ `filter(condition1, !condition2)` will return all rows where condition one is true but condition 2 is not.  
+ `filter(condition1 | condition2)` will return rows where condition 1 or condition 2 is met.  
In this case, we filter out the penguins with a body mass over 5000g and a bill length over 50mm or a bill depth over 18mm.
```{r}
filter(penguins, body_mass_g > 5000, (bill_length_mm > 50 | bill_depth_mm > 18))
```
## Pipes `%>%` 
In order to start combining `select()`, `filter()`, and other functions efficiently, we need to learn pipes. Pipes feed the output from one function into the input of another function.  
Let's say we are only interested in the body mass of Gentoo penguins. We can use pipes to first select the species and body mass columns then filter that output only for Gentoo penguins. Notice that we only need to call the data one time.
```{r}
penguins %>%
  select(species, body_mass_g) %>% 
  filter(species == "Gentoo")
```
We can sort the output of `filter()` using the `arrange()` function. Here we filter for Gentoo penguins and then sort them by body mass in ascending order.
```{r}
penguins %>%
  select(species, body_mass_g) %>% 
  filter(species == "Gentoo") %>%
  arrange(body_mass_g)
```
To sort in descending order, wrap the variable name in `desc()`.
```{r}
penguins %>%
  select(species, body_mass_g) %>% 
  filter(species == "Gentoo") %>%
  arrange(desc(body_mass_g)) 
```
`slice()`
```{r}
penguins %>%
  select(species, body_mass_g) %>% 
  filter(species == "Gentoo") %>%
  slice_max(body_mass_g, n = 1) #slice_max gets largest bodymass, slice_min gets smallest 
```
## Practice
Let's use the fish data again to keep things simple.
1. We are interested in the radii length and scalelength of fish with the id `300`. Extract this information from the `fish` data. Please sort the data by radii length.
```{r}
names(fish)
```
```{r}
fish %>%
  select(fish_id, radii_length_mm, scalelength) %>% 
  filter(fish_id == 300) %>%
  arrange(radii_length_mm) 
```
2. We are interested in the fish from the lakes "AL" and "AR" with a radii length between 2 and 4. Extract this information from the `fish` data. Please sort the data in descending order by radii length.
```{r}
fish %>%
  select(lakeid, radii_length_mm) %>% 
  filter(lakeid == "AL" | lakeid == "AR") %>%
  filter(between(radii_length_mm, 2, 4)) %>%
  arrange(desc(radii_length_mm)) 
```

Hw 5:
## Data
**Database of vertebrate home range sizes.** 
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  
**1. Load the data into a new object called `homerange`.**
```{r}
homerange <- read.csv("data/Tamburelloetal_HomeRangeDatabase.csv")
```
**2. What is the structure of the `homerange` data? Does it have any NA's? Do a quick exploratory analysis of your choice below.**
```{r}
glimpse(homerange)
```
The homerange data does have NA values in "N", "preymass", "log10.preymass", "PPMR", and "prey.size.reference". 
```{r}
summary(homerange)
```
**3. Which animal groups are represented in the data? Make a plot that shows the number in each class.**
```{r}
names(homerange)
```
```{r}
ggplot(data = homerange,
      mapping = aes(x = class)) + 
      geom_bar(mapping = aes(fill = class)) +
      labs(title = "Number of Animal Groups in Each Class",
       x = "Class",
       y = "Number of Animal Groups")
```
**4. Are there more carnivores or herbivores represented in the data? Make a plot that shows the number in each trophic guild.**
```{r}
ggplot(data = homerange,
      mapping = aes(x = trophic.guild)) + 
      geom_bar(mapping = aes(fill = trophic.guild)) +
      labs(title = "Number of Animal Groups in Each Trophic Guild",
       x = "Trophic Guild",
       y = "Number of Animal Groups")
```
There are more carnivores in the data than herbivores. 
**5. Which herbivorous mammals have homeranges greater than 1,000,000 m²? Show only the common name, order, mean homerange size, and mean body mass. Arrange the data in descending order by homerange size and show only the top 10 species.**
```{r}
homerange %>%
  select(common.name, order, mean.hra.m2, mean.mass.g) %>% 
  filter(mean.hra.m2 >= 1000000) %>%
  arrange(desc(mean.hra.m2)) %>%
  slice_max(mean.hra.m2, n = 10)
```
**6. Which two species have the largest and smallest homerange sizes?**
```{r}
homerange %>%
  select(common.name, order, mean.hra.m2, mean.mass.g) %>% 
  arrange(desc(mean.hra.m2)) %>%
  slice_max(mean.hra.m2, n = 1)
```
```{r}
homerange %>%
  select(common.name, order, mean.hra.m2, mean.mass.g) %>% 
  arrange(mean.hra.m2) %>%
  slice_min(mean.hra.m2, n = 1)
```
**7. Which carnivorous mammals have the largest homeranges?**
```{r}
homerange %>%
  select(common.name, order, mean.hra.m2, mean.mass.g, trophic.guild) %>% 
  filter(trophic.guild == "carnivore") %>%
  arrange(desc(mean.hra.m2)) %>%
  slice_max(mean.hra.m2, n = 10)
```
**8. Is there a relationship between body mass and homerange size? Make a scatterplot to show this relationship. Be sure to include appropriate labels and a title.**
```{r}
ggplot(data = homerange,
      mapping = aes(x = mean.mass.g,y = mean.hra.m2)) + 
      geom_point(mapping = aes(color = trophic.guild)) + 
      geom_smooth(method = lm, se = F, na.rm = T) +
      scale_x_log10() +
      scale_y_log10() + 
      labs(title = "Body Mass (g) vs. Homerange Size (m^2)",
       x = "Body Mass (g)",
       y = "Homerange Size (m^2)")
```
There is a relationship between body mass and homerange size.
**9. Use filter to extract the data for only birds and mammals. Then make a plot that compares the homerange sizes of these two groups. Be sure to include appropriate labels and a title.**
```{r}
homerangebm = filter(homerange, taxon == "birds" | taxon == "mammals")

ggplot(data = homerangebm,
      mapping = aes(x = taxon, y = mean.hra.m2)) + 
      geom_boxplot(mapping = aes(fill = taxon)) +
      scale_y_log10() + 
      labs(title = "Taxon vs. Mean Homerange Size (m^2)",
       x = "Taxon",
       y = "Mean Homerange Size (m^2)")
```
**10. Do one additional exploratory analysis of your choice. Before writing the code, write the question you are trying to answer. The question should be biologically meaningful and appropriate for the data. It must include at least one `filter()`, `select()`, and `arrange()`.**
The question I am trying to answer is out of all of the bird species in the data, do carnivores have a larger homerange size than herbivores?
```{r}
homerange %>%
  select(common.name, taxon, order, family, trophic.guild, mean.mass.g, mean.hra.m2) %>% 
  filter(taxon == "birds") %>%
  arrange(desc(mean.hra.m2)) %>%
  slice_max(mean.hra.m2, n = 10)
  

homerangeb = filter(homerange, taxon == "birds")
ggplot(data = homerangeb,
      mapping = aes(x = trophic.guild, y = mean.hra.m2)) + 
      geom_boxplot(mapping = aes(fill = trophic.guild)) +
      scale_y_log10() + 
      labs(title = "Trophic Guild vs. Mean Homerange Size (m^2)",
       x = "Trophic Guild",
       y = "Mean Homerange Size (m^2)")
```

Lab 6:
## Load the tidyverse
```{r}
library("tidyverse")
library("janitor") #this will help with cleaning
library("palmerpenguins") #load the palmerpenguins package
options(scipen=999) #turn off scientific notation
```
1. Select species, island, and body mass from the penguins data. Arrange results by body mass.
```{r}
penguins %>%
  select(species, island, body_mass) %>%
  arrange(body_mass)
```
2. Filter the penguins data to only include observations from Biscoe and Dream islands.
```{r}
penguins %>%
  select(species, island, body_mass) %>%
  filter(island == "Biscoe" | island == "Dream")  %>% #ideally put select before filter
  arrange(body_mass)
```
```{r}
penguins %>% #alt method with concatenation
  select(species, island, body_mass) %>%
  filter(island == c("Biscoe", "Dream"))  %>% #ideally put select before filter
  arrange(body_mass)
```
```{r}
penguins %>% #alt method with concatenation + %in%
  select(species, island, body_mass) %>%
  filter(island %in% c("Biscoe", "Dream"))  %>% #ideally put select before filter
  arrange(body_mass)
```
3. Make a plot that shows the relationship between body mass and flipper length. How does this compare among different species?
```{r}
penguins %>%
  select(body_mass, flipper_len) %>%
  ggplot(data = penguins,
      mapping = aes(x = body_mass, y = flipper_len)) + 
      geom_point(mapping = aes(color = species)) + #can just be geom_point() for no colors
      geom_smooth(se = FALSE, method = lm) + 
      labs(title = "Body Mass (g) vs. Flipper Length (mm)",
           x = "Body Mass (g)",
           y = "Flipper Length (mm)")
```
## `distinct()`
`distinct()` looks for all unique observations in rows. This is a little tricky because it can look like it is working column-wise, but it is actually working row-wise.  
One helpful approach to new data is to find any duplicated rows. If we first look at the dimensions of the penguins data, we see it has 344 rows and 8 columns.
Using `distinct()` across all rows, we see there are no duplicates. This means every row contains unique observations across all variables.
```{r}
penguins %>%
  distinct() #looks across dataframe and compares every row, if there's a row thats the same as the other one it will flag them to ignore in new output
```
But if we only look at species, we can see that there are only 3 unique species in the data.
```{r}
penguins %>%
  distinct(species)
```
What if we want to know which islands each species occurs on?
```{r}
penguins %>%
  distinct(species, island, .keep_all=T) #.keep_all shows all other variables 
```
## `mutate()`
`mutate()` is another verb that acts on columns. It allows us to create new columns from existing columns in a data frame. When we use `mutate()`, the columns are added to the end of the dataframe by default. Let's create a new column that converts body mass from grams to kilograms.
```{r}
penguins %>%
  mutate(body_mass_kg = body_mass / 1000) %>%
  select(species, body_mass, body_mass_kg) %>%
  arrange(body_mass_kg)
```
## `mutate()` and `across()`
We use `across()` within `mutate()` to apply a function to multiple columns. This is especially helpful when cleaning data. For example, let's say we want to convert all columns that end with `mm` to centimeters. We can use `across()` to do this.
```{r}
penguins %>%
  mutate(across(ends_with("len"), ~./10)) %>%
  select(species, bill_length_cm = bill_len, bill_depth_cm = bill_dep, flipper_length_cm = flipper_len) 
```
What does the `~./10` mean? The `~` indicates that what follows is a formula (lambda function). The `.` represents the current column being processed. So, `./10` means "take the current column and divide it by 10". This operation is applied to all columns that end with `mm`.
## Cleaning Data
Cleaning raw data is an essential, but tedious step in data analysis. It's impossible to predict every scenario that you will come across, but there are some common issues that we can learn to address.
We already learned how to use `rename()` to change column names. We also learned how to rename columns from within `select()`. But, this can be very inefficient if we have a large dataset.
Let's have a look at some new data focused on mammal lifehistories. The data are from: S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.   [link](http://esapubs.org/archive/ecol/E084/093/)  
```{r}
mammals <- read.csv("data/mammal_lifehistories_v2.csv") #%>% clean_names() *Note: clean_names() done immediately if possible
```
One thing to notice is the column names are inconsistent. This is going to cause problems for us down the line. We could rename each column, one at a time, using `rename()`, but that would be tedious. Instead, we can use the `clean_names()` function from the `janitor` package to fix all of the column names at once.
```{r}
mammals <- mammals %>%
  clean_names() 
```
Notice that `clean_names()` has converted all column names to lowercase and replaced spaces with underscores. But, no adjustments were made to the data itself. What if we want to change observations from upper case to lower case?
```{r}
mammals %>%
  mutate(across(c("order","family"), tolower)) #tolower makes specific columns all lowercase
```
This will change all columns to lower case. But, notice what happens to numeric columns.
```{r}
mammals %>%
  mutate(across(everything(), tolower)) #everything in database made lowercase, numerics turned into characters however, best suited for only character-based dataframes/categorical
```
For this reason, it might be better to use `where` so we can specify only character columns.
```{r}
mammals %>%
  mutate(across(where(is.character), tolower))
```
## `if_else()`
We briefly introduce `if_else()` here because it allows us to use `mutate()` but not have the entire column affected in the same way. With `ifelse()`, you first specify a logical statement, afterwards what needs to happen if the statement returns `TRUE`, and lastly what needs to happen if it's  `FALSE`.  
Have a look at the data from mammals below. Notice that the values for newborn include `-999.00`. This is sometimes used as a placeholder for NA (but, is a really bad idea). We can use `if_else()` to replace `-999.00` with `NA`.  
```{r}
mammals %>%
  select(genus, species, newborn) %>%
  arrange(newborn)
```
```{r}
mammals %>%
  select(genus, species, newborn) %>%
  mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))
```
1. Following the example above, convert all -999 values in the mammals dataframe to NA.
```{r}
mammals <- mammals %>%
  mutate(across(c(mass, wean_mass, gestation, max_life, newborn, weaning, litter_size, afr, litters_year),  ~ifelse(.== -999.00, NA, .)))

  #mutate(across(everything(), tolower)) #everything in database made lowercase, numerics turned into characters however, best suited for only character-based dataframes/categorical
```
2. In the mammals data, make a new column `mass_kg` that that converts mass from grams to kilograms. Select the columns genus, species, mass, and mass_kg, and arrange the data by mass_kg in descending order. What is the common name for the species with the highest mass?  
```{r}
mammals %>%
  mutate(mass_kg = mass / 1000) %>%
  select(genus, species, mass, mass_kg) %>%
  arrange(desc(mass_kg))
```
Common Name is Blue Whale. 
3. What is the relationship between gestation and newborn mass?
```{r}
mammals %>%
  select(gestation, newborn) %>%
  ggplot(data = mammals,
         mapping = aes(x = log10(gestation), y = log10(newborn))) +
         geom_point() + 
        # geom_smooth(se = F, method = lm) + 
         labs(title = "Gestation vs. Newborn Mass", 
              x = "Gestation", 
              y = "Newborn Mass")
```
4. Which mammal has the longest life span in years?  
```{r}
mammals %>%
  mutate(max_life_years = max_life / 12)  %>% #instructor example has mutate and select lines swapped)
  select(genus, species, max_life, max_life_years) %>%
  arrange(desc(max_life_years))
```
5. Do larger mammals live longer?
```{r}
mammals %>%
  ggplot(mapping = aes(x = log10(mass), y = max_life)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

Hw 6:
**1. Load `IvindoData_DryadVersion.csv` and store it as a new object called `gabon`.**
```{r}
gabon <- read.csv("data/IvindoData_DryadVersion.csv")
```
**2. Use one or more of the summary functions you have learned to get an idea of the structure of the data.**  
```{r}
glimpse(gabon)
```
```{r}
summary(gabon)
```
**3. Clean the data as necessary following the conventions from lab 6.**
```{r}
gabon <- gabon %>%
  clean_names() %>%
  mutate(across(where(is.character),tolower))  
```
**4. Use `mutate()` Change the variables `HuntCat`, `LandUse`, and `TransectID` to factors. Be sure to store the output as a new dataframe and use it for the remaining questions.**
```{r}
gabon <- gabon %>%
  mutate(across(c(hunt_cat, land_use, transect_id), as.factor))
```
**5. Make a plot that shows the number of transects for each land use type.**
```{r}
gabon %>%
  ggplot(mapping = aes(x = land_use)) +
    geom_bar(mapping = aes(fill = land_use)) +
    labs(title = "Number of Transects for Each Land Use Type",
      x = "Land Use Type",
      y = "Number of Transects")
```
**6. How does vegetation richness compare among land use types?**
```{r}
gabon %>%
  ggplot(mapping = aes(x = land_use, y = veg_rich)) +
    geom_boxplot(mapping = aes(fill = land_use)) +
    labs(title = "Vegetation Richness (# Species) Among Land Use Types",
      x = "Land Use Type",
      y = "Vegetation Richness (# Species)")
```
**7. For which land use type (national parks, logging, or neither) is average all species diversity the greatest?**
```{r}
gabon %>%
  ggplot(mapping = aes(x = land_use, y = diversity_all_species)) +
    geom_boxplot(mapping = aes(fill = land_use)) +
    labs(title = "All Species Diversity Among Land Use Types",
      x = "Land Use Type",
      y = "All Species Diversity")
```
**8. Is there a relationship between vegetation richness and mammal species richness?**
```{r}
gabon %>%
  ggplot(mapping = aes(x = veg_rich, y = rich_mammal_species)) + 
    geom_point(mapping = aes(color = land_use)) + 
    geom_smooth(method = lm, se = F, na.rm = T) +
    labs(title = "Vegetation Richness (# Species) vs. Mammal Species Richness (# Species)",
      x = "Vegetation Richness (# Species)",
      y = "Mammal Species Richness (# Species)")
```
There is no relationship between vegetation richness and mammal species richness (line of best fit doesn't fit well with the data plotted. 
**9. Use `filter` to find all transects that have greater than 15 tree species or a breast height diameter between 50 and 60cm.  **
```{r}
gabon %>%
  select(transect_id, land_use, veg_rich, veg_dbh) %>%
  filter(veg_rich >= 15 | between(veg_dbh, 50, 60)) %>%
  arrange(desc(veg_dbh))
```
**10.Which transects and land use types have more than 10 tree species and 10 mammal species? Use `arrange()` to sort by the number of tree species.**
```{r}
gabon %>%
  select(transect_id, land_use, veg_rich, rich_mammal_species) %>%
  filter(veg_rich >= 10 & rich_mammal_species >= 10) %>%
  arrange(desc(veg_rich))
```

Lab 7:
## Summary statistics
When first exploring data, it is often a good idea to produce some basic summary statistics. This can help you identify potential problems with the data (e.g., outliers, missing values, skew) as well as get a general sense of the distribution of values for a given variable. So far, you have learned to calculate summary statistics using _base R_ functions such as `mean()`, `sd()`, `min()`, and `max()`.  
```{r}
summary(msleep)
```
## Practice
1. Calculate the mean body weight of large and small mammals in msleep. Large mammals have a body weight greater than 150, while small mammals have a body weight less than 10. I did large mammals for you as an example.  
```{r}
large <- msleep %>% 
  filter(bodywt > 150) %>% 
  arrange(desc(bodywt))
large
```
```{r}
small <- msleep %>% 
  filter(bodywt < 10) %>% 
  arrange(bodywt)
small
```
```{r}
largem <- mean(large$bodywt)
largem
smallm <- mean(small$bodywt)
smallm
```
## `summarize()`
Fortunately, dplyr has some functions that make producing summary statistics easier. The verb `summarize()` produces summary statistics for a given variable in a data frame. Let's try the examples above using `summarize()`. 
Large mammals
```{r}
msleep %>%
  filter(bodywt > 150) %>%
  summarize(mean_bodywt_large = mean(bodywt)) 
```
Small mammals
```{r}
msleep %>%
  filter(bodywt < 10) %>%
  summarize(mean_bodywt_small = mean(bodywt)) 
```
You can also combine functions to make summaries for multiple variables.
```{r}
msleep %>%
  filter(bodywt > 150) %>%
  summarize(mean_bodywt_large = mean(bodywt), 
            min_bodywt_large = min(bodywt),
            max_bodywt_large = max(bodywt),
            sd_bodywt_large = sd(bodywt),
            total = n()) #number of observations
```
Maybe you want to summarize all of the numeric variables in the data frame. You can use the `where()` function to select only the numeric variables.
```{r}
msleep %>%
  select(where(is.numeric)) %>%
  summarize_all(mean)
```
Don't forget to remove NA's.
```{r}
msleep %>%
  select(where(is.numeric)) %>%
  summarize_all(mean, na.rm = T)
```
This is a little more efficient.
```{r}
msleep %>%
  summarize(across(where(is.numeric),
                   ~mean(.x, na.rum = T))) #.x means for every cell
```
## Practice
1. What is the mean, min, and max `bodywt` for the taxonomic order Primates? Provide the total number of observations.
```{r}
msleep %>%
  filter(order == "Primates") %>%
  summarize(mean_bodywt_p = mean(bodywt), 
            min_bodywt_p = min(bodywt),
            max_bodywt_p = max(bodywt),
            total = n()) #number of observations
```
`n_distinct()` is a handy way of cleanly presenting the number of distinct observations. Notice that there are 9 genera with over 100 in body weight.
```{r}
msleep %>%
  filter(bodywt > 150) %>%
  distinct()
```
Here we show the number of distinct genera over 100 in body weight.
```{r}
msleep %>%
  filter(bodywt > 100) %>%
  summarize(n_genara = n_distinct(genus))
```
Not to be confused with n(), which counts the total number of observations.
```{r}
msleep %>%
  filter(bodywt > 100) %>%
  summarize(total = n())
```
`pull()` is another useful verb that extracts a single column from a data frame as a vector.
```{r}
msleep %>%
  filter(bodywt > 100) %>%
  distinct() %>%
  pull(genus)
```
## Practice
1. How many genera are represented in the msleep data frame?
```{r}
msleep %>%
  summarize(n_genara = n_distinct(genus))
```
## Counts
Although `summarize()` is helpful, oftentimes we are only interested in counts. The [janitor package](https://garthtarr.github.io/meatR/janitor.html) does a lot with counts, but there are also functions that are part of dplyr that are useful.  
For example, if we wanted to count the number of observations for each `vore` category in the `msleep` data frame we could do the following:
```{r}
msleep %>%
  count(vore)
```
```{r}
msleep %>%
  count(vore, sort = T) %>% #sort = T sorts in descending order
  ggplot(aes(x = vore, y = n)) + 
  geom_col(mapping = aes(fill = vore)) + #use geom_col if we have an x and a y variable 
  labs(title = "Number of Species per Vore",
  x = "Vore",
  y = "Number of Species")
```
You can even combine multiple variables to get counts. For example, if we wanted to count the number of observations for each `vore` category within each `order` category we could do the following:
```{r}
msleep %>%
  count(vore, order)
```
The tabyl() function from the janitor package is also very useful for producing counts and percentages.  
```{r}
msleep %>%
  tabyl(vore)
```
## Practice
1. In the taxonomic order Carnivora, count the number of observations for each conservation status.
```{r}
msleep %>%
  filter(order == "Carnivora") %>%
  count(conservation, sort = T)
```
2. Among herbivores, which order is most represented? Return the top 5 orders with counts, sorted in descending order.
```{r}
msleep %>%
  filter(vore == "herbi") %>%
  count(order, sort = T) %>%
  slice_max(n, n = 5)
```

Hw 7:
**1. Before starting data analysis, read the abstract of the paper to get an idea of the questions being asked. In 2-3 sentences, describe what the study is testing and the variables involved.**  
The study interest of the paper was looking ito whether severe historical elepahnt harvests eliminated large tuskers among survivors and whether elephants born thereafter had smaller tusks. 
The paper looks into the tusk size of elephants sampled in different time periods, and also categorized the elephants by sex, age, tusk circumference, and length. 
**2. Load `elephants.csv` and store it as a new object called `elephants`.**  
```{r}
elephants <- read.csv("data/elephants.csv")
```
**3. Clean the data by converting variable names to lowercase with no spaces or special characters.**  
```{r}
elephants <- elephants %>%
  clean_names() %>%
  mutate(across(where(is.character),tolower))  
```
**4. Use one or more of the summary functions you have learned to get an idea of the structure of the data.**   
```{r}
glimpse(elephants)
```
```{r}
summary(elephants)
```
**5. Use `mutate()` Change the variables `years_of_sample_collection`, `elephant_id`, and `sex` to factors. Be sure to store the output as a new dataframe and use it for the remaining questions.**  
```{r}
elephants <- elephants%>%
  mutate(across(c(years_of_sample_collection, elephant_id, sex), as.factor))
```
**6. From which years were data collected? Show the sample periods below.**  
```{r}
elephants %>%
  select(years_of_sample_collection) %>%
  count(years_of_sample_collection)
```
**7. How many males and females were sampled in this study?**  
```{r}
elephants %>%
  select(sex) %>%
  count(sex)
```
**8. What is the mean, median, and standard deviation for age of males and females included in the study? Separate the results by year of sample collection. Does the sampling look even between years and sexes?**  
```{r}
elephants %>%
  filter(years_of_sample_collection == "1966-68",
         sex == "m") %>%
  summarize(mean_age_m = mean(estimated_age_years, na.rm = T), 
            min_age_m = min(estimated_age_years, na.rm = T),
            max_age_m = max(estimated_age_years, na.rm = T),
            sd_age_m = sd(estimated_age_years, na.rm = T),
            total = n()) 
```
```{r}
elephants %>%
  filter(years_of_sample_collection == "1966-68",
         sex == "f") %>%
  summarize(mean_age_f = mean(estimated_age_years, na.rm = T), 
            min_age_f = min(estimated_age_years, na.rm = T),
            max_age_f = max(estimated_age_years, na.rm = T),
            sd_age_f = sd(estimated_age_years, na.rm = T),
            total = n()) 
```
```{r}
elephants %>%
  filter(years_of_sample_collection == "2005-13",
         sex == "m") %>%
  summarize(mean_age_m = mean(estimated_age_years, na.rm = T), 
            min_age_m = min(estimated_age_years, na.rm = T),
            max_age_m = max(estimated_age_years, na.rm = T),
            sd_age_m = sd(estimated_age_years, na.rm = T),
            total = n()) 
```
```{r}
elephants %>%
  filter(years_of_sample_collection == "2005-13",
         sex == "f") %>%
  summarize(mean_age_f = mean(estimated_age_years, na.rm = T), 
            min_age_f = min(estimated_age_years, na.rm = T),
            max_age_f = max(estimated_age_years, na.rm = T),
            sd_age_f = sd(estimated_age_years, na.rm = T),
            total = n()) 
```
**9. Is age (independent variable) a positive predictor of tusk length (dependent variable)? Create a plot that shows the relationship between these variables and add a linear model fit line.**  
```{r}
elephants %>%
  ggplot(mapping = aes(x = estimated_age_years, y = tusk_length_in_cm)) +
  geom_point(mapping = aes(color = sex)) + 
  geom_smooth(method = lm, se = F, na.rm = T) + 
  labs(title = "Age vs. Tusk Length (cm)",
      x = "Age",
      y = "Tusk Length (cm)")
```
Based on the plot, age is a positive predictor of tusk length.
**10. Is shoulder height (independent variable) a positive predictor of tusk length (dependent variable)? Create a plot that shows the relationship between these variables and add a linear model fit line.**  
```{r}
elephants %>%
  ggplot(mapping = aes(x = shoulder_height_in_cm, y = tusk_length_in_cm)) +
  geom_point(mapping = aes(color = sex)) + 
  geom_smooth(method = lm, se = F, na.rm = T) + 
  labs(title = "Shoulder Height (cm) vs. Tusk Length (cm)",
      x = "Shoulder Height (cm)",
      y = "Tusk Length (cm)")
```
Shoulder height is a positive predictor of tusk length. 
**11. The authors argue that because poachers preferentially target elephants with large tusks, this has resulted in a decrease in average tusk length. Is this supported by the data? Show your code and calculations below.**  
```{r}
elephants %>%
  ggplot(mapping = aes(x = years_of_sample_collection, y = tusk_length_in_cm)) +
  geom_boxplot(mapping = aes(color = years_of_sample_collection)) + 
  labs(title = "Time Period vs. Tusk Length (cm)",
      x = "Time Period", 
      y = "Tusk Length (cm)") #for this problem, better to not plot and instead use means/max/min/sd 
```
The data does support that because poachers preferentially target elephants with large tusks, this has resulted in a decrease in average tusk length
```{r}
elephants %>%
    filter(years_of_sample_collection == "1966-68") %>%
  summarize(mean_tusk_length = mean(tusk_length_in_cm, na.rm = T), 
            min_tusk_length = min(tusk_length_in_cm, na.rm = T),
            max_tusk_length = max(tusk_length_in_cm, na.rm = T),
            total = n()) 
```
```{r}
elephants %>%
    filter(years_of_sample_collection == "2005-13") %>%
  summarize(mean_tusk_length = mean(tusk_length_in_cm, na.rm = T), 
            min_tusk_length = min(tusk_length_in_cm, na.rm = T),
            max_tusk_length = max(tusk_length_in_cm, na.rm = T),
            total = n()) 
```
**12. Male elephants reach effective sexual maturity at 25 years while females are sexually mature at 12 years. Make a new dataframe that extracts only the males and females at sexual maturity. Then, make a plot that shows the range of tusk length between the two sample periods for these mature elephants.**   
```{r}
elephants_sm <- elephants %>%
  filter((sex == "m" & estimated_age_years >= 25) | (sex == "f" & estimated_age_years >= 12)) #must be isolated from the code chunk with ggplot for ggplot to utilize the new dataframe created, done to help with pipe management (doesn't mix up piping and ggplot functions)
```
```{r}
elephants_sm %>% 
  ggplot(mapping = aes(x = years_of_sample_collection, y = tusk_length_in_cm)) +
  geom_boxplot(mapping = aes(color = sex)) + 
  labs(title = "Time Period vs. Tusk Length (cm) for Males and Females at Sexual Maturity",
      x = "Time Period",
      y = "Tusk Length (cm)") #just present this graph
```

Lab 8_1:
## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```
## Load the data
For this lab, let's go back to the elephants data from homework 7. The data are from: [Chiyo, Patrick I., Vincent Obanda, and David K. Korir. “Illegal tusk harvest and the decline of tusk size in the African elephant.” Ecology and Evolution 5, 22: 5216–5229 (2015)](https://doi.org/10.1002/ece3.1769). Data deposited at [Dryad Digital Repository](https://doi.org/10.5061/dryad.h6t7j). 
```{r message=FALSE, warning=FALSE}
elephants <- read_csv("data/elephants.csv") %>% 
  clean_names() %>% 
  mutate(across(where(is.character), as.factor))
```
## Review
`summarize()` is a verb in `dplyr` that allows us to produce summaries of data. It's output is a new dataframe. In lab 7, we learned how `summarize()` can be used to produce quick statistical summaries.  
1. What is the average age, shoulder height, tusk length, and tusk circumference for all elephants in the data set?
```{r}
elephants %>%
  summarize(average_age = mean(estimated_age_years, na.rm = T),
            average_shoulder_height = mean(shoulder_height_in_cm, na.rm = T),
            average_tusk_length = mean(tusk_length_in_cm, na.rm = T),
            average_tusk_circumference = mean(tusk_circumference_in_cm, na.rm = T))
# or summarize(across(where(is.numeric), ~mean.x, na.rm = T)))
```
We can include multiple functions using this same structure, but it becomes cumbersome.
```{r}
elephants %>% 
  summarize(
    across(
      where(is.numeric),
      list(
        mean   = ~mean(.x, na.rm = TRUE),
        median = ~median(.x, na.rm = TRUE),
        sd     = ~sd(.x, na.rm = TRUE),
        min    = ~min(.x, na.rm = TRUE),
        max    = ~max(.x, na.rm = TRUE),
        n      = ~sum(!is.na(.x))
      )
    )
  )
```
## `group_by()`
The `summarize()` function is often more useful when used with `group_by()`. `group_by()` allows us to group data by a categorical variable(s) of interest.   
```{r}
elephants %>%
  group_by(sex) 
```
Notice that the output doesn't look different, but two groups are indicated. Any verbs that follow `group_by()` will be applied to each group separately. For example, what is the average age, shoulder height, tusk length, and tusk circumference for males and females?
```{r}
elephants %>%
  group_by(sex) %>%
  summarize(across(where(is.numeric), ~mean(.x, na.rm = T)))
```
We can group by more than one variable.
```{r}
elephants %>%
  group_by(years_of_sample_collection, sex) %>%
  summarize(mean_estimated_age_years = mean(estimated_age_years, na.rm = T), n = n(), .groups = 'keep') #.groups removes error message 
```
Do you remember this question from homework 7?  
"What is the mean, median, and standard deviation for age of males and females included in the study? Separate the results by year of sample collection. Does the sampling look even between years and sexes?"
Without `group_by()`, we had to produce multiple code chunks- we are now more efficient!
```{r}
elephants %>%
  filter(years_of_sample_collection == "1966-68") %>%
  group_by(sex) %>%
  summarize(mean_age = mean(estimated_age_years, na.rm = T),
            median_age = median(estimated_age_years, na.rm = T),
            sd_age = sd(estimated_age_years, na.rm = T))
```
```{r}
elephants %>%
  filter(years_of_sample_collection == "2005-13") %>%
  group_by(sex) %>%
  summarize(mean_age = mean(estimated_age_years, na.rm = T),
            median_age = median(estimated_age_years, na.rm = T),
            sd_age = sd(estimated_age_years, na.rm = T))
```
```{r}
elephants %>%
  group_by(sex, years_of_sample_collection) %>%
  summarize(mean_age = mean(estimated_age_years, na.rm = T),
            median_age = median(estimated_age_years, na.rm = T),
            sd_age = sd(estimated_age_years, na.rm = T))
```
## Practice
Use the built-in `msleep` data to answer the following questions.  
1. Calculate the min, max, mean, and total number of observations for body weight by vore.  
```{r}
msleep <- msleep
```
```{r}
names(msleep)
```
```{r}
msleep %>%
  group_by(vore) %>%
  summarize(mean_bodywt = mean(bodywt, na.rm = T),
            median_bodywt = median(bodywt, na.rm = T),
            sd_bodywt = sd(bodywt, na.rm = T),
            n = n())
```
2. Calculate mean brain weight by taxonomic order in the msleep data.
```{r}
msleep %>%
  group_by(order) %>%
  summarize(mean_brainwt = mean(brainwt, na.rm = T))
```
3. Try running the code again, but this time add `na.rm=TRUE`. What is the problem with Cetacea?
```{r}
msleep %>%
  group_by(order) %>%
  summarize(mean_brainwt = mean(brainwt, na.rm = T))
```

Hw 8: (load tidyverse, janitor, scipen = 999)
Load the data `fisheries_clean.csv` as a new object titled `fisheries_clean`.
```{r message=FALSE, warning=FALSE}
fisheries_clean <- read_csv("data/fisheries_clean.csv")
```
1. Explore the data. What are the names of the variables, what are the dimensions, are there any NA's, what are the classes of the variables, etc.? You may use the functions that you prefer.
```{r}
glimpse(fisheries_clean)
```
```{r}
names(fisheries_clean)
```
```{r}
dim(fisheries_clean)
```
2. Convert the following variables to factors: `period`, `continent`, `geo_region`, `country`, `scientific_name`, `common_name`, `taxonomic_code`, and `status`.
```{r}
fisheries_clean <- fisheries_clean %>%
  mutate(across(c(period, continent, geo_region, country, scientific_name, common_name, taxonomic_code), as.factor))
fisheries_clean
```
4. How many countries are represented in the data?
```{r}
fisheries_clean %>%
  summarize(n_country = n_distinct(country))
```
5. The variables `common_name` and `taxonomic_code` both refer to species. How many unique species are represented in the data based on each of these variables? Are the numbers the same or different?
```{r}
fisheries_clean %>%
  group_by(common_name, taxonomic_code) %>%
  summarize(common_name, taxonomic_code, .groups = 'keep') %>%
  count(common_name) 
```
```{r}
fisheries_clean %>%
  summarize(n_species = n_distinct(taxonomic_code))
```
6. In 2023, what were the top five countries that had the highest overall catch?
```{r}
fisheries_clean %>% #group by country because you use summarize later, take sum of the total catch for 7 (make a new column)
  filter(period == 2023) %>%
  group_by(country) %>%
  summarize(total_catch = sum(catch, na.rm = T)) %>%
  arrange(desc(total_catch)) %>%
  slice_max(total_catch, n = 5)
```
7. In 2023, what were the top 10 most caught species? To keep things simple, assume `common_name` is sufficient to identify species. What does `NEI` stand for in some of the common names? How might this be concerning from a fisheries management perspective?
```{r}
fisheries_clean %>%
  filter(period == 2023) %>%
  group_by(common_name) %>%
  summarize(total_catch = sum(catch, na.rm = T)) %>%
  arrange(desc(total_catch)) %>%
  slice_max(total_catch, n = 10)
```
NEI stands for "Not Elsewhere Included" (or Indicated), where catches are reported that they cannot be precisely identified at the species level (like the specific species beyond marine fishes nei for example cannot be identified). This is concerning from a fisheries management perspective because this tracking is important for maintaining the sustainability of the ecosystem/environment and knowing when to catch/not catch certain species. 
8. For the species that was caught the most above (not NEI), which country had the highest catch in 2023?
```{r}
fisheries_clean %>%
  filter(period == 2023, common_name == "Alaska pollock(=Walleye poll.)") %>% #Alaska pollock is 3rd for all reported species, 1st of non NEI species reported 
  group_by(country) %>%
  summarize(total_catch = sum(catch, na.rm = T)) %>%
  arrange(desc(total_catch)) 
```
For the species that was caught the most above (Alaska pollock (Walleye poll.)), the country that had the highest catch in 2023 was the Russian Federation.
9. How has fishing of this species changed over the last decade (2013-2023)? Create a  plot showing total catch by year for this species.
```{r}
fisheries_clean_walleyepoll <- fisheries_clean %>%
  filter(common_name == "Alaska pollock(=Walleye poll.)", period %in% c(2013:2023))
```
```{r}
fisheries_clean_walleyepoll %>% 
  ggplot(mapping = aes(x = period, y = catch)) +
  geom_col(mapping = aes(fill = continent)) + 
  labs(title = "Total Catch of The Alaska Pollock Species per Year",
      x = "Year",
      y = "Total Catch") 
```

Lab 9_1:
## Tidy data
Most "wild" data are organized incorrectly for work in R and, as you might expect, the tools used to transform data are a core part of the tidyverse. I will attempt to summarize the most important points below, but you should read [chapter 5 of the data science text](https://r4ds.hadley.nz/data-tidy.html) for a more thorough explanation.  
`Tidy` data follows three conventions:   
(1) each variable has its own column  
(2) each observation has its own row  
(3) each value has its own cell  
## `pivot_longer()`  
Scientists use spreadsheets that are organized to make data entry efficient. This is often referred to as **wide format**. Unfortunately, the wide format creates a problem because column names may actually represent values of a variable. The command `pivot_longer()` shifts data from wide to long format.   
Rules:  
+ `pivot_longer`(cols, names_to, values_to)
+ `cols` - Columns to pivot to longer format
+ `names_to` - Name of the new column; it will contain the column names of gathered columns as values
+ `values_to` - Name of the new column; it will contain the data stored in the values of gathered columns
## Example 1: column names are data
The following data show results from a drug trial with four treatments on six patients. The values represent resting heart rate.
```{r}
heartrate <- read_csv("data/heartrate.csv")
heartrate
```
Let's assess whether or not these data are tidy.  
(1) each variable has its own column?  
 Each variable does not have its own column.
(2) each observation has its own row?  
Each observation does not have its own row.
(3) each value has its own cell?  
Each value does not have its own cell. 
Let's think about this another way- can we make a plot that shows the effect of each drug on Margaret's heart rate?  No, the data are not tidy.  
```{r}
heartrate %>%
  filter(patient == "Margaret") %>%
  ggplot(mapping = aes(x = a, y = patient)) +
  geom_boxplot(mapping = aes(color = a)) 
```
To fix this problem, we need to reshape the table to long format while keeping track of column names and values. We do this using `pivot_longer()`. Notice that the dimensions of the data frame change.  
```{r}
heartrate %>%
  pivot_longer(-patient,
               names_to = "drug",
               values_to = "heartrate")
```
```{r}
heartrate %>%
  pivot_longer(-patient,
               names_to = "drug",
               values_to = "heartrate") %>%
  filter(patient == "Margaret") %>%
  ggplot(mapping = aes(x = drug, y = heartrate)) +
  geom_col() 
```
## Practice
1. Import the file `relig_income.csv` and store it as a new object `relig_income`.  
```{r}
relig_income <- read.csv("data/relig_income.csv")
```
3. Use `pivot_longer()` to make the data tidy.  
```{r}
relig_income %>%
  pivot_longer(-religion, #don't move
               names_to = "income", #move to column
               values_to = "total") #move to column
```
## Example 2: some column names are data
Some (but not all) of the column names are data. We also have NA's.
```{r}
billboard <- read_csv("data/billboard.csv")
billboard
```
Solution 1: specify a range of columns that you want to pivot.
```{r}
billboard %>%
  pivot_longer(wk1:wk76,
               names_to = "week",
               values_to = "rank",
               values_drop_na = T)
```
Solution 2: OR, specify columns that you want to stay fixed.
```{r}
billboard %>%
  pivot_longer(-c(artist, track, date.entered), #not numerically ordered 
               names_to = "week",
               values_to = "rank",
               values_drop_na = T)
```
Solution 3: identify columns by a prefix, remove the prefix and all NA's.
```{r}
billboard %>%
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    names_prefix = "wk", #gets rid of the wk in each entry like wk1, wk2, wk3, etc.
    values_to = "rank",
    values_drop_na = T)
```
## Practice  
1. Import `plant_data.csv` as a new object `plant_data`.  
```{r}
plant_data <- read_csv("data/plant_data.csv")
```
2. Why are these data not tidy?  
Variable names are actually data entries.
3. Use `pivot_longer()` to make the data tidy. Focus the data only on genotype, water_sched_prog, and greenhouse.  
```{r}
plant_data %>%
  pivot_longer(day1:day30,
               names_to = "day",
               names_prefix = "day",
               values_to = "idk",
               values_drop_na = T)

 # pivot_longer(-c(genotype, water_sched_prog, greenhouse),
    #           names_to = "day",
     #          names_prefix = "day",
     #          values_to = "idk",
      #         values_drop_na = T)
```
## Example 3: more than one variable in a column name
In this case, there are two observations in each column name.
```{r}
qpcr_untidy <- read_csv("data/qpcr_untidy.csv")
qpcr_untidy
```
`names_sep` helps pull these apart, but we still have "exp" and "rep" to deal with.  
```{r}
qpcr_untidy <- qpcr_untidy %>% 
  pivot_longer(
    exp1_rep1:exp3_rep3,
    names_to = c("experiment", "replicate"),
    names_sep = "_",
    values_to = "mRNA expression"
  )
```
```{r}
qpcr_untidy %>%
  separate(experiment, into = c("experiment", "e_number"), sep = 3) %>%
  separate(replicate, into = c("replicate", "r_number"), sep = 3) %>%
  select(gene, experiment = e_number, replicate = r_number, 'mRNA expression')
```

Lab 9_2:
## Review
Last time we learned the fundamentals of tidy data and used the `pivot_longer()` function to wrangle a few examples of frequently encountered untidy data. In the second part of today's lab we will look at a few more examples of `pivot_longer()` but also use the `pivot_wider()` function to deal with another type of untidy data.  
## Load the tidyverse
```{r}
library("tidyverse")
```
## `pivot_longer()`
Recall that we use `pivot_longer()` when our column names actually represent variables. A classic example would be that the column names represent observations of a variable.
```{r}
datasets::USPersonalExpenditure
?USPersonalExpenditure
```
Here we add a new column of expenditure types, which are stored as rownames above, with `mutate()`. The `USPersonalExpenditures` data also needs to be converted to a data frame before we can use the tidyverse functions, because it comes as a matrix.
```{r}
expenditures <- as_tibble(USPersonalExpenditure) %>% 
  mutate(expenditure = rownames(USPersonalExpenditure)) %>% 
  relocate(expenditure)
expenditures
```
## Practice
1. Are these data tidy? Please use `pivot_longer()` to tidy the data.
```{r}
expenditures %>%
  pivot_longer(-expenditure,
    names_to = "year",
    values_to = "bn_dollars")
```
2. Restrict the data to medical and health expenditures only. Sort in ascending order.
```{r}
expenditures %>%
  pivot_longer(-expenditure,
    names_to = "year",
    values_to = "bn_dollars") %>%
    filter(expenditure == "Medical and Health") %>%
    arrange(bn_dollars)
```
## `separate()`
In this new heart rate example, we have the sex of each patient included with their name. Are these data tidy? No, there is more than one value per cell in the patient column and the columns a, b, c, d once again represent values.
```{r}
heartrate2 <- read_csv("data/heartrate2.csv")
heartrate2
```
We need to start by separating the patient names from their sexes. `separate()` needs to know which column you want to split, the names of the new columns, and what to look for in terms of breaks in the data.
```{r}
heartrate2 %>% 
  separate(patient, into = c("patient", "gender"), sep = "_")
```
Another way of using separate is to specify the position of the split. In this case, we want to split the patient column at the last character.
```{r}
heartrate2 %>% 
  separate(patient, into = c("patient", "gender"), sep = -1)
```
We would still need to deal with the "_" in the patient column. We can do this using the stringr package.
```{r}
heartrate2 %>% 
  separate(patient, into = c("patient", "gender"), sep = -2)
```
```{r}
heartrate2 %>% 
  separate(patient, into = c("patient", "gender"), sep = -1) %>% 
  mutate(patient = str_remove(patient, "_"))
```
## Practice
1. Re-examine `heartrate2`. Use `separate()` for the sexes, `pivot_longer()` to tidy, and `arrange()` to organize by patient and drug. Store this as a new object `heartrate3`.  
```{r}
heartrate3 <- heartrate2 %>% 
  separate(patient, into = c("patient", "gender"), sep = -1) %>% 
  mutate(patient = str_remove(patient, "_")) %>% 
  pivot_longer(-c("patient", "gender"),
               names_to = "drug",
               values_to = "heartrate") %>%
  arrange(desc(patient))
heartrate3
```
2. `unite()` is the opposite of separate(). Its syntax is straightforward. You only need to give a new column name and then list the columns to combine with a separation character.  Give it a try below by recombining patient and sex from `heartrate3`.  
```{r}
heartrate3 %>% 
  unite(patient_gender, "patient", "gender", sep = "_")
```
## `pivot_wider()`
The opposite of `pivot_longer()`. You use `pivot_wider()` when you have an observation scattered across multiple rows. In the example below, `cases` and `population` represent variable names not observations.  
Rules:  
+ `pivot_wider`(names_from, values_from)  
+ `names_from` - Values in the `names_from` column will become new column names  
+ `values_from` - Cell values will be taken from the `values_from` column  
```{r}
tb_data <- read_csv("data/tb_data.csv")
tb_data
```
When using `pivot_wider()` we use `names_from` to identify the variables (new column names) and `values_from` to identify the values associated with the new columns.
```{r}
tb_data %>% 
  pivot_wider(
    names_from = "key",
    values_from = "value"
  )
```
## Practice
1. Load the `gene_exp.csv` data as a new object `gene_exp`. Are these data tidy? Use `pivot_wider()` to tidy the data.
```{r}
gene_exp <- read_csv("data/gene_exp.csv")
```
```{r}
gene_exp %>% 
  pivot_wider(
    names_from = "type",
    values_from = "L4_values"
  )
```
## Practice
For the last practice example, I will use data from the [R Ladies Sydney](https://rladiessydney.org/courses/ryouwithme/02-cleanitup-5/) blog. This data is compiled by the NSW Office of Environment and contains the _Enterococci_ counts in water samples obtained from Sydney beaches as part of the Beachwatch Water Quality Program.  
1. Load the beachbugs data and have a look.
```{r}
beachbugs <- read_csv("data/beachbugs_long.csv")
```
2. Use `pivot_wider` to transform the data into wide format.
```{r}
beachbugswider <- beachbugs %>% 
  pivot_wider(
    names_from = "site",
    values_from = "buglevels"
  )
```
3. Now, use `pivot_longer` to transform them back to long!
```{r}
beachbugswider %>% 
  pivot_longer(-year,
    names_to = "site",
    values_to = "buglevels"
  )
```

Hw 9:
1. Start by loading the data `sydneybeaches`. Do some exploratory analysis to get an idea of the data structure.
```{r}
sydneybeaches <- read_csv("data/sydneybeaches.csv")
```
```{r}
glimpse(sydneybeaches)
```
2. Are these data "tidy" per the definitions of the tidyverse? How do you know? Are they in wide or long format?
```{r}
# The tidyverse defintion of the data being tidy is if in the data:
#(1) each variable has its own column  
#(2) each observation has its own row  
#(3) each value has its own cell  

# In this data, all of above conditions are fulfilled, so the data here is tidy. 
# This data is in the long format. *Note: long or wide data can be tidy
```
3. We are only interested in the variables site, date, and enterococci_cfu_100ml. Make a new object focused on these variables only. Name the object `sydneybeaches_long`
```{r}
sydneybeaches_long <- sydneybeaches %>% 
  select(Site, Date, `Enterococci (cfu/100ml)`) 
sydneybeaches_long
```
4. Pivot the data such that the dates are column names and each beach only appears once (wide format). Name the object `sydneybeaches_wide`
```{r}
sydneybeaches_wide <- sydneybeaches_long %>% 
  pivot_wider(
    names_from = "Date",
    values_from = "Enterococci (cfu/100ml)"
  )
sydneybeaches_wide
```
5. Pivot the data back so that the dates are data and not column names.  
```{r}
sydneybeaches_wide <- sydneybeaches_wide %>% 
  pivot_longer(-Site,
    names_to = "data",
    values_to = "enterococci_cfu_100ml",
  )
sydneybeaches_wide
```
6. We haven't dealt much with dates yet, but separate the date into columns day, month, and year. Do this on the `sydneybeaches_long` data.
```{r}
sydneybeaches_long <- sydneybeaches_long %>% 
    separate(Date, into = c("day", "month", "year"), sep = "/")
sydneybeaches_long
```
7. What is the average `enterococci_cfu_100ml` by year for each beach. Think about which data you will use- long or wide.
```{r}
sydneybeaches_avg <- sydneybeaches_long %>% 
  group_by(Site, year) %>% 
  summarize(mean_enterococci_cfu_100ml = mean(`Enterococci (cfu/100ml)`, na.rm = TRUE))
sydneybeaches_avg
```
8. Make the output from question 7 easier to read by pivoting it to wide format.
```{r}
sydneybeaches_avg <- sydneybeaches_avg %>% 
  pivot_wider(
    names_from = "Site",
    values_from = "mean_enterococci_cfu_100ml"
  ) 
sydneybeaches_avg
```
9. What was the most polluted beach in 2013?
```{r}
sydneybeaches_avg %>% 
  pivot_longer(-year, 
               names_to = "site", 
               values_to = "mean_enterococci_cfu_100ml"
               ) %>% 
  filter(year == "2013") %>% 
  arrange(desc(mean_enterococci_cfu_100ml))
```
10. Explore the data! Do one analysis of your choice that includes a minimum of three lines of code.  
```{r}
# Analysis will be for Little Bay Beach (the most polluted beach in 2013), which years had the highest average of Enterococci (cfu/100ml)?
sydneybeaches_avg %>% 
  pivot_longer(-year, 
               names_to = "site", 
               values_to = "mean_enterococci_cfu_100ml"
               ) %>% 
  filter(site == "Little Bay Beach") %>% 
  arrange(desc(mean_enterococci_cfu_100ml))
```

Lab 10_1:
## Review
We have already learned the basics of ggplot, but let's review the plot types we have learned thus far. We will use the mammal life history data to practice. The [data](http://esapubs.org/archive/ecol/E084/093/) are from: *S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.*
```{r message=FALSE, warning=FALSE}
life_history <- read_csv("data/mammal_lifehistories_v2.csv", na="-999") %>% 
  clean_names()
```
## Bar Plots
Recall that `geom_bar` is used when you want to count the number of observations in a categorical variable. `geom_col` is used when you have already counted the number of observations (or you have a defined x and y) and want to plot those counts.
Make two bar plots showing the number of observations for each order of mammals using both geom types.  
`geom_bar`
```{r}
life_history %>% 
  ggplot(mapping = aes(x = order)) +
  coord_flip() + 
  geom_bar(mapping = aes(fill = order)) + 
  labs(title = "Title",
      x = "x",
      y = "y") 
```
`geom_col`
```{r}
life_history %>% 
  count(order, sort = T) 
```
```{r}
life_history %>% 
  count(order, sort = T) %>% 
  ggplot(mapping = aes(x = n, y = order)) +
  geom_col(mapping = aes(fill = order)) + 
  labs(title = "Title",
      x = "x",
      y = "y") 
```
What if we wanted a bar plot of the mean mass for each order?
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass = mean(mass, na.rm = T)) %>% 
  arrange(desc(mean_mass))
```
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass = mean(mass, na.rm = T)) %>% 
  arrange(desc(mean_mass)) %>% 
  ggplot(mapping = aes(x = order, y = mean_mass)) +
  geom_col(mapping = aes(fill = order)) + 
  coord_flip() + 
  labs(title = "Title",
      x = "x",
      y = "y") 
```
There are a few problems here. First, the y-axis is in scientific notation. We can fix this by adjusting the options for the session.
```{r}
options(scipen=999) #cancels scientific notation for the session
```
Next, the y-axis is not on a log scale. We can fix this by adding `scale_y_log10()`.
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass = mean(mass, na.rm = T)) %>% 
  arrange(desc(mean_mass)) %>% 
  ggplot(mapping = aes(x = order, y = mean_mass)) +
  geom_col(mapping = aes(fill = order)) + 
  coord_flip() + 
  scale_y_log10() +
  labs(title = "Title",
      x = "x",
      y = "y") 
```
Lastly, we can adjust the x-axis labels to make them more readable. We do this using `reorder`.
```{r}
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass = mean(mass, na.rm = T)) %>% 
  arrange(desc(mean_mass)) %>% 
  ggplot(mapping = aes(x = reorder(order, mean_mass), y = mean_mass)) +
  geom_col(mapping = aes(fill = order)) + 
  coord_flip() + 
  scale_y_log10() +
  labs(title = "Title",
      x = "x",
      y = "y") 
```
## Scatterplots
Scatter plots allow for comparisons of two continuous variables. Make a scatterplot that compares gestation time and wean mass.
```{r}
life_history %>% 
  ggplot(mapping = aes(x = gestation, y = wean_mass)) +
  geom_point(mapping = aes(color = order)) + 
  geom_smooth(method = lm, se = F, na.rm = T) +
  scale_y_log10() +
  labs(title = "Title",
      x = "x",
      y = "y") 
```
## Boxplots
Box plots are used to visualize a range of values. So, on the x-axis we have a categorical variable and the y-axis is the range. Make a box plot that compares mass across orders.
```{r}
life_history %>% 
  ggplot(mapping = aes(x = order, y = mass)) + #can also do y = log10(mass) instead 
  geom_boxplot(mapping = aes(fill = order)) + 
  scale_y_log10() +
  coord_flip() +
  labs(title = "Title",
      x = "x",
      y = "y") 
```
## Aesthetics: Labels
Now that we have practiced scatter plots, bar plots, and box plots we need to learn how to adjust their appearance to suit our needs. Let's start with labeling x and y axes.  
```{r}
names(life_history)
```
Is there a relationship between mass and litter size; i.e. do larger mammals have more offspring?
```{r}
life_history %>% 
  ggplot(mapping = aes(x = mass, y = litter_size)) +
  geom_point(mapping = aes(color = order)) + 
  scale_y_log10() +
  scale_x_log10() +
  geom_smooth(method = lm, se = F, na.rm = T) +
  #coord_flip() + 
  labs(title = "Title",
      x = "x",
      y = "y") 
```
The plot looks clean, but it is incomplete. A reader unfamiliar with the data might have a difficult time interpreting the labels. To add custom labels, we use the `labs` command.
```{r}
life_history %>% 
  ggplot(mapping = aes(x = mass, y = litter_size)) +
  geom_point(mapping = aes(color = order)) + 
  scale_y_log10() +
  scale_x_log10() +
  geom_smooth(method = lm, se = F, na.rm = T) +
  #coord_flip() + 
  labs(title = "Mass vs. Litter Size",
      x = "Mass (g)",
      y = "Litter Size") 
```
We can adjust the plot further by specifying the size and face of the text. We do this using `theme()`. The `rel()` option changes the relative size of the title to keep things consistent. Adding `hjust` controls the title position.
```{r}
life_history %>% 
  ggplot(mapping = aes(x = mass, y = litter_size)) +
  geom_point(mapping = aes(color = order)) + 
  scale_y_log10() +
  scale_x_log10() +
  geom_smooth(method = lm, se = F, na.rm = T) +
  #coord_flip() + 
  labs(title = "Mass vs. Litter Size",
      x = "Mass (g)",
      y = "Litter Size") +
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```

Lab 10_2:
```{r message=FALSE, warning=FALSE}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv", na = c("", "NA", "\\")) %>% clean_names() #na command only done if you know how NAs are represented in the data 
```
## Aesthetics
There are many options to create nice plots in `ggplot`. One useful trick is to store the plot as a new object and then experiment with geom's and aesthetics. Let's setup a plot that compares `log10_mass` and `log10_hra.` Notice that we are not specifying a geom.
```{r}
p <- homerange %>% 
  ggplot(aes(x = log10_mass, y = log10_hra)) 
```
Play with point size by adjusting the `size` argument.
```{r}
p + geom_point(size = 2.25)
#p + geom_jitter(size = 3)
```
We can color the points by a categorical variable.
```{r}
p + geom_jitter(aes(color = thermoregulation), size = 2)
```
We can also map shapes to another categorical variable.
```{r}
p + geom_point(aes(shape = thermoregulation, color = thermoregulation), size = 2)
```
What about just filling in with a color preference or size?
```{r}
p + geom_point(size = 3, color = "black", alpha = 0.8, shape = 24, fill = "blue") #alpha controls transparency 
```
```{r}
colors()
```
```{r}
# Plot all available point shapes
df <- data.frame(x = 1:25, y = rep(1, 25), shape = 0:24)

ggplot(df, aes(x, y)) +
  geom_point(aes(shape = factor(shape)), size = 5, fill = "lightblue") +
  scale_shape_manual(values = 0:24) +
  theme_void() +
  theme(legend.position = "none") +
  geom_text(aes(label = shape), vjust = -1, size = 5)
```
## Barplots and multiple variables
At this point you should be comfortable building bar plots that show counts of observations using `geom_bar()`. Last time we explored the `fill` option as a way to bring color to the plot; i.e. we filled by the same variable that we were plotting. What happens when we fill by a different categorical variable?  
Let's start by counting how many observations we have in each taxonomic group.
```{r}
homerange %>% 
  count(taxon, sort = T)
```
Now let's make a bar plot of these data.
```{r}
homerange %>% 
  ggplot(mapping = aes(x = taxon)) + 
  geom_bar(aes(fill = taxon)) +
  coord_flip() +
  labs(title = "Observations by Taxon",
       x = "Taxonomic Group", 
       y = NULL) #removes y axis label 
```
By specifying `fill=trophic.guild` we build a stacked bar plot that shows the proportion of a given taxonomic group that is an herbivore or carnivore.
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic_guild)) + 
  geom_bar() +
  coord_flip() +
  labs(title = "Observations by Taxon",
       x = "Taxonomic Group", 
       y = NULL) #removes y axis label 
```
We can also have counts of each trophic guild within taxonomic group shown side-by-side by specifying `position="dodge"`.
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic_guild)) + 
  geom_bar(position = "dodge") +
  coord_flip() +
  labs(title = "Observations by Taxon",
       x = "Taxonomic Group", 
       y = NULL) #removes y axis label 
```
Here is the same plot oriented vertically.
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic_guild)) + 
  geom_bar(position = "dodge") +
  theme(axis.text.x=element_text(angle = 60, hjust = 0.85)) +
  labs(title = "Observations by Taxon",
       x = "Taxonomic Group", 
       y = NULL) #removes y axis label 
```
We can also scale all bars to a percentage.
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic_guild)) + 
  geom_bar(position = position_fill()) +
  scale_y_continuous(labels = scales::percent) + 
  coord_flip()
```
## Practice
For this practice, let's use the [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/articles/intro.html) data.  
1. Make a bar plot that shows counts of individuals by island. Fill by species, and try both a stacked bar plot and `position="dodge"`.
```{r}
penguins
```
```{r}
penguins %>% 
  ggplot(aes(x=island, fill = species)) +
  geom_bar()
```
```{r}
penguins %>% 
  ggplot(aes(x=island, fill = species)) +
  geom_bar(position = "dodge")
```
2. Make another bar plot that shows the number of individuals by sex on each island?
```{r}
penguins %>% 
  ggplot(aes(x = island, fill = sex)) +
  geom_bar(position = "dodge")
```
## Using `group`
In addition to `fill`, `group` is an aesthetic that accomplishes the same function but does not add color. It groups data for plotting.
Here is a box plot that shows `log10.mass` by taxonomic `class.`
```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10_mass)) +
  geom_boxplot()
```
I use `group` to make individual box plots for each taxon.
```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10_mass, group = taxon)) +
  geom_boxplot()
```
I can also use `fill` to associate the different taxa with a color coded key.
```{r}
homerange %>% 
  ggplot(aes(x = class, y = log10_mass, fill = taxon)) +
  geom_boxplot()
```

Hw 10:
## Load the data
Run the following code chunk to import the data. In this chunk, I am removing NA's.
```{r message=FALSE, warning=FALSE}
sharks <- 
  read_csv("data/SharkIncidents_1950_2022_220302.csv", 
           na=c("NA", "N/A", "Unknown", "NOT COUNTED", "", " ")) %>% 
  clean_names()
```
## Questions
1. Are there any "hotspots" for shark incidents in California? Show the top five counties with the most incidents.
```{r}
sharks %>% 
  count(county, sort = T) %>% 
  slice_max(n, n = 5)
```
2.  Make a plot that shows the total number of incidents per county.
```{r}
sharks %>% 
  ggplot(aes(x = county, fill = county)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  labs(title = "Total Number of Incidents Per County",
       x = "County",
       y = "Number of Incidents")
```
3. Are there months of the year when incidents are more likely to occur? Make a plot that shows the total number of incidents by month. Which month has the highest number of incidents?  
```{r}
sharks %>% 
  ggplot(aes(x = as_factor(month), fill = month)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  labs(title = "Total Number of Incidents By Month",
       x = "Month",
       y = "Number of Incidents")
```
```{r}
sharks %>% 
  count(month, sort = T) %>% 
  slice_max(n, n = 5)
```
4. Is there a time of day when most attacks occur? Make a plot that shows the total number of incidents by time of day. Which time of day has the highest number of incidents?  
```{r}
sharks %>% 
  filter(!is.na(time) & !time %in% c("early am", "afternoon")) %>%
  ggplot(aes(x = time, fill = time)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  labs(title = "Total Number of Incidents By Time of Day",
       x = "Time of Day",
       y = "Number of Incidents")
```
```{r}
sharks %>% 
  filter(!is.na(time) & !time %in% c("early am", "afternoon")) %>%
  count(time, sort = T) %>% 
  slice_max(n, n = 5)
```
5. How do the number and types of injuries compare by county? Make a table that shows the number of injury types by county. Which county has the highest number incidents? (hint: your table should be in wide format)
```{r}
sharks_injuries <- sharks %>% 
  group_by(county, injury) %>% 
  summarize(total = n()) %>% 
  pivot_wider(names_from = injury, values_from = total)
sharks_injuries
```
```{r}
sharks %>% 
  count(county, sort = TRUE) %>% 
  slice_max(n, n = 5)
```
6. Use the table from #5 to make a plot that shows the total number of incidents by county.
```{r}
sharks_injuries %>% 
  mutate(total = sum(across(where(is.numeric)), na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(county, total), y = total, fill = county)) +
  geom_col() +
  coord_flip() +
  labs(title = "Total Number of Incidents by County",
       x = "County",
       y = "Number of Incidents") 
```
7. In the data, `mode` refers to a type of activity. Which activity is associated with the highest number of incidents?
```{r}
sharks %>% 
  count(mode, sort = TRUE) %>% 
  slice_max(n, n = 5)
```
8. Make a plot that compares the number of incidents by activity.
```{r}
sharks %>% 
  ggplot(aes(x = mode, fill = mode)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Incidents by Activity",
       x = "Activity",
       y = "Number of Incidents") 
```
9. Which shark species is involved in the highest number of incidents?  
```{r}
sharks %>% 
  filter(!is.na(species)) %>% 
  count(species, sort = TRUE) %>% 
  slice_max(n, n = 5)
```
10. Are all incidents involving Great White's fatal? Make a plot that shows the number and types of incidents for Great White's only.  
```{r}
#Not all incidents involving Great White's are fatal (in fact more are non-fatal)
sharks %>% 
  filter(species == "White") %>% 
  ggplot(aes(x = injury, fill = injury)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Number and Types of Incidents for Great White's",
       x = "Injury Type",
       y = "Number of Incidents")
```

Lab 11_1:
## Data
For these exercises, we are going to use [Gapminder](https://www.gapminder.org/data/) data on population (pop), GDP per capita (gdp), and life expectancy (lex). These data were downloaded and compiled using the `gapminder_compile.Rmd` script in the data folder.
```{r message=FALSE, warning=FALSE}
gap <- read_csv("data/gapminder.csv") %>% clean_names()
```
## Line plots
Line plots are great when you need to show changes over time. Let's see if we can find downturns in the US economy between 2000-2025.  
Let's start by making a clear x and y so we know what we are going to plot.
```{r}
gap %>% 
  filter((year >= 2000 & year <= 2025) & country == "USA") %>% 
  ggplot(aes(x = year, y = gdp)) +
  geom_line() +
  geom_point(shape = 5) + 
  scale_x_continuous(breaks = 2000:2025, minor_breaks = NULL) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  labs(title = "GDP per Capita in the USA, 2000-2025",
      x = "Year",
       y = "GDP")
```
```{r}
gap %>% 
  filter((year >= 2000 & year <= 2025) & country == "USA") %>% 
  ggplot(aes(x = year, y = gdp)) +
  geom_line(color = "red") +
  geom_point(shape = 5) + 
  scale_x_continuous(breaks = seq(2000, 2025, by = 5), minor_breaks = NULL) + #minor_breaks are just extra line dividers that are not labelled (white vertical lines in this case)
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  labs(title = "GDP per Capita in the USA, 2000-2025",
      x = "Year",
       y = "GDP")
```
## Histograms
Histograms show the distribution of continuous variables. As students, you have seen histograms of grade distributions. A histogram `bins` the data and you specify the number of bins that encompass a range of observations. For something like grades, this is easy because the number of bins corresponds to the grades A-F. By default, R uses a formula to calculate the number of bins but some adjustment may be required.  
What was the distribution of life expectancy in 1935?
```{r}
gap %>% 
  filter(year == 1935) %>% 
  ggplot(aes(x = lex)) +
  geom_histogram() 
```
Let's rebuild the histogram, but this time we will specify the color and fill. Do a little experimentation on your own with the different colors.
```{r}
gap %>% 
  filter(year == 1935) %>% 
  ggplot(aes(x = lex)) +
  geom_histogram(color = "black", fill = "turquoise", bins = 30) +
  labs(title = "Distribution of Life Expectancy")
```
## Density plots  
Density plots are similar to histograms but they use a smoothing function to make the distribution more even and clean looking. They do not use bins.
```{r}
gap %>% 
  filter(year == 1935) %>% 
  ggplot(aes(x = lex)) +
  geom_density(color = "black", fill = "turquoise", alpha = 0.7) +
  labs(title = "Distribution of Life Expectancy")
```
I like to see both the histogram and the density curve so I often plot them together. Note that I assign the density plot a different color.
```{r}
gap %>% 
  filter(year == 1935) %>% 
  ggplot(aes(x = lex)) +
  geom_histogram(aes(y = after_stat(density)), color = "black", fill = "turquoise") + 
  geom_density(color = "red")
  labs(title = "Distribution of Life Expectancy")
```
## Practice
1. Antibiotics became widely available in the 1940's. How did life expectancy change between 1930 and 1950? Let's plot the density curves for both years on the same plot. Use the `fill` aesthetic to differentiate the two years.
```{r warning=FALSE}
gap %>% 
  filter(year %in% c(1930, 1950)) %>% #pull out years
  mutate(year = factor(year)) %>% #change to factor
  ggplot(aes(x = lex, color = year, fill = year)) +
  geom_density(alpha = 0.3) + 
  labs(title = "Change in Distribution of Life Expectancy between 1930 and 1950",
       x = "Life Expectancy (years)",
       y = NULL,
       color = "Year",
       fill = "Year")
```
## Create Categories with mutate and case_when()
`case_when()` is a very handy function from `dplyr` which allows us to calculate a new variable from other variables. We use `case_when()` within `mutate()` to do this.`case_when()` allows us to specify multiple conditions.  
The approximate World Bank Income Categories are:  
- Low income: $1,100 or less
- Lower middle income: $1,100 to $4,500
- Upper middle income: $4,500 to $13,900
- High income: $13,900 or more
Let's use `case_when()` to create a new column `income_category` that categorizes countries into these income categories based on their GDP per capita in 2000.
```{r}
income_cat2000 <- gap %>% 
  filter(year == 2000) %>% 
  mutate(income_category = case_when(gdp <= 1100 ~ "low",
                                     gdp > 1100 & gdp <= 4500 ~ "low_middle",
                                     gdp > 4500 & gdp <= 13900 ~ "upper_middle",
                                     gdp > 13900 ~ "high")) %>% 
  select(country, gdp, income_category) %>% 
  mutate(income_category = factor(income_category, 
                                  levels = c("low",
                                             "low_middle",
                                             "upper_middle",
                                             "high")))
income_cat2000
```
## Practice
1. How many countries are in each of the World Bank Income Categories as of the year 2000?
```{r}
income_cat2000 %>% 
    #summarize(n_country = n_distinct(country)) %>% 
    group_by(income_category) %>% 
    summarize(count = n())
```
2. Make a bar plot showing the number of countries in each income category in the year 2000.
```{r}
income_cat2000 %>% 
  drop_na(income_category) %>% #get rid of NA's
  ggplot(aes(x = income_category, fill = income_category)) + 
  geom_bar() + 
  labs(title = "Number of Countries in each Income Category in 2000",
       x = "Income Category",
       y = "Count") +
  theme(legend.position = "none")
```

Lab 11_2:
## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(janitor)
```
```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```
## Data
For this tutorial, we will use:  
[Homerange](http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1). The data are from: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
```{r message=FALSE, warning=FALSE}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv", 
                      na = c("", "NA", "\\")) %>% 
  clean_names()
```
## ggplot themes
There are many options to change the theme of your plots within ggplot. Have a look [here]https://ggplot2.tidyverse.org/reference/ggtheme.html) for a list of the themes.  
Let's start by building a simple barplot.  
```{r}
p <- homerange %>% 
  ggplot(aes(x=taxon, fill=thermoregulation))+
  geom_bar(na.rm=T, position="dodge")+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
p
```
Have a look at the `linedraw` theme; I am adding it as another layer.
```{r}
p + theme_linedraw()+
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```
## Practice
1. Build a scatterplot that compares `log10_mass` and `log10_preymass`. Color by thermoregulation. Store this plot as object `q`.
```{r}
q <- homerange %>% 
  drop_na(log10_mass, log10_preymass) %>% 
  ggplot(aes(x=log10_mass, y = log10_preymass, color = thermoregulation))+
  geom_point() + 
  geom_smooth(method = "lm", se = F, color = "red") + 
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
q
```
2. Use plot `q` but add the classic theme.
```{r}
q + theme_classic()
```
## Legends
There are lots of options to manipulate legends. Have a look [here](http://www.sthda.com/english/wiki/ggplot2-legend-easy-steps-to-change-the-position-and-the-appearance-of-a-graph-legend-in-r-software).
```{r}
p+theme_linedraw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
## Not enough? Try using `ggthemes`
There are many packages that include additional themes, one of which is [ggthemes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/). Some of these are nice because they are designed to mimic the look of popular publications.
```{r}
#install.packages("ggthemes")
library(ggthemes)
```
Here is a list of the `ggthemes`
```{r}
ls("package:ggthemes")[grepl("theme_", ls("package:ggthemes"))]
```
```{r}
p + 
  theme_few()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
## Practice
1. Use plot `q` and try a theme of your choice.
```{r}
q + theme_excel()
```
## RColorBrewer
The default colors used by ggplot are uninspiring. They don't make plots pop out in presentations or publications, and you may want to use a customized palette to make things visually consistent.  
Access the help for `RColorBrewer`.
```{r}
?RColorBrewer
```
The thing to notice is that there are three different color palettes: 1) sequential, 2) diverging, and 3) qualitative. Within each of these there are several selections. You can bring up the colors by using `display.brewer.pal()`. Specify the number of colors that you want and the palette name.
```{r}
display.brewer.pal(9,"Set1") #qualitative palette
```
The [R Color Brewer](http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) website is very helpful for getting an idea of the color palettes. To make things easy, use these two guidelines:
+`scale_colour_brewer()` is for points  
+`scale_fill_brewer()` is for fills  
Here I chose the `Set2` palette. Take a moment and experiment with other options.
```{r}
p+scale_fill_brewer(palette = "Set1")+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
## Practice
1. Keep using the `same `q` plot, but test out one of R Color Brewer palettes.
```{r}
q + scale_color_brewer(palette = "Pastel2")
```
## Paleteer
You can also use `paleteer` to build a custom palette. To access the `paleteer` collection, I add it to a new object.
```{r}
colors <- paletteer::palettes_d_names
```
Now we can display the palettes. Assign the palette to `my_palette` and then build this base R bar plot. There are a lot of options; `paleteer` is a collection of popular palettes. I really like the [`ggsci` package] (https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html)
```{r}
my_palette <- paletteer_d("ggsci::uniform_startrek")
my_palette
```
```{r}
barplot(rep(1,7), axes=FALSE, col=my_palette)
```
Now we just identify `my_palette` as part of `scale_fill_manual()`
```{r}
p+scale_fill_manual(values=my_palette)+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
You can use custom colors by specifying the hex code. You can find hex codes for colors using [this website](https://r-charts.com/colors/).
```{r}
library(scales)

my_cols <- c("#66CDAA",
             "#FFCD00FF")

show_col(my_cols)
```
## Practice
1. Keep using the `same `q` plot, but test out one of `paleteer` themes.
```{r}
my_palette2 <- paletteer_d("ggprism::pastels") #check package:palette in colors.csv
my_palette2
q + scale_color_manual(values=my_palette2)
```
## Faceting
[Faceting](https://ggplot2-book.org/facet.html) is one of the amazing features of ggplot. It allows us to make multi-panel plots for easy comparison.  Here is a boxplot that shows the range of log10_mass by taxon, filled by thermoregulation.
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10_mass, fill=thermoregulation))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```
`facet_wrap()` makes a ribbon of panels by a specified categorical variable and allows you to control how you want them arranged.  
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10_mass, fill=thermoregulation))+
  geom_boxplot()+
  facet_wrap(~trophic_guild, ncol = 2, labeller = label_both) + 
  theme(axis.text.x = element_text(angle = 60, hjust=1), 
        legend.position = "none",
        strip.text = element_text(size = 12, face = "bold"))
```
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10_mass, fill=thermoregulation))+
  geom_boxplot()+
  facet_wrap(~trophic_guild, nrow = 2, labeller = label_both) + 
  theme(axis.text.x = element_text(angle = 60, hjust=1), 
        legend.position = "none",
        strip.text = element_text(size = 12, face = "bold"))
```
`facet_grid()` allows control over the faceted variable; it can be arranged in rows or columns. rows~columns. 
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10_mass, fill=thermoregulation))+
  geom_boxplot()+
  facet_grid(thermoregulation~., labeller = label_both) + 
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10_mass, fill=thermoregulation))+
  geom_boxplot()+
  facet_grid(.~thermoregulation, labeller = label_both) + 
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```
`facet_grid()` will also allow the comparison of two categorical variables, just remember a~b where a is rows and b is columns.  
```{r}
homerange %>% 
  ggplot(aes(x=taxon, y=log10_mass, fill=thermoregulation))+
  geom_boxplot()+
  facet_grid(trophic_guild ~ thermoregulation, labeller = label_both) + 
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```
## Practice
1. Use facet_wrap to compare the distribution of `log10_mass` by `thermoregulation`.
```{r}
homerange %>% 
  ggplot(aes(x=log10_mass, fill=thermoregulation))+
  geom_density(fill = "steelblue", alpha = 0.3) + 
  facet_wrap(~thermoregulation, ncol = 2) + 
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```
2. Use facet_grid to compare the distribution of `log10_hra` by `locomotion` and `thermoregulation`.
```{r}
homerange %>% 
  ggplot(aes(x=log10_hra)) +
  geom_density(fill = "steelblue", alpha = 0.3)+
  facet_grid(thermoregulation ~ locomotion) + 
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```

Hw 11:
Let's focus our attention on data from the past 100 years, between 1925 and 2025.
```{r message=FALSE, warning=FALSE}
gapminder <- read_csv("data/gapminder.csv") %>% 
  clean_names() %>% 
  filter(between(year, 1925, 2025))
```
1. Use the function(s) of your choice to get an idea of the overall structure of the data, including its dimensions, column names, classes, etc. 
```{r}
glimpse(gapminder)
```
```{r}
summary(gapminder)
```
2. How many countries are represented in this dataset?
```{r}
gapminder %>% 
  summarize(n_country = n_distinct(country))
```
3. Which five countries have had the largest absolute population growth between 1925-2025? Show this as a table.
```{r}
gapminder_pop_growth <- gapminder %>% 
  filter(year == 1925 | year == 2025) %>% 
  select(country, year, pop) %>% 
  pivot_wider(names_from = year, 
              values_from = pop, 
              names_prefix = "pop_") %>% 
  mutate(pop_growth = pop_2025 - pop_1925) 

gapminder_pop_growth %>% 
  arrange(desc(pop_growth)) %>% 
  slice_head(n = 5)
```
4. Make a plot that shows population growth over time for the top country you found in question #3.
```{r}
# Top country found in Question #3 was India. 
gapminder %>% 
  filter(country == "India") %>% 
  ggplot(aes(x = year, y = pop)) +
  geom_line(color = "steelblue") +
  labs(title = "Population Growth Over Time (India)",
       x = "Year",
       y = "Population") 
```
5. Has global life expectancy changed between 1925 and 2025? Show the min, median, mean, and max for all countries (combined) in the dataset.
```{r}
gapminder %>% 
  filter(year == 1925) %>% 
  drop_na(lex) %>% 
  summarize(lex_min = min(lex),
         lex_med = median(lex),
         lex_mean = mean(lex),
         lex_max = max(lex))
```
```{r}
gapminder %>% 
  filter(year == 2025) %>% 
  drop_na(lex) %>% 
  summarize(lex_min = min(lex),
         lex_med = median(lex),
         lex_mean = mean(lex),
         lex_max = max(lex))
```
Differences across all of these measures indicates that global life expectancy changed between 1925 and 2025. 
6. Make a plot that shows the distribution of life expectancy for all countries in 1925 and 2025 (try using `geom_density`). What do you notice about the shift in life expectancy over time?
```{r}
gapminder %>% 
  filter(year == 1925 | year == 2025) %>% 
  mutate(year = as.factor(year)) %>% 
  ggplot(aes(x = lex, fill = year)) +
  geom_density(alpha = 0.5) +
  labs(title = "Distribution of Life Expectancy in All Countries in 1925 and 2025",
       x = "Life Expectancy (Years)",
       y = "Density") 
```
The life expectancy greatly shifted over to the left over time, indicating that the life expectancy increased over time (people in 2025 are expected to have longer lifespans than people in 1925). 
7. We are interested in the relationship between per capita GDP and life expectancy; i.e. does having more money help you live longer in 2025? Show this as a plot. (Suggestion! Remove Monaco as an extreme outlier.)
```{r}
gapminder %>% 
  filter(year == 2025, country != "Monaco") %>% 
  ggplot(aes(x = gdp, y = lex)) +
  geom_point(color = "steelblue") +
  geom_smooth(method = "lm", se = F, color = "red") +
  labs(title = "GDP per Capita vs. Life Expectancy (2025)",
       x = "GDP per Capita",
       y = "Life Expectancy") 
```
8. Which five countries have had the highest absolute GDP per capita growth over the past 50 years? Show this as a table. (Suggestion! Remove Monaco as an extreme outlier.)
```{r}
gapminder_gdp_growth <- gapminder %>% 
  filter(year == 1975 | year == 2025, country != "Monaco") %>% #past 50 years = 1975-2025
  select(country, year, gdp) %>% 
  pivot_wider(names_from = year,
              values_from = gdp, 
              names_prefix = "gdp_") %>% 
  mutate(gdp_growth = gdp_2025 - gdp_1975)  

gapminder_gdp_growth %>% 
  arrange(desc(gdp_growth)) %>% 
  slice_head(n = 5)
```
9. How does per capita GDP growth compare between these same five countries over the past 50 years? Show this as a plot.
```{r}
# Top 5 Countries: Singapore, Ireland, Luxembourg, Guyana, Norway  
gapminder %>% 
  filter((country == "Singapore" |
         country == "Ireland" |
         country == "Luxembourg" |
         country == "Guyana" |
         country == "Norway"),
         year >= 1975) %>% 
  ggplot(aes(x = year, y = gdp, color = country)) +
  geom_line() +
  labs(title = "GDP per Capita Growth (1975-2025)",
       x = "Year",
       y = "GDP per Capita") 
```

Lab 13_1:
## Install the package `naniar`
The [naniar](https://naniar.njtierney.com/) package includes some useful tools to manage NA's.
```{r}
#install.packages("naniar")
```
## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("naniar")
library("janitor")
```
## Review
When working with "wild" data, dealing with NA's is a fundamental part of the data cleaning process. Data scientists spend much of their time cleaning and transforming data- including managing NA's. There isn't a single approach that will always work so you need to be careful about using replacement strategies across an entire data set. And, as the data sets become larger NA's can become trickier to deal with.  
For the following, we will use life history data for mammals. The [data](http://esapubs.org/archive/ecol/E084/093/) are from:  
**S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.**  
## Load the mammals life history data and clean the names  
```{r message=FALSE, warning=FALSE}
life_history <- read_csv("data/mammal_lifehistories_v3.csv") %>% clean_names()
```
## Are there any NA's?
Sometimes using one or more of the summary functions can give us clues to how the authors have represented missing data. This doesn't always work, but it is a good place to start.  
```{r}
life_history %>% 
  glimpse()
```
```{r}
life_history %>% 
  summary()
```
## Where are the NA's?
This will give you a quick summary of the number of NA's in each variable. Notice that, at least for now, it doesn't look like there are NA's in most variables because they are still represented by -999.
```{r}
life_history %>% 
  summarize(across(everything(), ~ sum(is.na(.))))
```
Here we look for all of the -999's.
```{r}
life_history %>% 
  summarize(across(everything(), ~ sum(.==-999)))
```
## Dealing with NA's using `read_csv()`
In the case of `-999` as NA, we can use `read_csv` to replace these values when we import the data. But, I only do this once I understand how NA's are represented in the data. Sometimes, they are represented in multiple ways.
```{r message=FALSE, warning=FALSE}
life_history_no_nas <- 
  read_csv("data/mammal_lifehistories_v3.csv", na="-999") %>% 
  clean_names()
```
Rechecking for NA's. Now we see that there are NA's in many of the variables.
```{r}
life_history_no_nas %>% 
  summarize(across(everything(), ~ sum(is.na(.))))
```
## Did we catch them all?
Sometimes it can be helpful to do a quick scan using view() or glimpse() to see if there are any other odd values that might represent NA's. Notice that `not measured` is used in `newborn`.
```{r}
glimpse(life_history_no_nas)
```
```{r}
summary(life_history_no_nas)
```
Notice that `max_life` has no NA's. Does that make sense? How likely is it that we know the lifespan for all of the species in the data set?
```{r}
hist(life_history_no_nas$max_life)
```
```{r}
life_history_no_nas %>% 
  filter(max_life == 0) %>% 
  select(order, family, genus, species, max_life)
```
Let's use `mutate()` and use `na_if()` to replace 0's with NA's in `max_life`. This chunk allows us to address problems in a single variable.
```{r}
life_history_no_nas <- life_history_no_nas %>% 
  mutate(max_life = na_if(max_life, 0))
```
## `naniar`
[naniar](https://naniar.njtierney.com/) is a package that manages NA's. Many of the functions it performs can also be performed using tidyverse functions, but it provides some nice alternatives.  
`miss_var_summary` provides a summary of NA's across the data frame.
```{r}
miss_var_summary(life_history_no_nas)
```
A unique feature of `naniar` is that it can produce visuals to help evaluate NA's. Here we use `gg_miss_var` to visualize the number of NA's in each variable.
```{r}
life_history_no_nas %>% 
  gg_miss_var()
```
We can also use `geom_miss_point()` to visualize where NA's are located in a scatter plot. Here we see that there are NA's in `wean_mass` across the range of `mass`.
```{r}
life_history_no_nas %>% 
  ggplot(aes(x = log10(mass), y = log10(wean_mass))) +
  geom_point() + 
  geom_miss_point() 
```
We can also use `miss_var_summary` with `group_by()`. This helps us better evaluate where NA's are in the data.
```{r}
life_history_no_nas %>% 
  group_by(order) %>% 
  miss_var_summary(order = T)
```
`naniar` has nice replace functions which will allow you to precisely control which values you want replaced with NA's in each variable. This is a nice alternative to `mutate()` and `na_if()`.  
```{r}
life_history %>% #going back to the original data
  replace_with_na(replace = list(newborn = "not measured", 
                                 weaning= -999, 
                                 wean_mass= -999, 
                                 afr= -999, 
                                 max_life= 0, 
                                 litter_size= -999, 
                                 gestation= -999, 
                                 mass= -999)) %>% 
miss_var_summary()
```
You can also use naniar to replace a specific value (like -999) with NA across the entire data set.  
```{r}
life_history %>% 
  replace_with_na_all(condition = ~.x == -999) %>% 
  miss_var_summary()
```
Finally, naniar has some built-in examples of common values or character strings used to represent NA's. The chunk below will use these built-in parameters to replace NA's across the entire data set.
```{r}
common_na_strings
```
```{r}
common_na_numbers
```
```{r}
life_history %>% 
  replace_with_na_all(condition = ~.x %in% c(common_na_strings, common_na_numbers)) %>% 
  mutate(newborn = na_if(newborn, "not measured"))
```
## Practice
Let's practice evaluating NA's in a large data set. The data are compiled from [CITES](https://cites.org/eng). This is the international organization that tracks trade in endangered wildlife. You can find information about the data [here](https://www.kaggle.com/cites/cites-wildlife-trade-database).  
Some key information:  
[country codes](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2)  
1. Import the data and do a little exploration. Be sure to clean the names if necessary.
```{r message=FALSE, warning=FALSE}
cites <- read_csv("data/cites.csv") %>% 
clean_names()
```
```{r}
cites %>% 
  glimpse()
```
2. Use `naniar` to summarize the NA's in each variable.
```{r}
cites %>% 
  miss_var_summary()
```
3. Try using `group_by()` with `naniar`. Look specifically at `class` and `exporter_reported_quantity`. For which taxonomic classes do we have the highest number of missing export data?
```{r}
cites %>% 
  select(class, exporter_reported_quantity) %>% 
  group_by(class) %>% 
  miss_var_summary() %>% 
  arrange(desc(n_miss))
```

Lab 13_2:
## Load the libraries
```{r, message = FALSE}
library("tidyverse")
library("janitor")
library("lubridate") #this will help us manage dates
```
## Joining Files
Sometimes data we need is stored in a separate file or becomes available later and we need to join it to our existing data in order to work with it. Let's use an easy example to explore some of the different types of joins. This section was based on the excellent example provided by [Holly Emblem](https://hollyemblem.medium.com/joining-data-with-dplyr-in-r-874698eb8898).  
Let's assume we have a database of customers and their purchases.
```{r message=FALSE, warning=FALSE}
table_A = read_csv("data/Table_A.csv")
table_B = read_csv("data/Table_B.csv")
```
```{r}
table_A
```
```{r}
table_B
```
The general syntax for joins is...
```{r}
#join_type(firstTable, secondTable, by=columnTojoinOn)
```
`inner_join`  
Where did customers 3 and 4 go? Since they are not on both tables, they were dropped. So, `inner_join` only keeps the rows that have matching values in both tables; it is very strict about matching.
```{r}
inner_join(table_A, table_B, by = "customer_ID")
```
`left_join`
Where did customer 4 go? Notice the missing data for customer 3. So, `left_join` keeps all the rows in the first table and only the matching rows in the second table. If there is no match, it fills in with NA.
```{r}
left_join(table_A, table_B, by = "customer_ID")
```
`right_join`
Where did customer 3 go? Notice the missing data for customer 4. `right_join` does the opposite of `left_join`. It keeps all the rows in the second table and only the matching rows in the first table. If there is no match, it fills in with NA.
```{r}
right_join(table_A, table_B, by = "customer_ID")
```
`full_join`
`full_join` keeps all the rows in both tables and fills in with NA where there is no match.
```{r}
full_join(table_A, table_B, by = "customer_ID")
```
`anti_join`
Provides the rows in the first table for which there are not matching values in the second table.
```{r}
anti_join(table_A, table_B, by = "customer_ID") #used for trouble shooting problematic joins 
```
## Dates
The `lubridate` package was created to deal with dates of all types. There are many useful functions in `lubridate`, most of which are designed to make working with dates easier. For example, you can easily get the current date and time with `today()` and `now()`.
```{r}
today()
```
```{r}
now()
```
There are many different options for working with dates and date-times in lubridate including `ymd`, `mdy`, `dmy`, `mdy_hms`, `ymd_hms`. The formats below are common examples of how you might see dates in your data.
```{r}
dateformat1 <- "20200922"
dateformat2 <- "09-22-2020"
dateformat3 <- "22/09/2020"
dateformat4 <- "09-22-2020 17:00:00"
dateformat5 <- "20200922 170000"
```
## Converting dates 
We can convert each of the examples above using lubridate.
```{r}
ymd(dateformat1)
```
```{r}
mdy(dateformat2)
```
```{r}
dmy(dateformat3)
```
```{r}
mdy_hms(dateformat4)
```
```{r}
ymd_hms(dateformat5)
```
## Practice
For this practice, we will use data from my research on _Usofila_ spiders. These spiders live mostly in caves and the data were collected by building a database of all the known museum specimens in the US. There are two files: 1. spider_records.csv which contains all the records of spiders and 2. spiders_locations.csv which contains the lat/long for each record. An accession number is a unique identifier for each specimen.
1. Load the two files as separate dataframes. Make sure to clean the column names.
```{r message=FALSE, warning=FALSE}
records = read_csv("data/spider_records.csv") %>% clean_names()
locations = read_csv("data/spiders_locations.csv") %>% clean_names()
```
```{r, message = FALSE}
records
```
```{r}
locations
```
2. Join the files using an appropriate key and join type.
```{r}
left_join(records, locations, by = "accession")
```
```{r}
names(records)
```
3. Convert the date column to a date format using lubridate.
```{r}
records %>% 
  mutate(date = dmy(date))
```
4. Save your final data to a `.csv`.
```{r}
write.csv(records, file = "spiders_with_locs.csv", row.names = F)
```

Hw 13:
## Data
For this homework, we will use `amniota` data from: Myhrvold N, Baldridge E, Chan B, Sivam D, Freeman DL, Ernest SKM (2015). “An amniote life-history database to perform comparative analyses with birds, mammals, and reptiles.” _Ecology_, *96*, 3109. doi: 10.1890/15-0846.1 (URL: https://doi.org/10.1890/15-0846.1).
```{r message=FALSE, warning=FALSE}
amniota <- read_csv("data/amniota.csv") %>% 
  clean_names()
```
1. Do some exploratory analysis of the `amniota` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.
```{r}
amniota %>% 
  glimpse() 
```
```{r}
amniota %>% 
  summary() 
```
2. Make any necessary replacements in the data such that all NA's appear as "NA". 
```{r}
amniota <- amniota %>% 
  replace_with_na_all(condition = ~.x == -999) 
```
3. How many total NA's are in the data set? Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amniota` data.
```{r}
amniota %>% 
  miss_var_summary()
```
4. Double check your replacement using `summary`. Do you see any other potential issues? If so, fix them.
```{r}
amniota %>% 
  summary()
```
```{r}
amniota <- amniota %>% 
  replace_with_na_all(condition = ~.x == 0) %>% 
  replace_with_na(replace = list(
    female_maturity_d = -30258.71100
  ))
```
5. Use `naniar` to visualize the number of NA's in each variable. Which variables have the most NA's?
```{r}
amniota %>% 
  gg_miss_var()
```
The variables with the most NA's are subspecies (21322) and female_body_mass_at_maturity_g (21318). 
6. Calculate the number of NAs in the `egg_mass_g` column sorted by taxonomic class; i.e. how many NA's are present in the `egg_mass_g` column in birds, mammals, and reptiles? Does this results make sense biologically? How do these results affect your interpretation of NA's?
```{r}
amniota %>%
  select(class, egg_mass_g) %>% 
  group_by(class) %>%
  miss_var_summary() %>% 
  arrange(desc(pct_miss))
```
These results make sense as the Mammalia class (mammmals) do have 100% NA's. Mammals are species that do not lay eggs, so it makes sense that there is no egg mass measured for them. These results showcase to me that NA's don't always mean that entries are sometimes forgotten, but sometimes that they can be used to say that a certain parameter don't apply to certain entries. 

Lab 14_1:
## Vector vs. Raster Data
In lab we are using vector data, which represents geographic features as points, lines, and polygons. Raster data represents space as a regular grid of cells (pixels), where each cell stores a value for a variable (elevation, precipitation, temperature, reflectance, etc.).  
- Vector data is often used for discrete features (e.g., locations, roads, rivers, political boundaries).
- Raster data is often used for continuous variables (e.g., elevation, temperature, land cover).  
## Packages
There are many packages and techniques for working with spatial data in R. We will cover only some of the basics. One package is ggmap, which allows us to retrieve and display base maps from services such as Google Maps and OpenStreetMap (and related tile providers). It integrates with ggplot2, meaning maps and spatial data can be layered using the same grammar of graphics framework.  
We will also use leaflet, which is a package for creating interactive maps. It allows you to create maps that users can zoom and pan, and it supports a wide range of map tiles and markers.  
## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
```
## Install and load the `ggmap` and `leaflet` packages  
```{r message=FALSE, warning=FALSE}
#install.packages("ggmap")
library(ggmap)
#install.packages("leaflet")
library(leaflet)
```
## Register for a Stamen API Key
Because some of these maps are proprietary, you will need an API key to use them. We are going to use Stamen maps which requires registration, but no payment.   
Start by making an account with [Stadia Maps](https://client.stadiamaps.com/signup/). You need to add a "property" which is just a name for your project. Once you have an account, you can get an API key. Use the following code to register your key.  
```{r}
#register_stadiamaps("YOUR API KEY HERE", write = FALSE)
register_stadiamaps("efbbad0c-b639-4a5e-bc42-a820b574fc9b", write = FALSE)
```
## Load the Data
Let's load our processed spider data from the previous lab. There is an error in one of the coordinates that we will also fix here.
```{r message=FALSE, warning=FALSE}
spiders <- read_csv("data/spiders_with_locs.csv")%>% 
  clean_names() %>% 
  filter(latitude<=42)
```
## Create Base Map
Our goal is to plot the spiders locations from the columns which contain the latitude and longitude. First, we need to get a base map for plotting our points on. We could plot them without a base map, but that wouldn't give us any context as to where they are in space. To get a base map we specify a min and max of each x and y coordinate, and create a bounding box.  
We set the bounding box to a little outside our min and max locations with `f = 0.03`. This gives an extra 3% padding to the bounding box.
`summary()` gives us our min and max.
```{r}
spiders %>% 
  select(latitude, longitude) %>% 
  summary()
```
Now we set the bounding box. We use the min and max values for latitude and longitude to set the range.  
```{r}
lat <- c(34.67, 41.80)
long <- c(-124.1, -115.5)
bbox <- make_bbox(long, lat, f = 0.03)
```
Let's get a base map for our bounding box area. There are several different map [styles](https://docs.stadiamaps.com/themes/)
```{r}
map1 <- get_stadiamap(bbox, maptype = "stamen_terrain", zoom = 7)
```
Have a look at the map.  
```{r}
ggmap(map1)
```
## Adding Points to Base Map
To add our points we only need to specify the x and y location similar to how we made plots in `ggplot.` 
```{r}
ggmap(map1) + 
  geom_point(data = spiders, aes(longitude, latitude), size = 0.75, color = "black", alpha = 0.8) +
  labs(x = "Longitude",
       y = "Latitude",
       title = "California Usofila Records")
```
## Leaflet
If we want to make an interactive map, we can use the `leaflet` package. This allows us to zoom and pan around the map, and we can add popups with information about each point.  
Get the min/max from the data.
```{r}
spiders %>% 
  summarize(
  lon_min = min(longitude, na.rm = T),
  lon_max = max(longitude, na.rm = T),
  lat_min = min(latitude, na.rm = T),
  lat_max = max(latitude, na.rm = T))
```
Add this to four new variables which will be used to create the bounding box in leaflet.
```{r}
lon_min <- -124.13
lon_max <- -115.49
lat_min <- 34.67
lat_max <- 41.8 
```
Build the interactive map using Stadia maps.
```{r}
leaflet(spiders) %>% 
  addProviderTiles(providers$Stadia.StamenTerrain) %>% 
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude, 
    radius = 2,
    stroke = FALSE,
    fillOpacity = 0.7
  ) %>% 
  addScaleBar(position = "bottomleft") %>% 
  fitBounds(lng1 = lon_min,
            lat1 = lat_min,
            lng2 = lon_max,
            lat2 = lat_max)
```
Build the same interactive map using default OpenStreet maps.
```{r}
leaflet(spiders) %>% 
  addTiles() %>% 
  addProviderTiles(providers$Stadia.StamenTerrain) %>% 
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude, 
    radius = 2,
    stroke = FALSE,
    fillOpacity = 0.7
  ) %>% 
  addScaleBar(position = "bottomleft") %>% 
  fitBounds(lng1 = lon_min,
            lat1 = lat_min,
            lng2 = lon_max,
            lat2 = lat_max)
```
## Practice
For this practice we will use data on the distribution of grizzly bears in Alaska. The data are from: https://rcweb.dartmouth.edu/~f002d69/workshops/index_rspatial.html) represent sightings of grizzly bears (Ursos arctos) in Alaska.
1. Load the `grizzly` data and evaluate its structure.  
```{r message=FALSE, warning=FALSE}
grizzly <- read_csv("data/bear-sightings.csv")
```
```{r}
grizzly %>% 
  summary()
```
2. Use the range of the latitude and longitude to build an appropriate bounding box for your map.  
```{r}
lon_min <- -166.2
lon_max <- -131.3
lat_min <- 55.02
lat_max <- 70.37
```
3. Load a map from `stadiamaps` in a `stamen_terrain` projection and display the map.
```{r}
lat <- c(55.02, 70.37)
lon <- c(-166.2, -131.3)
  bbox <- make_bbox(lon, lat, f = 0.03)
```
```{r}
map1 <- get_stadiamap(bbox, maptype = "stamen_terrain", zoom = 8)
```
4. Build a final map that overlays the recorded observations of grizzly bears in Alaska.  
```{r}
ggmap(map1) +
  geom_point(data = grizzly, aes(longitude, latitude), size = 0.8) +
  labs(x = "Longitude",
       y = "Latitude",
       title = "Alaska Grizzly Bear Records")
```

Lab 14_2:
## Shiny
Shiny is an R package which makes it easy to design interactive web applications without knowing any html, css, or java script. First we need to make sure we have shiny installed and loaded.
```{r, echo=FALSE}
#install.packages("shiny")
library("shiny")
```
We will also need the `tidyverse`, `janitor`, and `palmerpenguins` to run the examples.
```{r echo=FALSE, message=FALSE, warning=FALSE}
library("tidyverse")
library("palmerpenguins")
library("janitor")
```
## Shiny Example
Shiny apps can be designed to do almost anything you can imagine. They are very useful for making specialized tools for specific needs in business and research, or just for fun! Let's run a basic example on our computer. Type the following into the R console. You should have the file `plot_iris.R` in your working directory. Use the drop down menus to change the x and y axes on the plot.
```{r, eval=F, echo=T}
#runApp("plot_iris.R")
```
## Built-in Examples
There are many built-in examples in shiny. You can run them by typing the following into the R console. Examples: "01_hello", "02_text", "03_reactivity", "04_mpg", "05_sliders", "06_tabsets", "07_widgets", "08_html", "09_upload", "10_download", "11_timer"
```{r}
#runExample("04_mpg")
```
## Other Examples
Shiny apps can be very powerful and complex. Here are some examples of shiny apps that have been built by the [life science community](https://www.appsilon.com/post/r-shiny-in-life-sciences-examples#dashboard-2).  
## Shiny App Skeleton
A shiny app consists of a user interface (ui) function, server function, and run function.
1.  The `ui` function controls the user inputs and the way the app will be displayed; i.e. this controls how your app looks.  
2.  The `server` function is the part of the app which takes the values of the user inputs, performs calculations and/or makes plots, and prepares the outputs for display; i.e. this controls how your app works.
3.  The `run` function combines the ui and server function to run the app.  
Below is a bare bones shiny app. You can get this code snippet by typing `shinyapp` + shift + tab. It doesn't do anything because we haven't created any inputs with the ui or any outputs with the server.  
```{r, eval=F, echo=T}
library(shiny)
ui <- fluidPage(
)
server <- function(input, output, session) { }
shinyApp(ui, server)
```
For the rest of the lab we will work on building an example app using `palmerpenguins`.
## Shiny Inputs
There are many types of inputs in shiny, see the [shiny cheat sheet](https://shiny.posit.co/r/articles/start/cheatsheet/). Our example app uses `selectInput()`, which takes a set of predefined choices and supplies them to the user in a drop down menu format. We are working with the `palmerpenguins` data set, so we want the selections to be the column names of the data frame.  
Let's get the names of the `palmerpenguins` data frame and add them as a `selectInput()` to our ui. `selectInput()` takes several arguments including the label that the server function will use, the label that will display to the user, the selection choices, and the initially selected choice. I always start by making a simple example plot so I know where I am going. 
```{r}
penguins %>% 
  ggplot(aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point(na.rm = T)
```
Since we will make a scatterplot, let's focus on the continuous variables. The first step is to setup the x variable input.
```{r, eval=F, echo=T}
library(shiny)
ui <- fluidPage(
  selectInput("x", 
              "Select X Variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm")
)
server <- function(input, output, session) { }
shinyApp(ui, server)
```
Let's repeat the process for the y variable input.
```{r, eval=F, echo=T}
library(shiny)
ui <- fluidPage(
  selectInput("x", 
              "Select X Variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm"),
  selectInput("y", 
              "Select Y Variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm"),
)
server <- function(input, output, session) { }
shinyApp(ui, server)
```
## Server Side
Great, we now have our user inputs set up! Next we need to work with those inputs so R can make the plot. This can be the tricky part. The processing is done in the server function. The ui passes the names of our choices via an object called `inputs` to the server function. We need to access our inputs by their names with the `$` symbol. To get the x input we type `input$x` and y is `input$y`.  
All reactive inputs need to be wrapped in a reactive environment called a reactive function. Reactivity is the core of shiny. Let's wrap our reactive input values in a reactive function on the server side and run our app as a test. What happens? Are we doing anything with the inputs yet?  
```{r, eval=F, echo=T}
library(shiny)
ui <- fluidPage(
  selectInput("x", 
              "Select X Variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm"),
  selectInput("y", 
              "Select Y Variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm"),
)
server <- function(input, output, session) {
  reactive({input$x
            input$y
            })
}
shinyApp(ui, server)
```
## Shiny Outputs
So we have our inputs in a reactive environment, but we want to actually use those inputs to make a plot and display it on the ui. To make and display the plot, we need to save it to a named output object that the ui can use. To do this we use the reactive expression `renderPlot()` and access the plot on the ui side with `plotOutput()`. The inputs from `selectInput()` are character strings, so we need to use `aes_string()` in `ggplot`.
```{r, eval=F, echo=T}
library(shiny)
ui <- fluidPage(
  selectInput("x", 
              "Select X Variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm"),
  selectInput("y", 
              "Select Y Variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm"),
  plotOutput("plot",
             width = "500px",
             height = "400px")
)
server <- function(input, output, session) {
  output$plot <- renderPlot({
    ggplot(data = penguins,
           aes_string(x = input$x,
                      y = input$y,
                      color = "species")) +
      geom_point(na.rm = T)
            })
}
shinyApp(ui, server)
```
```{r}
library(shiny)
ui <- fluidPage(
  selectInput("x", 
              "Select X Variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm"),
  selectInput("y", 
              "Select Y Variable",
              choices = c("bill_length_mm",
                          "bill_depth_mm",
                          "flipper_length_mm",
                          "body_mass_g"),
              selected = "bill_length_mm"),
  checkboxInput("show_lm",
                "Show best-fit line(lm)",
                value = FALSE),
  plotOutput("plot",
             width = "500px",
             height = "400px")
)
server <- function(input, output, session) {
  output$plot <- renderPlot({
    ggplot(penguins,
           aes(x = .data[[input$x]],
               y = .data[[input$y]])) +
      geom_point(na.rm = T) +
      if (input$show_lm)
        geom_smooth(method = "lm", se = F, na.rm = T)
      else 
        NULL
            })
}
shinyApp(ui, server)
```
## Practice
1. Please load the data as a new object `homerange` and have a look at the column names.
```{r echo=FALSE, message=FALSE, warning=FALSE}
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")
```
```{r}
homerange %>% 
  ggplot(aes(x = locomotion,
             fill = trophic.guild)) +
  geom_bar(position = "dodge")
```
```{r}
names(homerange)
```
2.  Build a shiny app that produces a barplot that explores `locomotion` type with a fill reaction for the variables `trophic_guild` and `thermoregulation`.
```{r, eval=F, echo=T}
library(shiny)
ui <- fluidPage(
  selectInput("x", 
              "Select Fill Variable",
              choices = c("trophic.guild", #clean_names for trophic_guild 
                          "thermoregulation"),
              selected = "trophic.guild"),
  plotOutput("plot",
             width = "500px",
             height = "400px")
)
server <- function(input, output, session) {
  output$plot <- renderPlot({
    ggplot(data = homerange,
           aes_string(x = "locomotion",
                      fill = input$x)) +
      geom_bar(position = "dodge", alpha = 0.8, color = "black") +
      labs(x = NULL, y = NULL, fill = "Fill Variable")
            })
}
shinyApp(ui, server)
```

Hw 14:
## Load the libraries  
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(ggmap)
library(leaflet)
``` 
## API Key
```{r}
register_stadiamaps("efbbad0c-b639-4a5e-bc42-a820b574fc9b", write = FALSE)
```
## Load the Data
For this homework, we will use the shark attack data to visualize where the attacks occurred. The [data](https://catalog.data.gov/dataset/shark-incident-database-california-56167) are from: State of California- Shark Incident Database.  
```{r message=FALSE, warning=FALSE}
sharks <- read_csv("data/SharkIncidents_1950_2022_220302.csv") %>% 
  clean_names() %>% 
  filter(longitude !="NA" & latitude !="NA") %>% # pulling out NA locations
  mutate(longitude = as.numeric(longitude)) # converting longitude to numeric
```
1. Use the range of the latitude and longitude to build an appropriate bounding box for your map.
```{r}
sharks %>% 
  summary()
```
```{r}
lon_min <- -124.7
lon_max <- -117.1
lat_min <- 32.59
lat_max <- 41.56
```
```{r}
lat <- c(lat_min, lat_max)
lon <- c(lon_min, lon_max)
  bbox <- make_bbox(lon, lat, f = 0.03)
```
2. Load a map from `stamen` in a terrain style projection and display the map.  
```{r, echo=T}
map_sharks <- get_stadiamap(bbox, maptype = "stamen_terrain", zoom = 7)
ggmap(map_sharks)
```
3. Build a map that overlays the recorded observations of shark attacks in California.  
```{r}
ggmap(map_sharks) +
  geom_point(data = sharks, aes(longitude, latitude),  color = "red", alpha = 0.8, size = 0.8) +
  labs(x = "Longitude",
       y = "Latitude",
       title = "California Shark Incidents (1950-2022)")
```
4. Build a map that only shows the fatalities.  
```{r}
fatalities <- sharks %>% 
  filter(injury == "fatal")

ggmap(map_sharks) + 
  geom_point(data = fatalities, aes(longitude, latitude), size = 2, color = "red") +
  labs(title = "California Fatal Shark Incidents (1950-2022)")
```
5. Build the same map (fatalities only), but this time make it interactive using `leaflet`. Also, color the points red.  
```{r}
leaflet(fatalities) %>% 
  addTiles() %>% 
  addProviderTiles(providers$Stadia.StamenTerrain) %>% 
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude, 
    radius = 2,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7
  ) %>% 
  addScaleBar(position = "bottomleft") %>% 
  fitBounds(lng1 = lon_min,
            lat1 = lat_min,
            lng2 = lon_max,
            lat2 = lat_max)
```
6. Which county has the highest number of recorded shark attacks? Make a map that only shows the attacks in this county. (hint: it may help improve the map if you remove duplicate locations, i.e. multiple attacks at the same location.)
```{r}
sharks %>% 
  count(county) %>% 
  arrange(desc(n))
```
```{r}
top_county_data <- sharks %>% 
  filter(county == "San Diego") 
```
```{r}
top_county_data %>% 
  summary()
```
```{r}
lon_min <- -118.1
lon_max <- -117.1
lat_min <- 32.59
lat_max <- 33.37
```
```{r}
lat <- c(lat_min, lat_max)
lon <- c(lon_min, lon_max)
  bbox <- make_bbox(lon, lat, f = 0.03)
```
```{r}
map_sharks <- get_stadiamap(bbox, maptype = "stamen_terrain", zoom = 7)
```
```{r}
ggmap(map_sharks) +
  geom_point(data = top_county_data, aes(longitude, latitude), color = "red") +
  labs(title = "Shark Incidents in San Diego County")
```

Lab 15_1:
## Load the Libraries  
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(shiny)
library(shinythemes)
library(janitor)
```
## Build a Basic Shiny App
Recall that every shiny app has a ui, server, and run function. The ui is the user interface, which is what the user sees and interacts with. The server is where the code that creates the outputs goes. The run function is where we call the ui and server to run the app. 
The basic skeleton of a shiny app looks like this:
```{r, eval=F, echo=T}
ui <- fluidPage(
)
server <- function(input, output, session) {
}
shinyApp(ui, server)
```
Let's use the built-in msleep data to build a basic app. The data are from: V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007.
```{r}
glimpse(msleep)
```
Since the data are focused on sleep patterns, let's make a plot that shows the range of `sleep_total` by `vore` type.
```{r}
msleep %>% 
  filter(vore != "NA") %>% 
  ggplot(aes(x = vore, y = sleep_total, fill = vore)) +
  geom_boxplot(alpha = 0.75) +
  labs(title = "Sleep Total By Vore Type",
       x = "Vore",
       y = "Sleep Total (Hours)")
```
```{r}
names(msleep)
```
If we want to make the plot interactive, we could allow users to select the sleep variable of interest. These include `sleep_total`, `sleep_rem`, `sleep_cycle`, and `awake.` We can use a selectInput to allow users to select the variable of interest in a drop-down menu.
```{r, eval=F, echo=T}
ui <- fluidPage(
  theme = shinythemes::shinytheme("cosmo"), 
  selectInput("y", 
              "Select Sleep Variable",
              choices = c("sleep_total", 
                          "sleep_rem",
                          "sleep_cycle",
                          "awake"),
              selected = "sleep_total"),
  plotOutput("plot",
             width = "600px",
             height = "500px")
)
server <- function(input, output, session) {
  output$plot <- renderPlot({
    msleep %>% 
      filter(vore != "NA") %>% 
      ggplot(aes(x = vore, 
                 y = .data[[input$y]],
                 fill = vore)) +
        geom_boxplot(alpha = 0.8, color = "black") +
        labs(title = "Sleep Variables by Vore Type", 
             x = "Vore", 
             fill = "Vore Type") +
             theme_minimal()
              })
}
shinyApp(ui, server)
```
Since we only have one interaction, we can move on to the server function. The server function is where we write the code that creates the outputs. In this case, we want to create a plot that shows the range of the selected sleep variable by vore type. We can use the `renderPlot` function to create the plot output.
```{r, eval=F, echo=T}

```
## Running A Shiny App from an R Script
In BIS 15L, we use RMarkdown to write code, including building shiny apps. But, in practice shiny apps are often built as stand-alone scripts. This way, even people with limited programming experience can run an app or it can be hosted online.  
Let's take our app above and make it into a standalone R Script.  
1. Navigate to the file tab in RStudio and click "New File" > "R Script".  
2. Copy/paste your app into the script.  
3. Make sure to add the libraries that are needed at the top.  
4. Save your app as `msleep.R`.  

Lab 15_2:
## Load the Data
For this lab, we will use data on the gestation length of eutherian mammals. The data are from: Thodoris Danis, Antonis Rokas; The evolution of gestation length in eutherian mammals. Proc. Biol. Sci. 1 October 2024; 291 (2033): 20241412. https://doi.org/10.1098/rspb.2024.1412
```{r message=FALSE, warning=FALSE}
gestation <- read_csv("data/Data_S1.csv") %>% 
  clean_names() %>% 
  rename(gestationlength=gestationength)
```
```{r}
names(gestation)
```
```{r}
gestation %>% 
  distinct(order)
```
## Aesthetics
There are many ways to adjust shiny apps for better visual appeal. Let's start with some more practice and then add details.  
1. Build an app that allows users to select a mammalian `order` and a variable of interest (body mass, longevity, or gestation length) and then plots the density distribution of that variable for the selected order.
```{r, eval=F, echo=T}
ui <- fluidPage(
  theme = shinythemes::shinytheme("cyborg"), 
  titlePanel("Distribution of Gestation Length by Order"),
  selectInput("x", 
              "Select Mammalian Order",
              choices = sort(unique(gestation$order)),
              selected = "Afrosoricida"),
  selectInput("y", 
              "Select Variable of Interest",
              choices = c("bodymass",
                          "longevity",
                          "gestationlength"),
              selected = "bodymass"),
  helpText("Testing 123"),
  plotOutput("plot",
             width = "600px",
             height = "500px")
)
server <- function(input, output, session) {
  output$plot <- renderPlot({
    gestation %>% 
      filter(input$x == .data[["order"]]) %>% 
      ggplot(aes(.data[[input$y]])) +
      geom_density(alpha = 0.8, fill = "steelblue", color = "black") +
      labs(title = "Density Distribution", 
           x = "Order", 
           fill = "Variable of Interest") +
      theme_minimal()
  })
}
shinyApp(ui, server)
```
## ui Layout with `shinydashboard`
We have a functional app, but the layout doesn't look very nice. We can change that with many different methods (and specialized libraries), but shiny dashboards are simple to use and offer a lot of functionality for making apps look professional quickly.  
We won't use `fluidPage()` with `shinydashboard` but rather `dashboardPage()`. A shiny dashboard contains a minimum of a header, a sidebar, and a body.
```{r, eval=F, echo=T}
ui <- 
  dashboardPage(
  dashboardHeader(title = "This is the Header"),
  dashboardSidebar(title = "This is the Sidebar"),
  dashboardBody() # this is where the output will go
)
server <- function(input, output) { }
shinyApp(ui, server)
```
Let's add our ui and server elements to make a functional dashboard. We will put our ui elements in the `dashboardBody()`. For now, we will turn off the sidebar.
```{r, eval=F, echo=T}
ui <-  dashboardPage(
  dashboardHeader(title = "Distribution of Gestation Length by Order"),
  dashboardSidebar(disable = T),
  dashboardBody(
  selectInput("x", 
              "Select Mammalian Order",
              choices = sort(unique(gestation$order)),
              selected = "Afrosoricida"),    
  selectInput("y", 
              "Select Variable of Interest",
              choices = c("bodymass",
                          "longevity",
                          "gestationlength"),
              selected = "bodymass"),    
  helpText("Testing 123"),
  plotOutput("plot",
             width = "600px",
             height = "500px")
  ) 
)
server <- function(input, output, session) { 
  output$plot <- renderPlot({
    gestation %>% 
      filter(input$x == .data[["order"]]) %>% 
      ggplot(aes(.data[[input$y]])) +
      geom_density(alpha = 0.8, fill = "steelblue", color = "black") +
      labs(title = "Density Distribution", 
           x = "Order", 
           fill = "Variable of Interest") +
      theme_minimal()
  })
} 
shinyApp(ui, server)
```
Now let's add our server functions.
```{r, eval=F, echo=T}
ui <-  dashboardPage(
  dashboardHeader(title = "Distribution of Gestation Length by Order"),
  dashboardSidebar(disable = T),
  dashboardBody(
  selectInput("x", 
              "Select Mammalian Order",
              choices = sort(unique(gestation$order)),
              selected = "Afrosoricida"),    
  selectInput("y", 
              "Select Variable of Interest",
              choices = c("bodymass",
                          "longevity",
                          "gestationlength"),
              selected = "bodymass"),    
  helpText("Testing 123"),
  plotOutput("plot",
             width = "600px",
             height = "500px")
  ) 
)
server <- function(input, output, session) { }
shinyApp(ui, server)
```
What if you want to turn on the sidebar and put the inputs there? We can do that by moving the `selectInput()` functions to the `dashboardSidebar()`.
```{r, eval=F, echo=T}
ui <-  dashboardPage(
  dashboardHeader(title = "Distribution of Gestation Length by Order"),
  dashboardSidebar(
   selectInput("x", 
              "Select Mammalian Order",
              choices = sort(unique(gestation$order)),
              selected = "Afrosoricida"),    
  selectInput("y", 
              "Select Variable of Interest",
              choices = c("bodymass",
                          "longevity",
                          "gestationlength"),
              selected = "bodymass")  
  ),
  dashboardBody(
  helpText("Testing 123"),
  plotOutput("plot",
             width = "600px",
             height = "500px")
  ) 
)
server <- function(input, output, session) {
  output$plot <- renderPlot({
    gestation %>% 
      filter(input$x == .data[["order"]]) %>% 
      ggplot(aes(.data[[input$y]])) +
      geom_density(alpha = 0.8, fill = "steelblue", color = "black") +
      labs(title = "Density Distribution", 
           x = "Order", 
           fill = "Variable of Interest") +
      theme_minimal()
  })
}
shinyApp(ui, server)
```
Whew! We made it! We created a nice looking app with `shinydashboard`. And now we can easily see how even a basic shiny app can become messy and why it's always helpful to include lots of annotations. chatGPT can really help troubleshoot errors and is especially helpful with annotation.    
## Practice
Let's try to build an app that allows users to explore sex composition of wolf packs by `pop`. These data are from: Brandell, Ellen E (2021), Serological dataset and R code for: Patterns and processes of pathogen exposure in gray wolves across North America, Dryad, [Dataset](https://doi.org/10.5061/dryad.5hqbzkh51)  
1. Start by loading and exploring the data.
```{r message=FALSE, warning=FALSE}
wolves <- read_csv("data/wolves_data/wolves_dataset.csv") %>% clean_names()
```
```{r}
names(wolves)
```
2. For the app, here is a faceted version of what we are looking for. We want the reactive part to be `pop` and you should try to use `shinydashboard`.
```{r}
wolves %>% 
  filter(sex!="NA") %>% 
  ggplot(aes(x=sex, fill=sex))+
  geom_bar()+
  facet_wrap(~pop)
```
```{r, eval=F, echo=T}
ui <-  dashboardPage(
  dashboardHeader(title = "Sex Composition by Wolf Population"),
  dashboardSidebar(disable = T),
  dashboardBody(
  selectInput("x", 
              "Select Population",
              choices = sort(unique(wolves$pop)),
              selected = "AK.PEN"),
  helpText("Testing 123"),
  plotOutput("plot",
             width = "600px",
             height = "500px")
  ) 
)
server <- function(input, output, session) {
  output$plot <- renderPlot({
    wolves %>% 
      filter(pop == input$x) %>% 
      ggplot(aes(x = sex, fill = sex)) +
      geom_bar(alpha = 0.8, color = "black") +
      labs(x = "Sex", y = NULL) +
      theme_minimal()
  })
}
shinyApp(ui, server)
```

Hw 15:
## Load the data
For this homework, we will use data on African elephants. These data are from: Phyllis C. Lee, Luc F. Bussière, C. Elizabeth Webber, Joyce H. Poole, Cynthia J. Moss; Enduring consequences of early experiences: 40 year effects on survival and success among African elephants (Loxodonta africana). Biol Lett 23 April 2013; 9 (2): 20130011. https://doi.org/10.1098/rsbl.2013.0011
```{r}
elephants <- read_csv("data/elephants_data/elephants.csv") %>%
  clean_names()
```

1. There are only three variables in the data. Two are continuous (age, height) and one is categorical (sex). Build an app that compares the range of age and height by sex. Users should be able to select either age or height. Make the app neat, clean, and professional. Use `shinydashboard` to improve the aesthetics of your app.
```{r}
elephants %>% 
  names()
```


```{r, eval=F, echo=T}
ui <- dashboardPage(
  
  dashboardHeader(title = "Elephant Age/Height vs. Sex"), 
  
  dashboardSidebar(
  selectInput("y", 
              "Select Variable of Interest",
              choices = c("age",
                          "height"),
              selected = "age")
  ),
  
  
  dashboardBody(
  helpText("Data source: Lee et al. (2013) Enduring consequences of early experiences: 40 year effects on survival and success among African elephants (Loxodonta africana)"),
  
  plotOutput("plot",
             width = "600px",
             height = "500px")
)
)

server <- function(input, output, session) {
  
  output$plot <- renderPlot({
    
    elephants %>% 
      ggplot(aes(x = sex, y = .data[[input$y]], fill = sex)) +
      geom_boxplot(alpha = 0.8) +
      labs(title = "Distribution of Age/Height by Sex", 
           x = "sex", 
           y = input$y) +
      theme_minimal()
  })
  
}

shinyApp(ui, server)
```

2. In addition to the .Rmd file, save your app as a standalone R script called `elephants.csv`.
(readcsv function)
```{r}
# Below is what the standalone R script looks like in elephants.R:
#
#library(tidyverse)
#library(shiny)
#library(shinythemes)
#library(janitor)
#ui <- fluidPage(
#  theme = shinythemes::shinytheme("cyborg"), 
# titlePanel("Elephant Age/Height vs. Sex"), 
#    selectInput("y", 
#                "Select Variable of Interest",
#                choices = c("age",
#                            "height"),
#                selected = "age"),
#    helpText("Data source: Lee et al. (2013) Enduring consequences of early experiences: 40 year effects on survival and success among African elephants (Loxodonta africana)"),
#    plotOutput("plot",
#               width = "600px",
#               height = "500px")
#)
#
#server <- function(input, output, session) {
#  
#  output$plot <- renderPlot({
#    elephants %>% 
#      ggplot(aes(x = sex, y = .data[[input$y]], fill = sex)) +
#      geom_boxplot(alpha = 0.8) +
#      labs(title = "Distribution of Age/Height by Sex", 
#           x = "sex", 
#           y = input$y) +
#      theme_minimal()
#  })
#}
#
#shinyApp(ui, server)
```



