---
title: "Lab 7"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## 

Lab 1_1: 
## RStudio Layout  
When you open RStudio, you will see that the window is divided into four sections. Each of these sections has a specific function that we will use throughout the course.  

1. **Source (Top Left)**: This is where you write and edit R scripts and RMarkdown files.  
2. **Console (Bottom Left)**: This is where R executes your commands.   
3. **Environment (Top Right)**: The Environment tab shows all the objects (like data frames, vectors, etc.) that you have created during your R session.  
4. **Files/Plots/Packages/Help/Viewer (Bottom Right)**: This section has multiple tabs. The Files tab allows you to navigate your file system. The Plots tab displays any graphs you create. The Packages tab lets you manage R packages. The Help tab provides access to R documentation, and the Viewer tab is used for viewing local web content. 

Lab 1_2:
## GitHub  
[GitHub](http://www.github.com) is a file storage and management site used by programmers. Programmers upload code to repositories (folders) and make it publicly available. We will use Github to store our work and manage workflows in BIS 15L. We have a main repository for our class [datascibiol](https://github.com/jmledford3115/datascibiol). You should **Star** this repository for future reference.  

## GitHub Desktop  
The easiest way to manage your GitHub repository is to use [GitHub desktop](https://desktop.github.com/). This is a  helpful tool that will allow you to manage all of your files. Open the program and login using your GitHub account.  

## Clone Your Repository  
In GitHub desktop, you will see your personal repository as an option under `clone repository`. Clone your repository to the desktop. Once you have done this any changes you make to files will be recorded and easily uploaded. Do not move this folder from your desktop.  

## Copy the Class Repository
In order to keep track of the class files, the last thing you need to do is make a copy of the class repository. I will make regular changes to the class repository including the addition of files and data that you will need. Do a search for `datascibiol` and download a copy to the desktop. You do this by navigating to the repository then going to the small green code button at top right. Click on the code button then go to Download ZIP.  

Copy the `lab1` folder from the class repository and paste it into your personal repository. By doing this, you have created your own copies of lab materials for lab 1.  

## Using GitHub  
You will use your repository to store all of your notes and work. There are three terms that we need to define:  
1. `Commit` takes a snapshot of your current progress.    
2. `Pull` this is the same as download.     
3. `Push` this is the same as upload.    

If you setup [GitHub desktop](https://desktop.github.com/), then you will see that it automatically tracks your changes. Clicking the `Commit to master` button in the bottom left corner means that you are recording a snapshot of your edits locally. Once this is done, you need to upload your changes to GitHub. This is done by clicking `Push origin`. Once you have pushed your changes you can double check by looking at your repository on github.com.    

## RMarkdown
The file format we use in class is called a markdown file. RMarkdown files have the extension `.Rmd` and can be created in RStudio by going to `File > New File > RMarkdown`. RMarkdown files allow you to combine text, code, and output in a single document.  

## Knitting
RMarkdown files need to be `knit` in order to produce a final document. This is done by clicking the `Knit` button at the top of the RStudio window. Knitting produces a final document in HTML format. You will turn in all homework assignments as knitted HTML files. If your code does not work, then the file will not knit.    

lab 2_intro:
1. What is the difference between R and RStudio?  R is the actual programming language doing the calculations/dataprocessing, RStudio is the dashboard/userinterface to make it easier to write code and manage files
2. What is GitHub and GitHub Desktop and why are they useful to programmers?  Github is a cloud-based data hosting service to store projects online and collaborate with others, Github Desktop is an application of Github outside of the website to sync files directly from the computer to the online cloud storage. 
3. What is version control and why is it useful to programmers?  Version control is a system of records changes to a file that can be accessed at any time. This lets you undo changes as needed and tracks the entire history of the file. 
4. What is a repository?  A repository is a project's folder containing all of its files, which is tracked online. 
5. What is the difference between committing and pushing?  committing will save the files locally on your computer, pushing will save the files online to the cloud

lab 2_1:
## Working Directories and Paths
Before we get started with R, it is important to understand the concept of working directories and paths. A working directory is the folder on your computer where R will look for files to read in and where it will save files that you create. You can check your current working directory using the `getwd()` function.
```{r}
getwd() #identify current working directory
```
What if you find that you are not in the correct working directory? You can change your working directory in two ways. The first is to look at the Session menu at the top of RStudio. Click on Session -> Set Working Directory -> Choose Directory. Then navigate to the folder you want to use as your working directory. 
The second way is to use the `setwd()` function. You will need to provide the full path to the folder you want to use. For example:
```{r}
#setwd("/Users/yourname/Documents/yourfolder") #uncomment and change to your path

#setwd("/Users/avongsav/Desktop/BIS15LW2026_avongsavanh/lab2")
```
What is a path? A path is the location of a file or folder on your computer. It tells R where to find files to read in or where to save files you create. Paths can be absolute or relative. An absolute path provides the full location of a file or folder, starting from the root directory. A relative path provides the location of a file or folder in relation to the current working directory.
## Objects
In order to access the potential of R we need to assign values or other types of data to `objects`. There is a specific format that I want you to follow, so please pay close attention.  

Assign a value to object 'x'. The '<-' symbol is read as 'gets'. In this case, x gets 42. Make sure that you are in the environment panel and you should see the value associated with 'x'. On a mac, you can push `option` and `-` to automatically generate the gets symbol.  
```{r}
x  <- 42 #x gets the value 42
```
Make two new objects, treatment and control. The value of treatment is 36 and the value of control is 38.
```{r}
treatment <- 36
control <- 38
```
What is the sum of treatment and control?
```{r}
treatment + control 
```
Here we make a new object `my_experiment` that is the sum of the treatment and control. Notice that I use `_` and not spaces. 
```{r}
my_experiment <- treatment + control
my_experiment
```
We can also use the function sum to do the same thing. Notice that if I give a new object the same name as an existing object, the old one is replaced.
```{r}
my_experiment <- sum(treatment,control)
```
To learn more about the sum function, use the help command `?`.
```{r}
?sum
```
What is the mean of 2+8+2? (Do the math in your head first). Does the following code match your prediction?
```{r}
sum <- 2 + 8 + 2 
mean <- sum / 3
``` 
Order of operations applies, so we need to tell R exactly what we want.
```{r}
(2+8+2)/3
```
Here we use the mean function. Notice that we use `c` which stands for concatenate. This combines the three numbers into a single object that the mean function can use. This type of data structure is called a vector.
```{r}
mean(c(2,8,2)) #c stands for combine and creates a vector
```
Why does this not work?
```{r}
mean(2,8,2) #outputs 2 
```
1. Create three new objects, `venom_GT`, `chiron`, and `veyron`. These are the fastest cars in the world. Assign each car to its top speed. The venom_GT can go 270, chiron is 261, and veyron is 268. 
```{r}
venom_GT <- 270
chiron <- 261
veyron <- 268
```
2. Use arithmetic to calculate the mean top speed for the cars.
```{r}
mean_speed <- (venom_GT + chiron + veyron)/3
mean_speed
```
3. Use the function `mean()` to calculate the mean top speed for the cars.
```{r}
mean_speed <- mean(c(venom_GT,chiron,veyron)) #make sure you use a vector using the c
```
## Types of Data
There are four frequently used `classes` of data: 1. numeric, 2. integer, 3. character, 4. logical.
```{r}
my_numeric <- 42
my_integer <- 2L #adding an L automatically denotes an integer
my_character <- "universe"
my_logical <- FALSE
```
To find out what type of data you are working with, use the `class()` function. This is important because sometimes we will need to change the type of data to perform certain analyses.
```{r}
class(my_numeric)
```
```{r}
class(my_integer)
```
You can use the `is()` and `as()` functions to clarify or specify a type of data.
```{r}
is.integer(my_numeric)
```
Let's convert my_numeric to an integer. 
```{r}
my_numeric <- as.integer(my_numeric) #important when we work with plots
```
```{r}
is.integer(my_numeric)
```
## Missing Data
R has a special way to designate missing data, the NA. NA values in R have specific properties which are very useful if your data contains any missing values. Later this quarter we will have a lab focused on dealing with NAs.  

NA values are used to designate missing data. `is.na` or `anyNA` are useful functions when dealing with NAs in data. 
```{r}
my_missing <- NA #NA stands for missing
```
```{r}
is.na(my_missing)
```
```{r}
anyNA(my_missing)
```
## Practice  
1. Let's create a vector that includes some missing data. For now, run the following code chunk.  
```{r}
new_vector <- c(7, 6.2, 5, 9, NA, 4, 9.8, 7, 3, 2)
```
2. Calculate the mean of `new_vector`.
```{r}
mean(new_vector)
```
4. Recalculate the mean using the following code chunk. Why is the useful?  
```{r}
mean(new_vector, na.rm = TRUE) #removes NA values, can use "T" instead of "TRUE"
```

hw2:
**4. Write a code chunk that divides `pi` by 2. Use the help command `?` to learn how to use the `round` function to limit your result to 4 significant digits.**  
```{r}
?round

piover2 <- pi/2
piover2rounded <- round(piover2,4)
piover2rounded
```
**5. Calculate the mean for: 2, 8, 4, 6, NA, 4, 9, 9, 10, NA. Please start by making a new vector `x` that holds these values then use `mean` to perform the calculation.**
```{r}
x = c(2, 8, 4, 6, NA, 4, 9, 9, 10, NA)
x
```
```{r}
mean(x, na.rm = T)
```

lab 3_1:
## Load the tidyverse
A library is a collection of R functions and data sets. The [tidyverse](https://tidyverse.org/) is a collection of R packages designed for data science. For this course, we will be using many of the packages in the tidyverse. We load the tidyverse with the command below.  
```{r}
library("tidyverse") 
#install.packages("package_name") #command to install packages
```
## Data Structures
In addition to classes of data, R also organizes data in different ways. These are called data structures and include vectors, lists, matrices, data frames, and factors. Here, we will introduce vectors and data frames.  
## Vectors
Vectors are a common way of organizing data in R.  We create vectors using the `c` command. The `c` stands for concatenate. We used this command in lab 2.  
A numeric vector.
```{r}
my_vector <- c(10, 20, 30) #this makes a new vector
```
A character vector.
```{r}
days_of_the_week <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
```
A convenient trick for creating a vector to play around with is to generate a sequence of numbers.
```{r}
my_vector_sequence <- c(1:100)
my_vector_sequence
```
## Identifying vector elements
We can use `[]` to pull out elements in a vector. We just need to specify their position in the vector; i.e. day 3 is Wednesday.
```{r}
days_of_the_week[4]
```
```{r}
my_vector_sequence[10]
```
## Practice
1. Use `[]` to determine which element in `my_vector_sequence` has a value of 15.
```{r}
my_vector_sequence[15]
```
2. We can use operators such as <, >, ==, <==, etc. Show all values in `my_vector_sequence` that are less than or equal to 10. 
```{r}
my_vector_sequence[my_vector_sequence <= 10] #important for dna sequence data 
```
3. If you use `[]` then you only get the values, not the logical evaluation of the entire vector. Experiment with this by adjusting the chunk below.    
```{r}
my_vector_sequence <= 10
```

## Data Frames
The data frame is the most common way to organize data within R. A data frame stores data of many different classes. Essentially, data frames are spreadsheets like you would find in Excel. We usually don't build data frames in RStudio from scratch, but this example will show you how they are structured.  
Let's build separate vectors that include length (in), weight (oz), and sex of three [ruby-throated hummingbirds](https://www.audubon.org/field-guide/bird/ruby-throated-hummingbird).  
```{r}
Sex <- c("male", "female", "male") #character vectors, strings of characters
Length <- c(3.2, 3.7, 3.4) #integer vectors
Weight <- c(2.9, 4.0, 3.1)
```
Since we work in the tidyverse, we use `tibble()` to create a data frame.  
```{r}
hbirds <- tibble(Sex, Length, Weight) #columns referred to as variables
hbirds 
```
What are the column names of our data frame? Notice that R defaulted to using the names of our vectors, but we could name them something else when creating the data frame, or rename them later.  
```{r}
names(hbirds)
```
What are the dimensions of the `hbirds` data frame? The `dim()` and `str()` commands provide this information.   
```{r}
dim(hbirds)
```
```{r}
str(hbirds) #str stands for structures
```
Let's use lowercase names when we create the data frame. We just changed to lowercase here, but we could use any names we wish.  
```{r}
hbirds <- tibble(sex = Sex, length = Length, weight = Weight)
hbirds
```
## Accessing Data Frame Columns and Rows 
The same methods of selecting elements in vectors and data matrices apply to data frames. We use `[]`. We have two positions where the first applies to the rows, and the second to the columns.  
The first row.  
```{r}
hbirds[1,]
```
The third column.  
```{r}
hbirds[3] #using hbirds[,3] also works
```
```{r}
hbirds[1,3] #row, column
```
## Calculations
We can use the `$` to access a column (variable) in a data frame. Here we calculate the mean length of the hummingbirds.  
```{r}
hbirdsmean <- mean(hbirds$length)
hbirdsmean
```
## Writing Data to File
We should save our hbirds data frame so we can use it again later! There are many ways to save data in R, here we write our data frame to a csv file. We use `row.names = FALSE` to avoid row numbers from printing out (on the first row when opened on a spreadsheet)
```{r}
#write.csv(hbirds, "hbirds_data.csv", row.names = FALSE) 
```

lab 3_2:
## Change column classes
When making plots or performing statistical modeling, it may be important to convert column classes. For example, if a column contains categorical data (e.g., scientist names, treatment groups, etc.) it may be best to convert that column to a factor class.  
Recall that factors are categorical variables that have a fixed number of possible values called levels.  
What is the class of the scientist column? Let's change it to factor and then show the levels of that factor.  
```{r}
names(hot_springs)
```
```{r}
class(hot_springs$scientist)
```
Did our change work? We can also check the levels of each column.
```{r}
hot_springs$scientist <- as.factor(hot_springs$scientist)
```
```{r}
class(hot_springs$scientist)
```
## Practice
1. Change the class of the springs column to factor.
```{r}
hot_springs$spring <- as.factor(hot_springs$spring)
```
2. What are the levels of the springs column?
```{r}
levels(hot_springs$spring)
```
```{r}
table(hot_springs$spring)
```
We can summarize our data frame with the`summary()` function.  (uses fish <- (another csv file))
```{r}
names(fish)
```
```{r}
str(fish)
```
```{r}
summary(fish)
```
`glimpse()` is another useful summary function.
```{r}
glimpse(fish)
```
`nrow()` gives the numbers of rows.
```{r}
nrow(fish)
```
`ncol` gives the number of columns.
```{r}
ncol(fish)
```
`dim()` gives the dimensions.
```{r}
dim(fish)
```
`names` gives the column names.
```{r}
names(fish)
```
`head()` prints the first n rows of the data frame.
```{r}
head(fish)
```
`tail()` prinst the last n rows of the data frame.
```{r}
tail(fish)
```
`table()` is useful when you have a limited number of categorical variables. It produces fast counts of the number of observations in a variable.
```{r}
table(fish$fish_id)
```

hw3 not included

lab4: 
## Libraries
```{r}
library(tidyverse)
library(palmerpenguins)
```
## Grammar of Graphics
The ability to quickly produce and customize graphs is a strength of R. Data visualizations are produced by the package `ggplot2` and it is a core part of the tidyverse. The syntax for using ggplot is specific and common to all types of plots. This is what Hadley Wickham calls a [Grammar of Graphics](http://vita.had.co.nz/papers/layered-grammar.pdf). The "gg" in `ggplot` stands for grammar of graphics.

## Philosophy
What makes a good chart? In my opinion a good chart is elegant in its simplicity. It provides a clean, clear visual of the data without being overwhelming to the reader. This can be hard to do and takes some careful thinking. Always keep in mind that the reader will almost never know the data as well as you do so you need to be mindful about how you present the facts.  

## Data Types
We first need to define some of the data types we will use to build plots.  

+ `discrete` quantitative data that only contains integers
+ `continuous` quantitative data that can take any numerical value (like weight, height)
+ `categorical` qualitative data that can take on a limited number of values 

## Basics
The syntax used by ggplot takes some practice to get used to, especially for customizing plots, but the basic elements are the same. It is helpful to think of plots as being built up in layers.  

In short, **plot= data + geom_ + aesthetics**.  

We start by calling the ggplot function, identifying the data, and specifying the axes. We then add the `geom` type to describe how we want our data represented. Each `geom_` works with specific types of data and R is capable of building plots of single variables, multiple variables, and even maps. Lastly, we add aesthetics.
Let's start by asking the question: How does body mass vary among penguin species?

To make a plot, we need to first specify the data and map the aesthetics. The aesthetics include how each variable in our data set will be used. In the example below, I am using the aes() function to identify the x and y variables in the plot.
```{r}
ggplot(data=penguins, #specifies the data
       mapping = aes(x = species,y = body_mass_g)) #map the aesthetics
```

Notice that we have a nice background, labeled axes, and even a value range of our variables on the y-axis- but no plot. This is because we need to tell ggplot how we want our data represented. This is called the geometry or `geom()`. There are many types of `geom`, see the ggplot [cheatsheet](https://posit.co/resources/cheatsheets/).

Here we want a boxplot, specified by `geom_boxplot()`. We will explore boxplots in more detail later, but for now we just need an example.
```{r}
ggplot(data=penguins, #specifies the data
       mapping = aes(x = species,y = body_mass_g)) + #map the aesthetics, note the plus sign
  geom_boxplot()
```
## Scatterplots and barplots
Now that we have a general idea of the syntax, let's explore two common plots: 1) scatter plots and 2) bar plots.

### 1. Scatter Plots
Scatter plots are good at revealing relationships that are not readily visible in the raw data.   
Let's ask the question: Is there a relationship between body mass and flipper length?
```{r}
names(penguins)
```
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point()
```
Notice the warning! R is telling us that there are some missing values in our data. This is common in real world data sets. R automatically omits these missing values when plotting. We can also deal with the NA's explicitly using the `na.rm=T` function.  
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point(na.rm = T)
```
To add a regression (best of fit) line, we add another layer.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point( na.rm = T) +
  geom_smooth(method = lm, se = T,  na.rm = T) #linear model, se = standard error
```
This graph is fine, but it doesn't distinguish between species. It might be helpful for the reader to see how each species is represented. We can do this by mapping the color aesthetic to species.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm, color = species)) +
  geom_point(na.rm = T) +
  geom_smooth(method = lm, se = T, na.rm = T) #linear model, se = standard error
```
The plot above looks good, but I think it's a bit messy having the regression line presented for each species. When we add the color aesthetic, it is passed down to all layers. To fix this, we can move the color aesthetic to just the geom_point layer.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species),na.rm = T) +
  geom_smooth(method = lm, se = T, na.rm = T)
```
Another helpful aesthetic is `shape` which can be used to distinguish points by shape instead of or in addition to color. A common approach is to map both color and shape to the same variable.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species, shape = species),na.rm = T) +
  geom_smooth(method = lm, se = T, na.rm = T)
```
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point(mapping = aes(color = bill_depth_mm, shape = species),na.rm = T) + #color for continuous example, not recommended for visuals
  geom_smooth(method = lm, se = T, na.rm = T)
```
But don't we want a title? We do this using the `labs()` function.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = body_mass_g,y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species, shape = species),na.rm = T) +
  geom_smooth(method = lm, se = T, na.rm = T) + 
  labs(title = "Body Mass (g) vs. Flipper Length (mm)",
       x = "Body Mass (g)",
       y = "Flipper Length (mm)") #labs stands for title
```
### Bar Plot: `geom_bar()`
The simplest type of bar plot counts the number of observations in a categorical variable. In this case, we want to know how many observations are present in the variable `species`. Notice that we do not specify a y-axis because it is count by default.  
```{r}
names(penguins)
```
How many individuals per species?
```{r}
ggplot(data=penguins, 
       mapping = aes(x = species)) + 
    geom_bar(mapping = aes(x = species))
```
What if we want to use the color aesthetic like we did for `geom_point` above? Let's try...
```{r}
ggplot(data=penguins, 
       mapping = aes(x = species)) + 
    geom_bar(mapping = aes(fill = species))
```
## Practice
1. Make a bar plot showing the number of penguins on each island.
```{r}
ggplot(data=penguins, 
       mapping = aes(x = island)) + 
    geom_bar(mapping = aes(fill = island))
```
2. Make a bar plot showing the number of penguins of each species on each island. (Hint: use `fill` to distinguish species)
```{r}
ggplot(data=penguins, 
       mapping = aes(x = island)) + 
    geom_bar(mapping = aes(fill = species), position = "dodge") + 
    labs(title = "#Penguins for each species by island")
```

hw4:
## Load the tidyverse
```{r}
library("tidyverse")
options(scipen = 999) #disable scientific notation
```
For this homework, we will be using the `msleep` dataset from the `ggplot2` package. This dataset contains information about the sleep habits of various mammals. [Reference](https://ggplot2.tidyverse.org/reference/msleep.html): V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007.
```{r}
msleep <- msleep
```
**1. Use a summary function of your choice to get an idea of the structure of the msleep data.**
```{r}
glimpse(msleep)
```
**2. What are the names of the variables? To get a better idea of what the variables mean, you can use `?msleep`.**
```{r}
?msleep
names(msleep)
```
**3. Make a plot that shows the number of mammals in each vore type. Which vore type is most represented in the data?**
```{r}
ggplot(data = msleep,
      mapping = aes(x = vore)) + 
      geom_bar(mapping = aes(fill = vore)) +
      labs(title = "Number of Mammals In Each Vore Type",
       x = "Vore Type",
       y = "Number of Mammals")
```
The herbi vore type is most represented by the data.
**4. What is the average length of time that mammals sleep?**
```{r}
sleep_average = mean(msleep$sleep_total)
sleep_average
```
The average length of time that mammals sleep is 10.43373 hours.
**5. Use `summary()` to get a quick idea of the range of values for the variable `bodywt`.**
```{r}
summary(msleep$bodywt)
```
**6. Build a plot that shows the range of `bodywt.` by vore type. Because the range of `bodywt` is so variable, you will need to scale the y-axis. You do this by adding `scale_y_log10()` as another layer.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = vore,y = bodywt)) + 
      geom_boxplot(mapping = aes(fill = vore)) +
      scale_y_log10() + 
      labs(title = "Vore Type vs. Bodyweight (kg)",
       x = "Vore Type",
       y = "Bodyweight (kg)")
```
**7. Is there a relationship between body weight and total sleep time? Add appropriate labels and a title to your plot.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = bodywt,y = sleep_total)) + 
      geom_point(mapping = aes(color = vore)) + #color instead of fill for dots 
      geom_smooth(method = lm, se = F, na.rm = T) +
      scale_x_log10() + 
      labs(title = "Bodyweight (kg) vs. Total Sleep Time (hr)",
       x = "Bodyweight (kg)",
       y = "Total Sleep Time (hr)")
```
There is a relationship between bodyweight and total sleep time.
**8. Which vore type tends to sleep the most? Build a plot to support your answer.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = vore,y = sleep_total)) + 
      geom_boxplot(mapping = aes(fill = vore)) +
      labs(title = "Vore Type vs. Total Sleep Time (hr)",
       x = "Vore Type",
       y = "Total Sleep Time (hr)")
```
The insecti vore type tends to sleep the most. 
**9. Is there a relationship between brain weight and body weight? Build a plot to support your answer.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = brainwt,y = bodywt)) + 
      geom_point(mapping = aes(color = vore)) +
      geom_smooth(method = lm, se = F, na.rm = T) +
      scale_y_log10() + 
      scale_x_log10() + 
      labs(title = "Brain Weight (kg) vs. Body Weight (kg)",
       x = "Brain Weight (kg)",
       y = "Body Weight (kg)")
```
There is a relationship between brain weight and body weight.
**10. Build one plot of your choice that provides additional insight into the msleep dataset. Be sure to include appropriate labels and a title.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = conservation,y = awake)) + 
      geom_boxplot(mapping = aes(fill = vore)) +
      labs(title = "Conservation Type vs. Time Awake (hr)",
       x = "Conservation Type",
       y = "Time Awake (hr)")
```

lab 5_1: ## Tidyverse
Recall that the [tidyverse](www.tidyverse.org) is a collection of packages that make workflow in R easier. The packages operate more intuitively than base R commands and share a common organizational philosophy. In lab 4, we learned how to use `ggplot2` to make visualization of data easier. In this lab, we will learn how to use the package `dplyr` to wrangle data. 
## dplyr
The first package that we will use to wrangle data is `dplyr`. `dplyr` is used to transform data frames by extracting, rearranging, and summarizing data such that they are focused on a question of interest. This is very helpful,  especially when wrangling large data, and makes dplyr one of most frequently used packages in the tidyverse. The two functions we will use most are `select()` and `filter()`.  
These functions are often called verbs, and the format used for each is the same. The output is always a new, more restricted dataframe.  
## `select()`
The verb `select()` allows you to pull out columns of interest from a dataframe; it does not affect the rows. To do this, just add the names of the columns of interest to the `select()` command. The order in which you add them, will determine the order in which they appear in the output.  
For smaller dataframes, `select()` may not make much sense. However, for larger dataframes with many variables, `select()` is very useful. Let's look at the fish data again.
```{r}
names(penguins)
```
We are only interested in species and body mass. We can use `select()` to extract these columns.
```{r}
select(penguins, species, body_mass_g)
```
To add a range of columns use `start_col:end_col`.
```{r}
select(penguins, species:flipper_length_mm)
```
```{r}
penguins <- penguins
penguins
```
The ! operator is useful in select. It allows us to select everything except the specified variables. (! is "naught")
```{r}
select(penguins, !body_mass_g)
```
Alternatively, you can use the `c()` function to exclude multiple columns. (c() is concatenate)
```{r}
select(penguins, !c(species, island, year, sex))
```
For very large data frames with lots of variables, `select()` utilizes lots of different operators to make things easier. Let's say we are only interested in the variables that deal with length.
```{r}
select(penguins,contains("mm"))
```
`starts_with()`
```{r}
select(penguins,starts_with("bill"))
```
`ends_with()`
```{r}
select(penguins,ends_with("mm"))
```
You can also select columns based on the class of data.  
```{r}
select(penguins, where(is.numeric))
```
There are a few verbs that go with `select()` because they deal with columns. One of these is `rename()`. `rename()` allows you to rename columns in a dataframe. The format is `new_name = old_name`.
```{r}
rename(penguins, body_mass=body_mass_g)
```
Alternatively, you can rename from within select.
```{r}
select(penguins, species, body_mass=body_mass_g)
```
The second is `relocate()`. `relocate()` allows you to move columns to different locations within a dataframe. Let's say we want to move `length` to be the first column in the dataframe.
```{r}
relocate(penguins, year)
```

lab 5_2:
## Review  
In the previous lab, we used `select()` to extract columns of interest from a data frame. This helps us focus our attention on the variables relevant to our question. However, it doesn't allow us to extract information in rows. The `filter()` function allows us to extract data in rows that meet specific criteria.  
## `filter()`
Unlike `select()`, `filter()` allows us to extract data that meet specific criteria within a row. Let's say that we are interested only in the Adelie penguins from the `penguins` data frame. We can use `filter()` to extract these observations.  
```{r}
glimpse(penguins)
```
```{r}
filter(penguins, species == "Adelie")
```
Similarly, if we are only interested in penguins with a body mass greater than or equal to 5500g we can use `filter()` to extract these observations.  
```{r}
filter(penguins, body_mass_g >= 5500)
```
+ `filter()` allows all of the expected operators; i.e. >, >=, <, <=, != (not equal), and == (equal).  
Using the `!` operator allows for the exclusion of specific observations.
```{r}
filter(penguins, species!= "Chinstrap")
```
## Using `filter()` with multiple observations  
The `%in%` [operator](https://www.tutorialspoint.com/r/r_operators.htm) determines whether or not a value is part of a vector.    
```{r}
filter(penguins, body_mass_g %in% c(5250,5500))
```
Alternatively, you can use `between` if you are looking for a range of specific values.  
```{r}
filter(penguins, between(body_mass_g, 5250, 5500))
```
How would you use `%in%` to get the same result?
```{r}
filter(penguins, body_mass_g %in% c(5250:5500)f) # or c(5250, 5300, 5350, 5400, 5450, 5500) if you know those are the other values
```
You can also extract observations "near" a certain value but you need to specify a tolerance.  
```{r}
filter(penguins, near(body_mass_g, 5400, tol = 100))
```
## Practice
The data are from: Gaeta J., G. Sass, S. Carpenter. 2012. Biocomplexity at North Temperate Lakes LTER: Coordinated Field Studies: Large Mouth Bass Growth 2006. Environmental Data Initiative.  [link](https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-ntl&identifier=267)  
1. Load the data
```{r}
fish <- read.csv("data/Gaeta_etal_CLC_data.csv")
fish
```
2. Filter the `fish` data to include the samples from lake "BO".
```{r}
names(fish)
```
```{r}
filter(fish,lakeid == "BO")
```
3. Filter the data to include all lakes except "AR".
```{r}
filter(fish,lakeid != "AR")
```
4. Filter the fish data to include all fish with a scalelength within 0.25 of 8.
```{r}
filter(fish, near(scalelength, 8, tol = 0.25))
```
5. Filter the fish data to include fish with a scalelength between 2 and 4.
```{r}
filter(fish, between(scalelength, 2, 4))
```
## Using `filter()` on multiple conditions
You can also use `filter()` to extract data based on multiple conditions. Below we extract the Chinstrap penguins that have a body mass greater than 4500g.
```{r}
names(penguins)
```
```{r}
filter(penguins, species == "Chinstrap" & body_mass_g > 4500) #can use , instead of &
```
Notice that the `|` operator generates a different result. Why?
```{r}
filter(penguins, species == "Chinstrap" | body_mass_g > 4500) # | is an "OR" so as long as one condition is fulfilled it will be included
```
Rules:  
+ `filter(condition1, condition2)` will return rows where both conditions are met. By default the , means &.    
+ `filter(condition1, !condition2)` will return all rows where condition one is true but condition 2 is not.  
+ `filter(condition1 | condition2)` will return rows where condition 1 or condition 2 is met.  
In this case, we filter out the penguins with a body mass over 5000g and a bill length over 50mm or a bill depth over 18mm.
```{r}
filter(penguins, body_mass_g > 5000, (bill_length_mm > 50 | bill_depth_mm > 18))
```
## Pipes `%>%` 
In order to start combining `select()`, `filter()`, and other functions efficiently, we need to learn pipes. Pipes feed the output from one function into the input of another function.  
Let's say we are only interested in the body mass of Gentoo penguins. We can use pipes to first select the species and body mass columns then filter that output only for Gentoo penguins. Notice that we only need to call the data one time.
```{r}
penguins %>%
  select(species, body_mass_g) %>% 
  filter(species == "Gentoo")
```
We can sort the output of `filter()` using the `arrange()` function. Here we filter for Gentoo penguins and then sort them by body mass in ascending order.
```{r}
penguins %>%
  select(species, body_mass_g) %>% 
  filter(species == "Gentoo") %>%
  arrange(body_mass_g)
```
To sort in descending order, wrap the variable name in `desc()`.
```{r}
penguins %>%
  select(species, body_mass_g) %>% 
  filter(species == "Gentoo") %>%
  arrange(desc(body_mass_g)) 
```
`slice()`
```{r}
penguins %>%
  select(species, body_mass_g) %>% 
  filter(species == "Gentoo") %>%
  slice_max(body_mass_g, n = 1) #slice_max gets largest bodymass, slice_min gets smallest 
```
## Practice
Let's use the fish data again to keep things simple.
1. We are interested in the radii length and scalelength of fish with the id `300`. Extract this information from the `fish` data. Please sort the data by radii length.
```{r}
names(fish)
```
```{r}
fish %>%
  select(fish_id, radii_length_mm, scalelength) %>% 
  filter(fish_id == 300) %>%
  arrange(radii_length_mm) 
```
2. We are interested in the fish from the lakes "AL" and "AR" with a radii length between 2 and 4. Extract this information from the `fish` data. Please sort the data in descending order by radii length.
```{r}
fish %>%
  select(lakeid, radii_length_mm) %>% 
  filter(lakeid == "AL" | lakeid == "AR") %>%
  filter(between(radii_length_mm, 2, 4)) %>%
  arrange(desc(radii_length_mm)) 
```

hw5:
## Data
**Database of vertebrate home range sizes.** 
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  
**1. Load the data into a new object called `homerange`.**
```{r}
homerange <- read.csv("data/Tamburelloetal_HomeRangeDatabase.csv")
```
**2. What is the structure of the `homerange` data? Does it have any NA's? Do a quick exploratory analysis of your choice below.**
```{r}
glimpse(homerange)
```
The homerange data does have NA values in "N", "preymass", "log10.preymass", "PPMR", and "prey.size.reference". 
```{r}
summary(homerange)
```
**3. Which animal groups are represented in the data? Make a plot that shows the number in each class.**
```{r}
names(homerange)
```
```{r}
ggplot(data = homerange,
      mapping = aes(x = class)) + 
      geom_bar(mapping = aes(fill = class)) +
      labs(title = "Number of Animal Groups in Each Class",
       x = "Class",
       y = "Number of Animal Groups")
```
**4. Are there more carnivores or herbivores represented in the data? Make a plot that shows the number in each trophic guild.**
```{r}
ggplot(data = homerange,
      mapping = aes(x = trophic.guild)) + 
      geom_bar(mapping = aes(fill = trophic.guild)) +
      labs(title = "Number of Animal Groups in Each Trophic Guild",
       x = "Trophic Guild",
       y = "Number of Animal Groups")
```
There are more carnivores in the data than herbivores. 
**5. Which herbivorous mammals have homeranges greater than 1,000,000 m²? Show only the common name, order, mean homerange size, and mean body mass. Arrange the data in descending order by homerange size and show only the top 10 species.**
```{r}
homerange %>%
  select(common.name, order, mean.hra.m2, mean.mass.g) %>% 
  filter(mean.hra.m2 >= 1000000) %>%
  arrange(desc(mean.hra.m2)) %>%
  slice_max(mean.hra.m2, n = 10)
```
**6. Which two species have the largest and smallest homerange sizes?**
```{r}
homerange %>%
  select(common.name, order, mean.hra.m2, mean.mass.g) %>% 
  arrange(desc(mean.hra.m2)) %>%
  slice_max(mean.hra.m2, n = 1)
```
```{r}
homerange %>%
  select(common.name, order, mean.hra.m2, mean.mass.g) %>% 
  arrange(mean.hra.m2) %>%
  slice_min(mean.hra.m2, n = 1)
```
**7. Which carnivorous mammals have the largest homeranges?**
```{r}
homerange %>%
  select(common.name, order, mean.hra.m2, mean.mass.g, trophic.guild) %>% 
  filter(trophic.guild == "carnivore") %>%
  arrange(desc(mean.hra.m2)) %>%
  slice_max(mean.hra.m2, n = 10)
```
**8. Is there a relationship between body mass and homerange size? Make a scatterplot to show this relationship. Be sure to include appropriate labels and a title.**
```{r}
ggplot(data = homerange,
      mapping = aes(x = mean.mass.g,y = mean.hra.m2)) + 
      geom_point(mapping = aes(color = trophic.guild)) + 
      geom_smooth(method = lm, se = F, na.rm = T) +
      scale_x_log10() +
      scale_y_log10() + 
      labs(title = "Body Mass (g) vs. Homerange Size (m^2)",
       x = "Body Mass (g)",
       y = "Homerange Size (m^2)")
```
There is a relationship between body mass and homerange size.
**9. Use filter to extract the data for only birds and mammals. Then make a plot that compares the homerange sizes of these two groups. Be sure to include appropriate labels and a title.**
```{r}
homerangebm = filter(homerange, taxon == "birds" | taxon == "mammals")

ggplot(data = homerangebm,
      mapping = aes(x = taxon, y = mean.hra.m2)) + 
      geom_boxplot(mapping = aes(fill = taxon)) +
      scale_y_log10() + 
      labs(title = "Taxon vs. Mean Homerange Size (m^2)",
       x = "Taxon",
       y = "Mean Homerange Size (m^2)")
```
**10. Do one additional exploratory analysis of your choice. Before writing the code, write the question you are trying to answer. The question should be biologically meaningful and appropriate for the data. It must include at least one `filter()`, `select()`, and `arrange()`.**
The question I am trying to answer is out of all of the bird species in the data, do carnivores have a larger homerange size than herbivores?
```{r}
homerange %>%
  select(common.name, taxon, order, family, trophic.guild, mean.mass.g, mean.hra.m2) %>% 
  filter(taxon == "birds") %>%
  arrange(desc(mean.hra.m2)) %>%
  slice_max(mean.hra.m2, n = 10)
  

homerangeb = filter(homerange, taxon == "birds")
ggplot(data = homerangeb,
      mapping = aes(x = trophic.guild, y = mean.hra.m2)) + 
      geom_boxplot(mapping = aes(fill = trophic.guild)) +
      scale_y_log10() + 
      labs(title = "Trophic Guild vs. Mean Homerange Size (m^2)",
       x = "Trophic Guild",
       y = "Mean Homerange Size (m^2)")
```

lab6:
## Load the tidyverse
```{r}
library("tidyverse")
library("janitor") #this will help with cleaning
library("palmerpenguins") #load the palmerpenguins package
options(scipen=999) #turn off scientific notation
```
1. Select species, island, and body mass from the penguins data. Arrange results by body mass.
```{r}
penguins %>%
  select(species, island, body_mass) %>%
  arrange(body_mass)
```
2. Filter the penguins data to only include observations from Biscoe and Dream islands.
```{r}
penguins %>%
  select(species, island, body_mass) %>%
  filter(island == "Biscoe" | island == "Dream")  %>% #ideally put select before filter
  arrange(body_mass)
```
```{r}
penguins %>% #alt method with concatenation
  select(species, island, body_mass) %>%
  filter(island == c("Biscoe", "Dream"))  %>% #ideally put select before filter
  arrange(body_mass)
```
```{r}
penguins %>% #alt method with concatenation + %in%
  select(species, island, body_mass) %>%
  filter(island %in% c("Biscoe", "Dream"))  %>% #ideally put select before filter
  arrange(body_mass)
```
3. Make a plot that shows the relationship between body mass and flipper length. How does this compare among different species?
```{r}
penguins %>%
  select(body_mass, flipper_len) %>%
  ggplot(data = penguins,
      mapping = aes(x = body_mass, y = flipper_len)) + 
      geom_point(mapping = aes(color = species)) + #can just be geom_point() for no colors
      geom_smooth(se = FALSE, method = lm) + 
      labs(title = "Body Mass (g) vs. Flipper Length (mm)",
           x = "Body Mass (g)",
           y = "Flipper Length (mm)")
```
## `distinct()`
`distinct()` looks for all unique observations in rows. This is a little tricky because it can look like it is working column-wise, but it is actually working row-wise.  
One helpful approach to new data is to find any duplicated rows. If we first look at the dimensions of the penguins data, we see it has 344 rows and 8 columns.
Using `distinct()` across all rows, we see there are no duplicates. This means every row contains unique observations across all variables.
```{r}
penguins %>%
  distinct() #looks across dataframe and compares every row, if there's a row thats the same as the other one it will flag them to ignore in new output
```
But if we only look at species, we can see that there are only 3 unique species in the data.
```{r}
penguins %>%
  distinct(species)
```
What if we want to know which islands each species occurs on?
```{r}
penguins %>%
  distinct(species, island, .keep_all=T) #.keep_all shows all other variables 
```
## `mutate()`
`mutate()` is another verb that acts on columns. It allows us to create new columns from existing columns in a data frame. When we use `mutate()`, the columns are added to the end of the dataframe by default. Let's create a new column that converts body mass from grams to kilograms.
```{r}
penguins %>%
  mutate(body_mass_kg = body_mass / 1000) %>%
  select(species, body_mass, body_mass_kg) %>%
  arrange(body_mass_kg)
```
## `mutate()` and `across()`
We use `across()` within `mutate()` to apply a function to multiple columns. This is especially helpful when cleaning data. For example, let's say we want to convert all columns that end with `mm` to centimeters. We can use `across()` to do this.
```{r}
penguins %>%
  mutate(across(ends_with("len"), ~./10)) %>%
  select(species, bill_length_cm = bill_len, bill_depth_cm = bill_dep, flipper_length_cm = flipper_len) 
```
What does the `~./10` mean? The `~` indicates that what follows is a formula (lambda function). The `.` represents the current column being processed. So, `./10` means "take the current column and divide it by 10". This operation is applied to all columns that end with `mm`.
## Cleaning Data
Cleaning raw data is an essential, but tedious step in data analysis. It's impossible to predict every scenario that you will come across, but there are some common issues that we can learn to address.
We already learned how to use `rename()` to change column names. We also learned how to rename columns from within `select()`. But, this can be very inefficient if we have a large dataset.
Let's have a look at some new data focused on mammal lifehistories. The data are from: S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.   [link](http://esapubs.org/archive/ecol/E084/093/)  
```{r}
mammals <- read.csv("data/mammal_lifehistories_v2.csv") #%>% clean_names() *Note: clean_names() done immediately if possible
```
One thing to notice is the column names are inconsistent. This is going to cause problems for us down the line. We could rename each column, one at a time, using `rename()`, but that would be tedious. Instead, we can use the `clean_names()` function from the `janitor` package to fix all of the column names at once.
```{r}
mammals <- mammals %>%
  clean_names() 
```
Notice that `clean_names()` has converted all column names to lowercase and replaced spaces with underscores. But, no adjustments were made to the data itself. What if we want to change observations from upper case to lower case?
```{r}
mammals %>%
  mutate(across(c("order","family"), tolower)) #tolower makes specific columns all lowercase
```
This will change all columns to lower case. But, notice what happens to numeric columns.
```{r}
mammals %>%
  mutate(across(everything(), tolower)) #everything in database made lowercase, numerics turned into characters however, best suited for only character-based dataframes/categorical
```
For this reason, it might be better to use `where` so we can specify only character columns.
```{r}
mammals %>%
  mutate(across(where(is.character), tolower))
```
## `if_else()`
We briefly introduce `if_else()` here because it allows us to use `mutate()` but not have the entire column affected in the same way. With `ifelse()`, you first specify a logical statement, afterwards what needs to happen if the statement returns `TRUE`, and lastly what needs to happen if it's  `FALSE`.  
Have a look at the data from mammals below. Notice that the values for newborn include `-999.00`. This is sometimes used as a placeholder for NA (but, is a really bad idea). We can use `if_else()` to replace `-999.00` with `NA`.  
```{r}
mammals %>%
  select(genus, species, newborn) %>%
  arrange(newborn)
```
```{r}
mammals %>%
  select(genus, species, newborn) %>%
  mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))
```
1. Following the example above, convert all -999 values in the mammals dataframe to NA.
```{r}
mammals <- mammals %>%
  mutate(across(c(mass, wean_mass, gestation, max_life, newborn, weaning, litter_size, afr, litters_year),  ~ifelse(.== -999.00, NA, .)))

  #mutate(across(everything(), tolower)) #everything in database made lowercase, numerics turned into characters however, best suited for only character-based dataframes/categorical
```
2. In the mammals data, make a new column `mass_kg` that that converts mass from grams to kilograms. Select the columns genus, species, mass, and mass_kg, and arrange the data by mass_kg in descending order. What is the common name for the species with the highest mass?  
```{r}
mammals %>%
  mutate(mass_kg = mass / 1000) %>%
  select(genus, species, mass, mass_kg) %>%
  arrange(desc(mass_kg))
```
Common Name is Blue Whale. 
3. What is the relationship between gestation and newborn mass?
```{r}
mammals %>%
  select(gestation, newborn) %>%
  ggplot(data = mammals,
         mapping = aes(x = log10(gestation), y = log10(newborn))) +
         geom_point() + 
        # geom_smooth(se = F, method = lm) + 
         labs(title = "Gestation vs. Newborn Mass", 
              x = "Gestation", 
              y = "Newborn Mass")
```
4. Which mammal has the longest life span in years?  
```{r}
mammals %>%
  mutate(max_life_years = max_life / 12)  %>% #instructor example has mutate and select lines swapped)
  select(genus, species, max_life, max_life_years) %>%
  arrange(desc(max_life_years))
```
5. Do larger mammals live longer?
```{r}
mammals %>%
  ggplot(mapping = aes(x = log10(mass), y = max_life)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

hw6:
**1. Load `IvindoData_DryadVersion.csv` and store it as a new object called `gabon`.**
```{r}
gabon <- read.csv("data/IvindoData_DryadVersion.csv")
```
**2. Use one or more of the summary functions you have learned to get an idea of the structure of the data.**  
```{r}
glimpse(gabon)
```
```{r}
summary(gabon)
```
**3. Clean the data as necessary following the conventions from lab 6.**
```{r}
gabon <- gabon %>%
  clean_names() %>%
  mutate(across(where(is.character),tolower))  
```
**4. Use `mutate()` Change the variables `HuntCat`, `LandUse`, and `TransectID` to factors. Be sure to store the output as a new dataframe and use it for the remaining questions.**
```{r}
gabon <- gabon %>%
  mutate(across(c(hunt_cat, land_use, transect_id), as.factor))
```
**5. Make a plot that shows the number of transects for each land use type.**
```{r}
gabon %>%
  ggplot(mapping = aes(x = land_use)) +
    geom_bar(mapping = aes(fill = land_use)) +
    labs(title = "Number of Transects for Each Land Use Type",
      x = "Land Use Type",
      y = "Number of Transects")
```
**6. How does vegetation richness compare among land use types?**
```{r}
gabon %>%
  ggplot(mapping = aes(x = land_use, y = veg_rich)) +
    geom_boxplot(mapping = aes(fill = land_use)) +
    labs(title = "Vegetation Richness (# Species) Among Land Use Types",
      x = "Land Use Type",
      y = "Vegetation Richness (# Species)")
```
**7. For which land use type (national parks, logging, or neither) is average all species diversity the greatest?**
```{r}
gabon %>%
  ggplot(mapping = aes(x = land_use, y = diversity_all_species)) +
    geom_boxplot(mapping = aes(fill = land_use)) +
    labs(title = "All Species Diversity Among Land Use Types",
      x = "Land Use Type",
      y = "All Species Diversity")
```
**8. Is there a relationship between vegetation richness and mammal species richness?**
```{r}
gabon %>%
  ggplot(mapping = aes(x = veg_rich, y = rich_mammal_species)) + 
    geom_point(mapping = aes(color = land_use)) + 
    geom_smooth(method = lm, se = F, na.rm = T) +
    labs(title = "Vegetation Richness (# Species) vs. Mammal Species Richness (# Species)",
      x = "Vegetation Richness (# Species)",
      y = "Mammal Species Richness (# Species)")
```
There is no relationship between vegetation richness and mammal species richness (line of best fit doesn't fit well with the data plotted. 
**9. Use `filter` to find all transects that have greater than 15 tree species or a breast height diameter between 50 and 60cm.  **
```{r}
gabon %>%
  select(transect_id, land_use, veg_rich, veg_dbh) %>%
  filter(veg_rich >= 15 | between(veg_dbh, 50, 60)) %>%
  arrange(desc(veg_dbh))
```
**10.Which transects and land use types have more than 10 tree species and 10 mammal species? Use `arrange()` to sort by the number of tree species.**
```{r}
gabon %>%
  select(transect_id, land_use, veg_rich, rich_mammal_species) %>%
  filter(veg_rich >= 10 & rich_mammal_species >= 10) %>%
  arrange(desc(veg_rich))
```

lab7:
## Summary statistics
When first exploring data, it is often a good idea to produce some basic summary statistics. This can help you identify potential problems with the data (e.g., outliers, missing values, skew) as well as get a general sense of the distribution of values for a given variable. So far, you have learned to calculate summary statistics using _base R_ functions such as `mean()`, `sd()`, `min()`, and `max()`.  
```{r}
summary(msleep)
```
## Practice
1. Calculate the mean body weight of large and small mammals in msleep. Large mammals have a body weight greater than 150, while small mammals have a body weight less than 10. I did large mammals for you as an example.  
```{r}
large <- msleep %>% 
  filter(bodywt > 150) %>% 
  arrange(desc(bodywt))
large
```
```{r}
small <- msleep %>% 
  filter(bodywt < 10) %>% 
  arrange(bodywt)
small
```
```{r}
largem <- mean(large$bodywt)
largem
smallm <- mean(small$bodywt)
smallm
```
## `summarize()`
Fortunately, dplyr has some functions that make producing summary statistics easier. The verb `summarize()` produces summary statistics for a given variable in a data frame. Let's try the examples above using `summarize()`. 
Large mammals
```{r}
msleep %>%
  filter(bodywt > 150) %>%
  summarize(mean_bodywt_large = mean(bodywt)) 
```
Small mammals
```{r}
msleep %>%
  filter(bodywt < 10) %>%
  summarize(mean_bodywt_small = mean(bodywt)) 
```
You can also combine functions to make summaries for multiple variables.
```{r}
msleep %>%
  filter(bodywt > 150) %>%
  summarize(mean_bodywt_large = mean(bodywt), 
            min_bodywt_large = min(bodywt),
            max_bodywt_large = max(bodywt),
            sd_bodywt_large = sd(bodywt),
            total = n()) #number of observations
```
Maybe you want to summarize all of the numeric variables in the data frame. You can use the `where()` function to select only the numeric variables.
```{r}
msleep %>%
  select(where(is.numeric)) %>%
  summarize_all(mean)
```
Don't forget to remove NA's.
```{r}
msleep %>%
  select(where(is.numeric)) %>%
  summarize_all(mean, na.rm = T)
```
This is a little more efficient.
```{r}
msleep %>%
  summarize(across(where(is.numeric),
                   ~mean(.x, na.rum = T))) #.x means for every cell
```
## Practice
1. What is the mean, min, and max `bodywt` for the taxonomic order Primates? Provide the total number of observations.
```{r}
msleep %>%
  filter(order == "Primates") %>%
  summarize(mean_bodywt_p = mean(bodywt), 
            min_bodywt_p = min(bodywt),
            max_bodywt_p = max(bodywt),
            total = n()) #number of observations
```
`n_distinct()` is a handy way of cleanly presenting the number of distinct observations. Notice that there are 9 genera with over 100 in body weight.
```{r}
msleep %>%
  filter(bodywt > 150) %>%
  distinct()
```
Here we show the number of distinct genera over 100 in body weight.
```{r}
msleep %>%
  filter(bodywt > 100) %>%
  summarize(n_genara = n_distinct(genus))
```
Not to be confused with n(), which counts the total number of observations.
```{r}
msleep %>%
  filter(bodywt > 100) %>%
  summarize(total = n())
```
`pull()` is another useful verb that extracts a single column from a data frame as a vector.
```{r}
msleep %>%
  filter(bodywt > 100) %>%
  distinct() %>%
  pull(genus)
```
## Practice
1. How many genera are represented in the msleep data frame?
```{r}
msleep %>%
  summarize(n_genara = n_distinct(genus))
```
## Counts
Although `summarize()` is helpful, oftentimes we are only interested in counts. The [janitor package](https://garthtarr.github.io/meatR/janitor.html) does a lot with counts, but there are also functions that are part of dplyr that are useful.  

For example, if we wanted to count the number of observations for each `vore` category in the `msleep` data frame we could do the following:
```{r}
msleep %>%
  count(vore)
```
```{r}
msleep %>%
  count(vore, sort = T) %>% #sort = T sorts in descending order
  ggplot(aes(x = vore, y = n)) + 
  geom_col(mapping = aes(fill = vore)) + #use geom_col if we have an x and a y variable 
  labs(title = "Number of Species per Vore",
  x = "Vore",
  y = "Number of Species")
```
You can even combine multiple variables to get counts. For example, if we wanted to count the number of observations for each `vore` category within each `order` category we could do the following:
```{r}
msleep %>%
  count(vore, order)
```
The tabyl() function from the janitor package is also very useful for producing counts and percentages.  
```{r}
msleep %>%
  tabyl(vore)
```
## Practice
1. In the taxonomic order Carnivora, count the number of observations for each conservation status.
```{r}
msleep %>%
  filter(order == "Carnivora") %>%
  count(conservation, sort = T)
```
2. Among herbivores, which order is most represented? Return the top 5 orders with counts, sorted in descending order.
```{r}
msleep %>%
  filter(vore == "herbi") %>%
  count(order, sort = T) %>%
  slice_max(n, n = 5)
```

hw7:
**1. Before starting data analysis, read the abstract of the paper to get an idea of the questions being asked. In 2-3 sentences, describe what the study is testing and the variables involved.**  
The study interest of the paper was looking ito whether severe historical elepahnt harvests eliminated large tuskers among survivors and whether elephants born thereafter had smaller tusks. 
The paper looks into the tusk size of elephants sampled in different time periods, and also categorized the elephants by sex, age, tusk circumference, and length. 
**2. Load `elephants.csv` and store it as a new object called `elephants`.**  
```{r}
elephants <- read.csv("data/elephants.csv")
```
**3. Clean the data by converting variable names to lowercase with no spaces or special characters.**  
```{r}
elephants <- elephants %>%
  clean_names() %>%
  mutate(across(where(is.character),tolower))  
```
**4. Use one or more of the summary functions you have learned to get an idea of the structure of the data.**   
```{r}
glimpse(elephants)
```
```{r}
summary(elephants)
```
**5. Use `mutate()` Change the variables `years_of_sample_collection`, `elephant_id`, and `sex` to factors. Be sure to store the output as a new dataframe and use it for the remaining questions.**  
```{r}
elephants <- elephants%>%
  mutate(across(c(years_of_sample_collection, elephant_id, sex), as.factor))
```
**6. From which years were data collected? Show the sample periods below.**  
```{r}
elephants %>%
  select(years_of_sample_collection) %>%
  count(years_of_sample_collection)
```
**7. How many males and females were sampled in this study?**  
```{r}
elephants %>%
  select(sex) %>%
  count(sex)
```
**8. What is the mean, median, and standard deviation for age of males and females included in the study? Separate the results by year of sample collection. Does the sampling look even between years and sexes?**  
```{r}
elephants %>%
  filter(years_of_sample_collection == "1966-68",
         sex == "m") %>%
  summarize(mean_age_m = mean(estimated_age_years, na.rm = T), 
            min_age_m = min(estimated_age_years, na.rm = T),
            max_age_m = max(estimated_age_years, na.rm = T),
            sd_age_m = sd(estimated_age_years, na.rm = T),
            total = n()) 
```
```{r}
elephants %>%
  filter(years_of_sample_collection == "1966-68",
         sex == "f") %>%
  summarize(mean_age_f = mean(estimated_age_years, na.rm = T), 
            min_age_f = min(estimated_age_years, na.rm = T),
            max_age_f = max(estimated_age_years, na.rm = T),
            sd_age_f = sd(estimated_age_years, na.rm = T),
            total = n()) 
```
```{r}
elephants %>%
  filter(years_of_sample_collection == "2005-13",
         sex == "m") %>%
  summarize(mean_age_m = mean(estimated_age_years, na.rm = T), 
            min_age_m = min(estimated_age_years, na.rm = T),
            max_age_m = max(estimated_age_years, na.rm = T),
            sd_age_m = sd(estimated_age_years, na.rm = T),
            total = n()) 
```
```{r}
elephants %>%
  filter(years_of_sample_collection == "2005-13",
         sex == "f") %>%
  summarize(mean_age_f = mean(estimated_age_years, na.rm = T), 
            min_age_f = min(estimated_age_years, na.rm = T),
            max_age_f = max(estimated_age_years, na.rm = T),
            sd_age_f = sd(estimated_age_years, na.rm = T),
            total = n()) 
```
**9. Is age (independent variable) a positive predictor of tusk length (dependent variable)? Create a plot that shows the relationship between these variables and add a linear model fit line.**  
```{r}
elephants %>%
  ggplot(mapping = aes(x = estimated_age_years, y = tusk_length_in_cm)) +
  geom_point(mapping = aes(color = sex)) + 
  geom_smooth(method = lm, se = F, na.rm = T) + 
  labs(title = "Age vs. Tusk Length (cm)",
      x = "Age",
      y = "Tusk Length (cm)")
```
Based on the plot, age is a positive predictor of tusk length.
**10. Is shoulder height (independent variable) a positive predictor of tusk length (dependent variable)? Create a plot that shows the relationship between these variables and add a linear model fit line.**  
```{r}
elephants %>%
  ggplot(mapping = aes(x = shoulder_height_in_cm, y = tusk_length_in_cm)) +
  geom_point(mapping = aes(color = sex)) + 
  geom_smooth(method = lm, se = F, na.rm = T) + 
  labs(title = "Shoulder Height (cm) vs. Tusk Length (cm)",
      x = "Shoulder Height (cm)",
      y = "Tusk Length (cm)")
```
Shoulder height is a positive predictor of tusk length. 
**11. The authors argue that because poachers preferentially target elephants with large tusks, this has resulted in a decrease in average tusk length. Is this supported by the data? Show your code and calculations below.**  
```{r}
elephants %>%
  ggplot(mapping = aes(x = years_of_sample_collection, y = tusk_length_in_cm)) +
  geom_boxplot(mapping = aes(color = years_of_sample_collection)) + 
  labs(title = "Time Period vs. Tusk Length (cm)",
      x = "Time Period", 
      y = "Tusk Length (cm)") #for this problem, better to not plot and instead use means/max/min/sd 
```
The data does support that because poachers preferentially target elephants with large tusks, this has resulted in a decrease in average tusk length
```{r}
elephants %>%
    filter(years_of_sample_collection == "1966-68") %>%
  summarize(mean_tusk_length = mean(tusk_length_in_cm, na.rm = T), 
            min_tusk_length = min(tusk_length_in_cm, na.rm = T),
            max_tusk_length = max(tusk_length_in_cm, na.rm = T),
            total = n()) 
```
```{r}
elephants %>%
    filter(years_of_sample_collection == "2005-13") %>%
  summarize(mean_tusk_length = mean(tusk_length_in_cm, na.rm = T), 
            min_tusk_length = min(tusk_length_in_cm, na.rm = T),
            max_tusk_length = max(tusk_length_in_cm, na.rm = T),
            total = n()) 
```
**12. Male elephants reach effective sexual maturity at 25 years while females are sexually mature at 12 years. Make a new dataframe that extracts only the males and females at sexual maturity. Then, make a plot that shows the range of tusk length between the two sample periods for these mature elephants.**   
```{r}
elephants_sm <- elephants %>%
  filter((sex == "m" & estimated_age_years >= 25) | (sex == "f" & estimated_age_years >= 12)) #must be isolated from the code chunk with ggplot for ggplot to utilize the new dataframe created, done to help with pipe management (doesn't mix up piping and ggplot functions)
```
```{r}
elephants_sm %>% 
  ggplot(mapping = aes(x = years_of_sample_collection, y = tusk_length_in_cm)) +
  geom_boxplot(mapping = aes(color = sex)) + 
  labs(title = "Time Period vs. Tusk Length (cm) for Males and Females at Sexual Maturity",
      x = "Time Period",
      y = "Tusk Length (cm)") #just present this graph
```
## Submit the Homework
1. Save your work and knit the .rmd file.  
2. Open the .html file and "print" it to a .pdf file in Google Chrome (not Safari).    
3. Go to the class Canvas page and open Gradescope.  
4. Submit your .pdf file to the homework assignment- be sure to assign the pages to the correct questions.  
5. Commit and push your work to your repository.






