---
title: "Midterm 2 W26 - Corrections"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: no
---

## Instructions
This is an in-person exam. Your exam must be complete and uploaded to Gradescope by 2:00p.  

Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code must be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Be sure to add your name to the author header above. 

Your code must knit in order to be considered. If you are stuck and cannot answer a question, then comment out your code and knit the document. You may use your notes, labs, and homework to help you complete this exam. Do not use any other resources- including AI assistance or other students' work.  

Don't forget to answer any questions that are asked in the prompt! Each question must be coded; it cannot be answered by a sort in a spreadsheet or a written response only.  

For all plots you create, a title and clearly labeled axes must be provided.  

Be sure to push your completed midterm to your repository and upload the document to Gradescope. _Incorrectly uploading the exam to Gradescope will result in a 5 point deduction._ This exam is worth 50 points. 

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(naniar)
library(ggmap)
library(lubridate)
library(shiny)
``` 

## API Key
Since you will build a map as part of the exam, take a moment to register your API key. You may need to look up your [stadia maps](https://client.stadiamaps.com/dashboard/overview) API key online (please do that now).
```{r}
register_stadiamaps("efbbad0c-b639-4a5e-bc42-a820b574fc9b", write = FALSE)
```

## About the data
This study is focused on the fasting behavior of polar bears in the Southern Beaufort (SB) and Chukchi seas (CS). The data were collected from 1983-2017 and include measurements of blood urea nitrogen (BUN) and creatinine, which are used to calculate a ratio that indicates whether a bear is fasting or not. The original paper is included in the data folder. _It may help to quickly read the abstract before starting the exam._  

The data are from: Rode, K.D., Wilson, R.R., Douglas, D.C., Muhlenbruch, V., Atwood, T.C., Regehr, E.V., Richardson, E.S., Pilfold, N.W., Derocher, A.E., Durner, G.M., Stirling, I., Amstrup, S.C., St. Martin, M., Pagano, A.M., and others, 2017, Spring fasting behavior in a marine apex predator provides an index of ecosystem productivity: Global Change Biology, v. 24, no. 1, p. 410–423, https://doi.org/10.1111/gcb.13933.  

## Data resources
- https://data.usgs.gov/datacatalog/data/USGS:ASC141
- https://www.sciencebase.gov/catalog/item/61a1a9e7d34eb622f69743d2

## The variables include:  
- `bear_id`: A unique number given to each bear.    
- `capture_date`: The data when the bear was tranquilized and assessed.  
- `popn`: The population to which the bear belongs.  
- `sex `: Sex of each bear.    
- `age_class`: The age class of each bear (A=adult, S=subadult).   
- `mating`: Whether or not the bear is pursuing a mate.  
- `denning`: Whether or not the bear denned last year.  
- `lat`: Latitude where the bear was located.  
- `long`: Longitude where the bear was located.  
- `urea_nitrogen`: Concentration of urea nitrogen.  
- `creatinine`: Concentration of creatinine.  

## Load the data
Run the code chunk below to load the data, clean the names, and do some tidying.
```{r message=FALSE, warning=FALSE}
polar_bears <- 
  read_csv("data/polarBear_serumUreaCreatinine_beaufortChukchi_rode.csv") %>% 
  clean_names() %>% 
  select(-den_method, -no_cubs, -cub_age) %>%
  filter(age_class==c("A", "S")) %>% 
  mutate(sex=toupper(sex), 
         age_class=toupper(age_class))
```

```{r}
polar_bears <- polar_bears %>% #going back to the original data
  replace_with_na(replace = list(mating = "unknown", 
                                 denning = "unknown")) #found unknown in denning as well through inspection
```

```{r}
polar_bears %>% 
  summarize(across(everything(), ~ sum(is.na(.)))) 
```

3. (6 points) Is sampling even for both sexes? Make a count the shows the number of males and females sampled by population and make a plot to visualize the counts.
```{r}
polar_bears %>%
  select(sex) %>%
  count(sex)
```
The sampling is not even as there are more much more females than males.

```{r}
polar_bears %>% 
  ggplot(mapping = aes(x = popn, fill = sex)) +
  geom_bar(position = "dodge") + 
  labs(title = "Number of Male and Female Polar Bears by Population",
      x = "Population",
      y = "Number of Polar Bears")
```

The error made above was that the count was only by sex and not by population and sex. The corrected code is below: 
```{r}
polar_bears %>%
  select(sex, popn) %>%
  count(sex, popn)
```
The sampling is not even as there are more much more females than males between the two populations.

```{r}
polar_bears %>% 
  ggplot(mapping = aes(x = popn, fill = sex)) +
  geom_bar(position = "dodge") + 
  labs(title = "Number of Male and Female Polar Bears by Population",
      x = "Population",
      y = "Number of Polar Bears")
```

4. (6 points) How many bears were sampled in each year for both populations? Make a line plot(s) to visualize sampling by year for both populations. (hint: you will need to pull apart the date to get the year in its own column)
```{r}
polar_bears <- polar_bears %>%
  separate(capture_date, into = c("year", "month", "day"), sep = "-") 
polar_bears
```


```{r}
polar_bears %>%
  filter(popn == "SB") %>% 
  select(year) %>% 
  count(year) %>% 
  arrange(desc(n))
```
```{r}
polar_bears %>%
  filter(popn == "CS") %>% 
  select(year) %>% 
  count(year) %>% 
  arrange(desc(n))
```

```{r}
polar_bears %>% #Make a line plot(s) to visualize sampling by year for both populations.
  filter(popn == "SB") %>% 
  select(year) %>% 
  count(year) %>% 
  ggplot(aes(x = year, y = n)) +
  geom_line() +
  geom_point(shape = 5) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  labs(title = "Polar Bear Sampling by Year - SB Population",
      x = "Year",
       y = "Number of Polar Bears Sampled")
```
```{r}
polar_bears %>% #Make a line plot(s) to visualize sampling by year for both populations.
  filter(popn == "CS") %>% 
  select(year) %>% 
  count(year) %>% 
  ggplot(aes(x = year, y = n)) +
  geom_line() +
  geom_point(shape = 5) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  labs(title = "Polar Bear Sampling by Year",
      x = "Year",
       y = "Number of Polar Bears Sampled")
```

The error made above was that 

```{r}
polar_bears <- polar_bears %>%
  mutate(un_creat_ratio = (urea_nitrogen * 0.466) / creatinine) 
polar_bears
```
```{r}
polar_bears <- polar_bears %>% 
  mutate(fasting = case_when(un_creat_ratio <= 10 ~ "yes",
                             un_creat_ratio > 10 ~ "no"))  
polar_bears
```

8. (6 points) The authors found that the number of fasting bears increased in the SB population but decreased in the CS population. Make a plot(s) that shows the number of fasting bears by year for both populations to visualize this trend.
```{r}
polar_bears %>% #Make a line plot(s) to visualize sampling by year for both populations.
  filter(popn == "SB", fasting == "yes") %>% 
  select(year) %>% 
  count(year) %>% 
  ggplot(aes(x = year, y = n)) +
  geom_line(color = "red") +
  geom_point(shape = 5) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  labs(title = "Fasting Polar Bear Sampling by Year - SB Population",
      x = "Year",
       y = "Number of Fasting Polar Bears Sampled")
```
```{r}
polar_bears %>% #Make a line plot(s) to visualize sampling by year for both populations.
  filter(popn == "CS", fasting == "yes") %>% 
  select(year) %>% 
  count(year) %>% 
  ggplot(aes(x = year, y = n)) +
  geom_line(color = "red") +
  geom_point(shape = 5) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  labs(title = "Fasting Polar Bear Sampling by Year - CS Population",
      x = "Year",
       y = "Number of Fasting Polar Bears Sampled")
```

The error made above was that 



## Submit the Midterm
1. Save your work and knit the .rmd file.  
2. Open the .html file and "print" it to a .pdf file in Google Chrome (not Safari).    
3. Go to the class Canvas page and open Gradescope.  
4. Submit your .pdf file to the midterm assignment- be sure to assign the pages to the correct questions.  
5. Commit and push your work to your repository. 