---
title: "Homework 4"
author: "Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---

## Instructions
Answer the following questions and/or complete the exercises in RMarkdown. Please embed all of your code and push the final work to your repository. Your report should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Any plots must have appropriate titles and aesthetics.  

## Load the tidyverse
```{r}
library("tidyverse")
options(scipen = 999) #disable scientific notation
```

For this homework, we will be using the `msleep` dataset from the `ggplot2` package. This dataset contains information about the sleep habits of various mammals. [Reference](https://ggplot2.tidyverse.org/reference/msleep.html): V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007.

```{r}
msleep <- msleep
```


**1. Use a summary function of your choice to get an idea of the structure of the msleep data.**
```{r}
glimpse(msleep)
```

**2. What are the names of the variables? To get a better idea of what the variables mean, you can use `?msleep`.**
```{r}
?msleep
names(msleep)
```

**3. Make a plot that shows the number of mammals in each vore type. Which vore type is most represented in the data?**
```{r}
ggplot(data = msleep,
      mapping = aes(x = vore)) + 
      geom_bar(mapping = aes(fill = vore)) +
      labs(title = "Number of Mammals In Each Vore Type",
       x = "Vore Type",
       y = "Number of Mammals")
```
The herbi vore type is most represented by the data.

**4. What is the average length of time that mammals sleep?**
```{r}
sleep_average = mean(msleep$sleep_total)
sleep_average
```
The average length of time that mammals sleep is 10.43373 hours.

**5. Use `summary()` to get a quick idea of the range of values for the variable `bodywt`.**
```{r}
summary(msleep$bodywt)
```

**6. Build a plot that shows the range of `bodywt.` by vore type. Because the range of `bodywt` is so variable, you will need to scale the y-axis. You do this by adding `scale_y_log10()` as another layer.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = vore,y = bodywt)) + 
      geom_boxplot(mapping = aes(fill = vore)) +
      scale_y_log10() + 
      labs(title = "Vore Type vs. Bodyweight (kg)",
       x = "Vore Type",
       y = "Bodyweight (kg)")
```

**7. Is there a relationship between body weight and total sleep time? Add appropriate labels and a title to your plot.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = bodywt,y = sleep_total)) + 
      geom_point(mapping = aes(color = vore)) + #color instead of fill for dots 
      geom_smooth(method = lm, se = F, na.rm = T) +
      scale_x_log10() + 
      labs(title = "Bodyweight (kg) vs. Total Sleep Time (hr)",
       x = "Bodyweight (kg)",
       y = "Total Sleep Time (hr)")
```
There is a relationship between bodyweight and total sleep time.

**8. Which vore type tends to sleep the most? Build a plot to support your answer.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = vore,y = sleep_total)) + 
      geom_boxplot(mapping = aes(fill = vore)) +
      labs(title = "Vore Type vs. Total Sleep Time (hr)",
       x = "Vore Type",
       y = "Total Sleep Time (hr)")
```
The insecti vore type tends to sleep the most. 

**9. Is there a relationship between brain weight and body weight? Build a plot to support your answer.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = brainwt,y = bodywt)) + 
      geom_point(mapping = aes(color = vore)) +
      geom_smooth(method = lm, se = F, na.rm = T) +
      scale_y_log10() + 
      scale_x_log10() + 
      labs(title = "Brain Weight (kg) vs. Body Weight (kg)",
       x = "Brain Weight (kg)",
       y = "Body Weight (kg)")
```
There is a relationship between brain weight and body weight.

**10. Build one plot of your choice that provides additional insight into the msleep dataset. Be sure to include appropriate labels and a title.**
```{r}
ggplot(data = msleep,
      mapping = aes(x = conservation,y = awake)) + 
      geom_boxplot(mapping = aes(fill = vore)) +
      labs(title = "Conservation Type vs. Time Awake (hr)",
       x = "Conservation Type",
       y = "Time Awake (hr)")
```

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  